---
title: "Lesson 18: Scatterplots"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to show how much two numbers are related.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

# Scatterplots

## The basics

Scatterplots are very good at showing **relationships between two numbers**. They can help us ask questions of data and find people or teams who are doing better or worse than we would expect, provided there's a strong relationship in the data. 

If this sounds like the correlations and regressions tutorial, you're right. Scatterplots and regressions fit together like a hand in a glove. Scatterplots help you *see* what the regression is trying to say. 

So let's ask a silly question. Nebraska is a volleyball school. We're good at it. We recruit well for it. But is volleyball recruiting as simple as recruiting the tallest athlete and let them pound the ball at shorter athletes? Put more simply: Does height dictate attacking success? Taller athlete, more kills?

First, we need libraries and every college volleyball player from the last season. **What we're interested in is less about a specific player and more about a general point: Are these numbers related and by how much? What can they tell you about your players in general?**

Here is that data. Per usual: **For purposes of this exercise, you don't need to do this. The data is included here if you want to try this in your own notebook.**

```{r echo=FALSE, class.output="bg-info", results="asis",  message=FALSE,  warning=FALSE}
library(downloadthis)
library(glue)

dllink <- download_link(
  link = "http://mattwaite.github.io/sportsdatafiles/ncaa_womens_volleyball_playerstats_2021.csv",
  button_label = "Download csv file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

glue("<pre><p><strong>For this walkthrough:</strong></p><p>{dllink}</p></pre>")
```

We'll need the tidyverse.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
```
```{r load-tidyverse-solution}
library(tidyverse)
```
```{r load-tidyverse-check}
grade_this_code()
```

And the data.

```{r scatter-load-data, message=FALSE, warning=FALSE}
players <- read_csv("http://mattwaite.github.io/sportsdatafiles/ncaa_womens_volleyball_playerstats_2021.csv")

tall <- players %>% 
  mutate(height = (12 * feet) + inches)

newtall <- tall %>%
  filter(
    height > 0 &
      s > 65
         )
```
```{r scatter-load-data-exercise, exercise = TRUE}
players <- read_csv("http://mattwaite.github.io/sportsdatafiles/ncaa_womens_volleyball_playerstats_2021.csv")
```
```{r scatter-load-data-exercise-solution}
players <- read_csv("http://mattwaite.github.io/sportsdatafiles/ncaa_womens_volleyball_playerstats_2021.csv")
```
```{r scatter-load-data-exercise-check}
grade_this_code()
```

Let's glimpse it quick to see what we have: 

```{r glimpse-data, exercise=TRUE, exercise.setup = "scatter-load-data"}
glimpse(???????)
```
```{r glimpse-data-solution}
glimpse(players)
```
```{r glimpse-data-check}
grade_this_code()
```

We now have every player and their season stats. So the question is, how much does does a thing matter? That's the question you're going to answer. 

## Scatterplots chart relationships

In our case, we want to know how much height influence kills? How much difference can we explain in kills by knowing how tall an athlete is? 

The first thing we need to do is turn height into a single number. If you look at the data, you'll see height is expressed in feet and inches. We need to figure out how to put that into one number -- and no 5'11" is not a number. But how do we do that? 

### Exercise 1: Make your data.

By converting it all to inches and adding them up. We're going to make a new dataframe called `tall`. We have the `feet` and `inches` of every player. How many inches are in a foot?

```{r scatter1, exercise=TRUE, exercise.setup = "scatter-load-data", message=FALSE}
???? <- players %>% 
  mutate(height = (?? * feet) + inches)
```
```{r scatter1-solution, exercise.reveal_solution = FALSE}
tall <- players %>% 
  mutate(height = (12 * feet) + inches)
```
```{r scatter1-check}
grade_this_code()
```

Now we can go to the scatterplot. 

### Exercise 2: Your first scatterplot

With a scatterplot, we put what predicts the thing on the X axis, and the thing being predicted on the Y axis. `geom_point` works just like other geoms -- we need an x and a y and a data element. In this case, x is our height, y is our kills. 

```{r scatter2, exercise=TRUE, exercise.setup = "scatter-load-data", message=FALSE}
ggplot() + 
  geom_point(data=????, aes(x=??????, y=?????))
```
```{r scatter2-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(data=tall, aes(x=height, y=kills))
```
```{r scatter2-check}
grade_this_code()
```

Let's talk about this. This ... doesn't seem right. Seems ... really weird, no? 

We have two problems -- one is obvious, one is not immediately. First, the obvious one: Can you be zero inches tall and ... exist? Let alone play college volleyball? So obviously we have some missing data. The second issue, which is not as obvious, is what else could determine kills? What happens if you're 7 feet tall but ride the bench? Are you still racking up kills? No.

### Exercise 3: Filtering and trying again

Let's fix this by filtering some things out. First, we need to filter out people who have a height of 0. Second, the mean number of sets played by all players who played more than 0 sets is 65. Let's filter only players who played more than 65 sets. We'll call it `newtall`. 

```{r scatter4, exercise=TRUE, exercise.setup = "scatter-load-data", message=FALSE}
??????? <- tall %>%
  filter(
    height > ? &
      s > ??
         )
```
```{r scatter4-solution, exercise.reveal_solution = FALSE}
newtall <- tall %>%
  filter(
    height > 0 &
      s > 65
         )
```
```{r scatter4-check}
grade_this_code()
```

And let's chart that new dataframe. 

```{r scatter5, exercise=TRUE, exercise.setup = "scatter-load-data", message=FALSE}
ggplot() + 
  geom_point(data=???????, aes(x=??????, y=?????))
```
```{r scatter5-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(data=newtall, aes(x=height, y=kills))
```
```{r scatter5-check}
grade_this_code()
```

Better, but woof is there a big difference starting around 67 inches -- or about 5'7". There are players who are 5'7" who have zero kills and players the same height with more than 375. There's a player who is 6'10" who has fewer kills than a player who is 5'6". 

So is height predictive? We can add a new geom to this, called `geom_smooth` that will add a regression line -- a prediction line -- that is all but identical to our `geom_point`. The only thing added is the `method="lm"`.

```{r scatter6, exercise=TRUE, exercise.setup = "scatter-load-data", message=FALSE}
ggplot() + 
  geom_point(data=newtall, aes(x=height, y=kills)) +
  geom_smooth(data=newtall, aes(x=height, y=kills), method="lm")
```
```{r scatter6-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(data=newtall, aes(x=height, y=kills)) +
  geom_smooth(data=newtall, aes(x=height, y=kills), method="lm")
```
```{r scatter6-check}
grade_this_code()
```

A straight line across is bad. It means there's no pattern here. But there is a line that would indicate that height is somewhat predictive. But how predictive? Can we measure that? 

### Exercise 4: Regressions are back

Of course we can. We can apply a linear model to this -- remember Chapter 9? We're going to create an object called fit, and then we're going to put into that object a linear model -- `lm` -- and the way to read this is "wins are predicted by opponent threes". Then we just want the summary of that model.

```{r scatter7, exercise=TRUE, exercise.setup = "scatter-load-data", message=FALSE}
fit <- lm(????? ~ ??????, data = newtall)
summary(fit)
```
```{r scatter7-solution, exercise.reveal_solution = FALSE}
fit <- lm(kills ~ height, data = newtall)
summary(fit)
```
```{r scatter7-check}
grade_this_code()
```

Remember from Chapter 9: There's just a few things you really need.

The first thing: R-squared. In this case, the Adjusted R-squared value is 0.3022, which we can interpret as the the height of the player predicts about 30 percent of the variance in kills. 

Second: The P-value. We want anything less than .05. If it's above .05, the difference between them is not statistically significant -- it's probably explained by random chance. In our case, we have < 2.2e-16. So it's not random. Height really does impact kills -- but we kinda knew that already. 

Normally, we'd stop here, but let's look at the third element: The coefficient. In this case, the coefficient for height is 21.003. What this model predicts, given that and the intercept of -1348.654, is this: Every player starts about 1350 kills in the hole. For every inch they grow, they get 21 kills. Remember your algebra and y = mx + b. In this case, y is the wins, m is the coefficient, x is the number of threes given up and b is the intercept. 

Let's use Madi Kubik as an example. She's 6'3" on the roster, or 75 inches.  

y = 21.003 * 75 - 1348.654 or 226.5 kills. 

How many wins did Kubik have last seaon? 412. 

If we built a model to predict kills based purely on height, we missed Kubik's actual results by almost a factor of 2. Not great. 

**This is an important lesson: Not all scatterplots work. Not all scatterplots tell a clear story. If you try a scatterplot and get this level of inconclusiveness, you need to try something else.**