---
title: "Sports Data Lesson 20: Bubble charts"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to show relationships, but now with magnitude.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(ggrepel)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

The goal of this lesson is to introduce you to bubble charts, an advanced form of scatterplot that incorporates a third dimension of data through the size of the plotted points. By the end of this tutorial, you'll understand when to use bubble charts, how to create them using ggplot2, and how to interpret the complex relationships they reveal. You'll learn to prepare data for bubble charts, plot points with varying sizes to represent an additional variable, add reference lines for context, and use ggrepel to label points without overlap. We'll apply these concepts to real college football data, analyzing the relationship between offensive and defensive performance while also considering team wins. This lesson will enhance your ability to visualize multidimensional data relationships in sports, a crucial skill for understanding complex performance factors and making nuanced comparisons between teams.

## The Basics

Here is the real talk: Bubble charts are hard. The reason they are hard is not because of the code, or the complexity or anything like that. They're a scatterplot with magnitude added -- the size of the dot in the scatterplot has meaning. The hard part is seeing when a bubble chart works and when it doesn't. 

If you want to see it work spectacularly well, [watch a semi-famous Ted Talk](https://www.youtube.com/watch?v=hVimVzgtD6w) by Hans Rosling from 2006 where bubble charts were the centerpiece. It's worth watching. It'll change your perspective on the world. No seriously. It will.

And since then, people have wanted bubble charts. And we're back to the original problem: They're hard. There's a finite set of circumstances where they work. 

First, I'm going to show you an example of them not working to illustrate the point. 

For this example, I want to look at where Big Ten teams compare to the rest of college football last season. Is the Big Ten's reputation for tough games and defenses earned? Can we see patterns in good team vs bad teams? 

I'm going to create a scatterplot with offensive yards per play on the X axis and defensive yards per play on the y axis. We can then divide the grid into four quadrants. Teams with high yards per offensive play and low defensive yards per play are teams with good offenses and good defenses. The opposite means bad defense, bad offense. Then, to drive the point home, I'm going to make the dot the size of the total wins on the season -- the bubble in my bubble charts. 

We'll need the tidyverse and a library called ggrepel that we'll talk about later.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
library(ggrepel)
```
```{r load-tidyverse-solution}
library(tidyverse)
library(ggrepel)
```
```{r load-tidyverse-check}
grade_this_code()
```

And the data.

```{r bubble-load-data, message=FALSE, warning=FALSE}
logs <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballlogs24.csv")

winlosslogs <- logs |> 
  mutate(
    wins = case_when(
      grepl("W", Outcome) ~ 1, 
      grepl("L", Outcome) ~ 0)
  )
    
ypp <- winlosslogs |> 
  group_by(Team, Conference) |> 
  summarise(
    TotalPlays = sum(OffensivePlays), 
    TotalYards = sum(OffensiveYards), 
    DefensivePlays = sum(DefPlays), 
    DefensiveYards = sum(DefYards), 
    TotalWins = sum(wins)) |> 
  mutate(
    OffensiveYPP = TotalYards/TotalPlays, 
    DefensiveYPP = DefensiveYards/DefensivePlays)

bigten <- ypp |> filter(Conference == "Big Ten Conference")
```
```{r bubble-load-data-exercise, exercise = TRUE}
logs <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballlogs24.csv")
```
```{r bubble-load-data-exercise-solution}
logs <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballlogs24.csv")
```
```{r bubble-load-data-exercise-check}
grade_this_code()
```

To do this, I've got some work to do. First, I need to mutate the outcomes of the games into 1s and 0s so I can add up the wins. 

```{r bubble1, exercise=TRUE, exercise.setup = "bubble-load-data", message=FALSE}
winlosslogs <- logs |> 
  mutate(
    wins = case_when(
      grepl("W", Outcome) ~ 1, 
      grepl("L", Outcome) ~ 0)
)
```
```{r bubble1-solution, exercise.reveal_solution = FALSE}
winlosslogs <- logs |> 
  mutate(
    wins = case_when(
      grepl("W", Outcome) ~ 1, 
      grepl("L", Outcome) ~ 0)
)
```
```{r bubble1-check}
grade_this_code()
```

Now I have some more work to do. 

My football logs data has the yards per play of each game, and I could average those together and get something very close to what I'm going to do, but averaging each games yards per play is not the same thing as calculating it, so we're going to calculate it. 

I'm going to give you the data. It's pretty straightforward. We're making a new dataframe called ypp. It's going to group and sum wins, plays and yards -- offensive and defensive -- so we can divide them and get the season long yards per play on both offense and defense.

```{r bubble2, exercise=TRUE, exercise.setup = "bubble-load-data", message=FALSE}
ypp <- winlosslogs |> 
  group_by(Team, Conference) |> 
  summarise(
    TotalPlays = sum(OffensivePlays), 
    TotalYards = sum(OffensiveYards), 
    DefensivePlays = sum(DefPlays), 
    DefensiveYards = sum(DefYards), 
    TotalWins = sum(wins)) |> 
  mutate(
    OffensiveYPP = TotalYards/TotalPlays, 
    DefensiveYPP = DefensiveYards/DefensivePlays)
```
```{r bubble2-solution, exercise.reveal_solution = FALSE}
ypp <- winlosslogs |> 
  group_by(Team, Conference) |> 
  summarise(
    TotalPlays = sum(OffensivePlays), 
    TotalYards = sum(OffensiveYards), 
    DefensivePlays = sum(DefPlays), 
    DefensiveYards = sum(DefYards), 
    TotalWins = sum(wins)) |> 
  mutate(
    OffensiveYPP = TotalYards/TotalPlays, 
    DefensiveYPP = DefensiveYards/DefensivePlays)
```
```{r bubble2-check}
grade_this_code()
```

A bubble chart is just a scatterplot with one additional element in the aesthetic -- a size. Here's the scatterplot version. 

```{r bubble3, exercise=TRUE, exercise.setup = "bubble-load-data", message=FALSE}
ggplot() + geom_point(data=ypp, aes(x=OffensiveYPP, y=DefensiveYPP))
```
```{r bubble3-solution, exercise.reveal_solution = FALSE}
ggplot() + geom_point(data=ypp, aes(x=OffensiveYPP, y=DefensiveYPP))
```
```{r bubble3-check}
grade_this_code()
```

Looks kind of random, eh? In this case, that's not that bad because we're not claiming a relationship. We're saying the location on the chart has meaning. 

So, do teams on the bottom right -- good offense, good defense -- win more games? 

### Exercise 1: The bubble chart

Let's add the size element. From our data work above, we want to add the column that has the total wins in it. That's a hint, look at the code.

```{r bubble4, exercise=TRUE, exercise.setup = "bubble-load-data", message=FALSE}
ggplot() + 
  geom_point(
    data=____, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=____)
    )
```
```{r bubble4-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(
    data=ypp, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=TotalWins)
    )
```
```{r bubble4-check}
grade_this_code()
```

What does this chart tell you? We can see a general pattern that there are more big dots on the bottom right than the upper left. 

### Exercise 2: Adding transparency and scale

But we can make this more readable by adding an `alpha` element outside the aesthetic -- `alpha` in this case is transparency -- and we can manually change the size of the dots by adding `scale_size` and a `range`. Let's try an `alpha` of .3. 

```{r bubble5, exercise=TRUE, exercise.setup = "bubble-load-data", message=FALSE}
ggplot() + 
  geom_point(
    data=ypp, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=____),
    alpha = ____) + 
  scale_size(range = c(3, 8), name="Wins")
```
```{r bubble5-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(
    data=ypp, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=TotalWins),
    alpha = .3) + 
  scale_size(range = c(3, 8), name="Wins")
```
```{r bubble5-check}
grade_this_code()
```

Better?

### Exercise 3: Layering

And by now, you now know to add in the Big Ten as a layer, I would hope.

```{r bubble6, exercise=TRUE, exercise.setup = "bubble-load-data", message=FALSE}
bigten <- ypp |> filter(Conference == "____ ____ ____")

ggplot() + 
  geom_point(
    data=ypp, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=____), 
    color="grey", 
    alpha=.3) + 
  geom_point(
    data=bigten, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=____), 
    color="red")
```
```{r bubble6-solution, exercise.reveal_solution = FALSE}
bigten <- ypp |> filter(Conference == "Big Ten Conference")

ggplot() + 
  geom_point(
    data=ypp, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=TotalWins), 
    color="grey", 
    alpha=.3) + 
  geom_point(
    data=bigten, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=TotalWins), 
    color="red")
```
```{r bubble6-check}
grade_this_code()
```

It's clearer where Big Ten teams are. But which Big Ten team is which?

### Exercise 4: Adding reference lines 

Let's add some things to this chart to help us out. First, let's add lines that show us the average of all teams for those two metrics. So first, we need to calculate those. Because I have grouped data, it's going to require me to ungroup it so I can get just the total average of those two numbers. 

```{r bubble7, exercise=TRUE, exercise.setup = "bubble-load-data", message=FALSE}
ypp |> 
  ungroup() |> 
  summarise(
    offense = mean(OffensiveYPP), 
    defense = mean(DefensiveYPP)
    )
```
```{r bubble7-solution, exercise.reveal_solution = FALSE}
ypp |> 
  ungroup() |> 
  summarise(
    offense = mean(OffensiveYPP), 
    defense = mean(DefensiveYPP)
    )
```
```{r bubble7-check}
grade_this_code()
```

Now we can use those averages to add two more geoms -- geom_vline and geom_hline, for vertical lines and horizontal lines. The defense is the horizontal line, the offense is the vertical line.

```{r bubble8, exercise=TRUE, exercise.setup = "bubble-load-data", message=FALSE}
ggplot() + 
  geom_point(
    data=ypp, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=____), 
    color="grey", 
    alpha=.3) + 
  geom_point(
    data=bigten, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=____), 
    color="red") + 
  geom_vline(xintercept = ____) + 
  geom_hline(yintercept = ____)
```
```{r bubble8-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(
    data=ypp, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=TotalWins), 
    color="grey", 
    alpha=.3) + 
  geom_point(
    data=bigten, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=TotalWins), 
    color="red") + 
  geom_vline(xintercept = 5.763149) + 
  geom_hline(yintercept = 5.601157)
```
```{r bubble8-check}
grade_this_code()
```

Getting better.

### Exercise 5: Adding names to the dots

Now it's time for ggrepel. What ggrepel does is help us label the dots without overwriting other labels. By importing ggrepel, we now have access to a new geom -- `geom_text_repel` -- which is a modification of `geom_text` that we've used before that automatically keeps labels from overlapping. What do we want to label? The Big Ten teams in our bigten dataframe. The three elements we need in `geom_text_repel` are an x and y coordinate and a label. The label, in this case, is the name of the Team.

```{r bubble9, exercise=TRUE, exercise.setup = "bubble-load-data", message=FALSE}
ggplot() + 
  geom_point(
    data=ypp, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=____), 
    color="grey", 
    alpha=.3) + 
  geom_point(
    data=bigten, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=____), 
    color="red") + 
  geom_vline(xintercept = ____) + 
  geom_hline(yintercept = ____) +
  geom_text_repel(
    data=bigten, 
    aes(x=OffensiveYPP, y=DefensiveYPP, label=____)
  )
```
```{r bubble9-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(
    data=ypp, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=TotalWins), 
    color="grey", 
    alpha=.3) + 
  geom_point(
    data=bigten, 
    aes(x=OffensiveYPP, y=DefensiveYPP, size=TotalWins), 
    color="red") + 
  geom_vline(xintercept = 5.763149) + 
  geom_hline(yintercept = 5.601157) +
  geom_text_repel(
    data=bigten, 
    aes(x=OffensiveYPP, y=DefensiveYPP, label=Team)
  )
```
```{r bubble9-check}
grade_this_code()
```

Where is Nebraska? And what story does that tell? 

## The Recap

In this lesson, we've explored the power and complexity of bubble charts for visualizing multidimensional relationships in sports data. We started by learning how to prepare data for bubble charts, including grouping and summarizing game-level statistics into team-level metrics. We then progressed from creating basic scatterplots to more complex bubble charts by incorporating a size aesthetic to represent wins. We covered important techniques such as using transparency (alpha) to manage overlapping points, scaling bubble sizes for better readability, and layering data to highlight specific groups like Big Ten teams. We also learned how to add context to our charts with reference lines representing average performance metrics. Finally, we introduced the ggrepel package to add non-overlapping labels to our chart, enhancing its interpretability. Throughout the lesson, we've seen how bubble charts can reveal complex relationships between offensive performance, defensive performance, and overall success in college football. Remember, while bubble charts are powerful tools for visualizing three-dimensional data, they require careful design and interpretation to effectively communicate insights. Always consider whether the added complexity of a bubble chart enhances or potentially obscures your data story.

## Terms to Know

- Bubble Chart: A variation of a scatterplot where a third variable is represented by the size of the plotted points or "bubbles".
- `geom_point()`: A ggplot2 function used to create scatterplots and bubble charts by plotting individual data points.
- `alpha`: A parameter in ggplot2 that controls the transparency of graphical elements, useful for managing overlapping points in bubble charts.
- `scale_size()`: A ggplot2 function used to adjust the range of sizes for bubbles in a bubble chart.
- `geom_vline()` and `geom_hline()`: ggplot2 functions used to add vertical and horizontal reference lines to a chart, respectively.
- `ggrepel`: An R package that provides geoms for ggplot2 to repel overlapping text labels.
- `geom_text_repel()`: A function from the ggrepel package used to add non-overlapping text labels to points in a chart.