---
title: "Sports Data Lesson 14: Stacked bar charts"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to add nuance to a bar chart.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

The goal of this lesson is to build upon your basic bar chart skills and introduce you to creating stacked bar charts using ggplot2. Stacked bar charts are powerful tools for visualizing the composition of a whole and comparing parts within it. By the end of this tutorial, you'll understand how to prepare your data for stacked bar charts, use the fill aesthetic to create layers within bars, and interpret the resulting visualizations. You'll learn how to create charts that show both total values and their components. This lesson will enhance your ability to create more nuanced and informative visualizations, allowing you to present complex relationships in sports data in a clear and compelling manner.

## The Basics

One of the elements of data visualization excellence is **inviting comparison**. Often that comes in showing **what proportion a thing is in relation to the whole thing**. With bar charts, we're showing magnitude of the whole thing. If we have information about the parts of the whole, **we can stack them on top of each other to compare them, showing both the whole and the components**. And it's a simple change to what we've already done. 

We're going to use a dataset of college football games from this season. 

Load the tidyverse.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
```
```{r load-tidyverse-solution}
library(tidyverse)
```
```{r load-tidyverse-check}
grade_this_code()
```

And the data.

```{r stacked-load-data, message=FALSE, warning=FALSE}
football <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballlogs24.csv")

big <- football |> 
  group_by(Conference, Team) |> 
  summarise(
    SeasonRushingYards = sum(RushingYds),
    SeasonPassingYards = sum(PassingYds),
  ) |> filter(Conference == "Big Ten Conference")

biglong <- big |> 
  pivot_longer(
    cols=starts_with("Season"), 
    names_to="Type", 
    values_to="Yards")
```
```{r stacked-load-data-exercise, exercise = TRUE}
football <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballlogs24.csv")
```
```{r stacked-load-data-exercise-solution}
football <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballlogs24.csv")
```
```{r stacked-load-data-exercise-check}
grade_this_code()
```

What we have here is every game in college football in 2022. Let's quick take a glimpse.

```{r glimpse-data, exercise=TRUE, exercise.setup = "stacked-load-data"}
glimpse(____)
```
```{r glimpse-data-solution}
glimpse(football)
```
```{r glimpse-data-check}
grade_this_code()
```

The question we want to answer is this: Who had the most prolific offenses in the Big Ten? And how did they get there? 

So to make this chart, we have to just add one thing to a bar chart like we did in the previous chapter. However, it's not that simple.

### Exercise 1: Preparing your data

We have game data, and we need season data. To get that, we need to do some group by and sum work. And since we're only interested in the Big Ten, we have some filtering to do too. For this, we're going to measure offensive production by rushing yards and passing yards. So if we have all the games a team played, and the rushing and passing yards for each of those games, what we need to do to get the season totals is just add them up. We'll put all of that into a new dataframe called big

```{r stacked-big, exercise=TRUE, exercise.setup = "stacked-load-data", message=FALSE}
____ <- football |> 
  group_by(Conference, Team) |> 
  summarise(
    SeasonRushingYards = sum(____),
    SeasonPassingYards = sum(____)
  ) |> 
  filter(Conference == "Big Ten Conference")
```
```{r stacked-big-solution, exercise.reveal_solution = FALSE}
big <- football |> 
  group_by(Conference, Team) |> 
  summarise(
    SeasonRushingYards = sum(RushingYds),
    SeasonPassingYards = sum(PassingYds)
  ) |> 
  filter(Conference == "Big Ten Conference")
```
```{r stacked-big-check}
grade_this_code()
```

By looking at this, we can see we got what we needed. We have 14 teams (though remember that head only shows you six) and numbers that look like season totals for yards. 

```{r head-data, exercise=TRUE, exercise.setup = "stacked-load-data"}
head(big)
```
```{r head-data-solution}
head(big)
```
```{r head-data-check}
grade_this_code()
```

Now, the problem we have is that ggplot wants long data and this data is wide. 

### Exercise 2: Making wide data long

Remember transforming data? Lesson 6? This is where that work is going to pay off. We need to pivot this data longer. In order to do that, we need to say which columns are being pivoted. Note in the head block above, we have four columns. Two columns are the Team and the Conference. The other columns start with the word Season. That, my friends, is a gigantic hint.

We're going to save this work to a new dataframe called biglong. 

```{r stacked-pivot, exercise=TRUE, exercise.setup = "stacked-load-data", message=FALSE}
____ <- ____ |> 
  pivot_longer(
    cols=starts_with("____"), 
    names_to="Type", 
    values_to="Yards")
```
```{r stacked-pivot-solution, exercise.reveal_solution = FALSE}
biglong <- big |> 
  pivot_longer(
    cols=starts_with("Season"), 
    names_to="Type", 
    values_to="Yards")
```
```{r stacked-pivot-check}
grade_this_code()
```

If this worked, you'll have two rows for each team: One for rushing yards, one for passing yards. This is what ggplot needs. 

```{r head2-data, exercise=TRUE, exercise.setup = "stacked-load-data"}
head(biglong)
```
```{r head2-data-solution}
head(biglong)
```
```{r head2-data-check}
grade_this_code()
```

### Exercise 3: Making your first plot

Building on what we learned in the last chapter, we know we can turn this into a bar chart with an x value, a weight and a geom_bar. What we are going to add is a `fill`. The `fill` will stack bars on each other based on which element it is. In this case, we can fill the bar by Type, which means it will stack the number of rushing yards on top of passing yards and we can see how they compare. 

```{r stacked-plot1, exercise=TRUE, exercise.setup = "stacked-load-data", message=FALSE}
ggplot() + 
  geom_bar(data=____, aes(x=____, weight=Yards, fill=____)) + 
  coord_flip()
```
```{r stacked-plot1-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bar(data=biglong, aes(x=Team, weight=Yards, fill=Type)) + 
  coord_flip()
```
```{r stacked-plot1-check}
grade_this_code()
```

What's the problem with this chart? 

There's a couple of things, one of which we'll deal with now: The ordering is alphabetical (from the bottom up). So let's `reorder` the teams by Yards.

### Exercise 4: Reordering your bars

```{r stacked-plot2, exercise=TRUE, exercise.setup = "stacked-load-data", message=FALSE}
ggplot() + 
  geom_bar(data=____, aes(x=reorder(____, ____), weight=Yards, fill=____)) + 
  coord_flip()
```
```{r stacked-plot2-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bar(data=biglong, aes(x=reorder(Team, Yards), weight=Yards, fill=Type)) + 
  coord_flip()
```
```{r stacked-plot2-check}
grade_this_code()
```

And just like that ... Penn State and Oregon are out on top. Where is Nebraska?

## The Recap

In this lesson, we've expanded our data visualization skills by learning how to create stacked bar charts using ggplot2. We started by preparing our data, using group_by() and summarise() to aggregate game-level data into season totals, and then transforming it from wide to long format using pivot_longer(). This data preparation step is crucial for creating effective stacked bar charts. We then built on our basic bar chart knowledge, adding the fill aesthetic to create layers within our bars that represent different components of the whole (in this case, rushing and passing yards). We also reinforced the importance of reordering our data for more meaningful visualizations. Through this process, we've learned how to create charts that not only show total offensive production for Big Ten football teams, but also reveal the balance between rushing and passing yards for each team. This type of visualization allows for rich comparisons and insights, demonstrating the power of stacked bar charts in sports data analysis. Remember, effective data visualization often requires careful data preparation and thoughtful design choices to clearly communicate your insights.

## Terms to Know

- Stacked Bar Chart: A type of bar chart where bars are divided into segments, each representing a subcategory of the main category. The total height of the bar represents the sum of all subcategories.
- `pivot_longer()`: A function from the tidyr package used to transform data from wide to long format, which is often necessary for creating stacked bar charts in ggplot2.
- `fill`: An aesthetic in ggplot2 used to assign different colors to subcategories within a stacked bar chart, allowing for visual distinction between components.
- `reorder()`: A function used to change the order of categorical variables in a plot, often used in bar charts to sort bars by their total values for better readability.
- Long format: A data structure where each row represents a single observation and each column represents a variable. This format is typically required for creating stacked bar charts in ggplot2.
