---
title: "Lesson 14: Circular bar charts"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to add some visual interestingness to your bar charts.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

# Circular bar plots

## The basics

At the 27:36 mark in the [Half Court Podcast](https://www.omaha.com/sports/podcasts/half-court-press/half-court-press-creighton-cruises-in-opener-nebraska-stunned-in/article_67081a35-3a8f-5e9e-ae67-e88fcacbb362.html), former Omaha World Herald Writer and CoJMC grad Chris Heady said "November basketball doesn't matter, but it shows you where you are."

It's a tempting phrase to believe, especially since this aired a day after Nebraska lost the first game of the Fred Hoiberg era at home to a baseball school, UC Riverside. And it wasn't close. The Huskers, because of a total roster turnover, were a complete mystery before the game. And what happened during it wasn't pretty, so there was a little soul searching going on in Lincoln.

But does November basketball really not matter?

Let's look, using a new form of chart called a circular bar plot. It's a chart type that combines several forms we've used before: bar charts to show magnitude, stacked bar charts to show proportion, but we're going to add bending the chart around a circle to add some visual interstingness to it. We're also going to use time as an x-axis value to make a not subtle circle of time reference -- a common technique with circular bar charts. 

We'll use a dataset of every college basketball game last season. Standard message applies. 

```{r echo=FALSE, class.output="bg-info", results="asis",  message=FALSE,  warning=FALSE}
library(downloadthis)
library(glue)

dllink <- download_link(
  link = "http://mattwaite.github.io/sportsdatafiles/logs21.csv",
  button_label = "Download csv file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

glue("<pre><p><strong>For this walkthrough:</strong></p><p>{dllink}</p></pre>")

```

Load the tidyverse and a library called `lubridate`, which we'll use later. It's got some very handy functions for working with dates.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
library(lubridate)
```
```{r load-tidyverse-solution}
library(tidyverse)
library(lubridate)
```
```{r load-tidyverse-check}
grade_this_code()
```

And the data.

```{r circular-load-data, message=FALSE, warning=FALSE}
logs <- read_csv("http://mattwaite.github.io/sportsdatafiles/logs21.csv")

winlosslogs <- logs %>% 
  filter(Date < as.Date("2021-03-07")) %>%
  mutate(winloss = case_when(
  grepl("W", W_L) ~ 1, 
  grepl("L", W_L) ~ 0)
) 

dates <- winlosslogs %>% group_by(Date, Conference) %>% summarise(wins = sum(winloss))

p5 <- c("SEC", "Big Ten", "Pac-12", "Big 12", "ACC")

months <- winlosslogs %>% 
  mutate(month = floor_date(Date, unit="months")) %>% 
  group_by(month, Conference) %>% 
  summarise(wins=sum(winloss)) %>% 
  filter(Conference %in% p5) 

weeks <- winlosslogs %>% 
  mutate(week = floor_date(Date, unit="weeks")) %>% 
  group_by(week, Conference) %>% 
  summarise(wins=sum(winloss)) %>% 
  filter(Conference %in% p5) 
```
```{r circular-load-data-exercise, exercise = TRUE}
logs <- read_csv("http://mattwaite.github.io/sportsdatafiles/logs21.csv")
```
```{r circular-load-data-exercise-solution}
logs <- read_csv("http://mattwaite.github.io/sportsdatafiles/logs21.csv")
```
```{r circular-load-data-exercise-check}
grade_this_code()
```

What we have here is every game in college basketball in 2020-2021. Let's quick take a glimpse.

```{r glimpse-data, exercise=TRUE, exercise.setup = "circular-load-data"}
glimpse(????)
```
```{r glimpse-data-solution}
glimpse(logs)
```
```{r glimpse-data-check}
grade_this_code()
```

## Does November basketball matter?

So let's test the notion of November Basketball Doesn't Matter. What matters in basketball? Let's start simple: Wins.

Sports Reference's win columns are weird, so we need to scan through them and find W and L and we'll give them numbers using `case_when`. I'm also going to filter out tournament basketball. I'm going to give you this because we haven't done a lot of this, but pay attention to the patterns. 

```{r circular-data, exercise=TRUE, exercise.setup = "circular-load-data", message=FALSE}
winlosslogs <- logs %>% 
  filter(Date < as.Date("2021-03-07")) %>%
  mutate(winloss = case_when(
  grepl("W", W_L) ~ 1, 
  grepl("L", W_L) ~ 0)
) 
```
```{r circular-data-solution, exercise.reveal_solution = FALSE}
winlosslogs <- logs %>% 
  filter(Date < as.Date("2021-03-07")) %>%
  mutate(winloss = case_when(
  grepl("W", W_L) ~ 1, 
  grepl("L", W_L) ~ 0)
) 
```
```{r circular-data-check}
grade_this_code()
```

Now we have a new column called winloss, where if it's a win, winloss will say 1. If it's a loss, winloss will say 0. We can group by date and conference and sum up the wins. How many wins by day does each conference get?

### Exercise 1: Wins by day

For this, we're going to create a new dataframe called dates. In it, we're going to group by Date and Conference, and in summarize, we'll add up all the winloss 1s and 0s. 

```{r circular-day, exercise=TRUE, exercise.setup = "circular-load-data", message=FALSE}
????? <- winlosslogs %>% group_by(????, Conference) %>% summarise(wins = sum(???????))
```
```{r circular-day-solution, exercise.reveal_solution = FALSE}
dates <- winlosslogs %>% group_by(Date, Conference) %>% summarise(wins = sum(winloss))
```
```{r circular-day-check}
grade_this_code()
```

Let's take a peek at that.

```{r head-data, exercise=TRUE, exercise.setup = "circular-load-data"}
head(dates)
```
```{r head-data-solution}
head(dates)
```
```{r head-data-check}
grade_this_code()
```

Looks like what we need. A date, a Conference and a number. 

Earlier, we did stacked bar charts. We have what we need to do that now.

### Exercise 2: Make a stacked bar chart.

For the stacked bar, the fill is going to be the Conference and the x-axis will be the Date.

```{r circular-plot1, exercise=TRUE, exercise.setup = "circular-load-data", message=FALSE}
ggplot() + 
  geom_bar(data=?????, aes(x=????, weight=wins, fill=??????????))
```
```{r circular-plot1-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bar(data=dates, aes(x=Date, weight=wins, fill=Conference))
```
```{r circular-plot1-check}
grade_this_code()
```

Eeek. This is already looking not great. But to make it a circular bar chart, we add `coord_polar()` to our chart.

```{r circular-plot2, exercise=TRUE, exercise.setup = "circular-load-data", message=FALSE}
ggplot() + 
  geom_bar(data=?????, aes(x=????, weight=wins, fill=??????????)) +
  coord_polar()
```
```{r circular-plot2-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bar(data=dates, aes(x=Date, weight=wins, fill=Conference)) +
  coord_polar()
```
```{r circular-plot2-check}
grade_this_code()
```

Based on that, the day is probably too thin a slice, and there's way too many conferences in college basketball. 

### Exercise 3: Better slices of data

Let's group this by months and filter out all but the power five conferences. Remember filtering in Lesson 5? We're going to use %in% again to make filtering the Power 5 easy. 

Then we're going to make a dataframe called months. In months, we'll create a new column called month. To get months, we're going to use a function in the library `lubridate` called `floor_date`, which combine with mutate will give us a field of just months. We'll then group by that new column and the Conference. That'll give us the number of wins each conference gets each month. 

```{r circular-month, exercise=TRUE, exercise.setup = "circular-load-data", message=FALSE}
p5 <- c("SEC", "Big Ten", "Pac-12", "Big 12", "ACC")

?????? <- winlosslogs %>% 
  mutate(month = floor_date(Date, unit="months")) %>% 
  group_by(?????, Conference) %>% 
  summarise(wins=sum(???????)) %>% 
  filter(Conference %in% p5) 
```
```{r circular-month-solution, exercise.reveal_solution = FALSE}
p5 <- c("SEC", "Big Ten", "Pac-12", "Big 12", "ACC")

months <- winlosslogs %>% 
  mutate(month = floor_date(Date, unit="months")) %>% 
  group_by(month, Conference) %>% 
  summarise(wins=sum(winloss)) %>% 
  filter(Conference %in% p5) 
```
```{r circular-month-check}
grade_this_code()
```

And let's take a peek at that:

```{r head-data2, exercise=TRUE, exercise.setup = "circular-load-data"}
head(months)
```
```{r head-data2-solution}
head(months)
```
```{r head-data2-check}
grade_this_code()
```

### Exercise 4: Our second crack at a circular bar chart.

Now we can use month instead of Date to make our circular bar chart of wins by month in the Power Five.

```{r circular-plot3, exercise=TRUE, exercise.setup = "circular-load-data", message=FALSE}
ggplot() + 
  geom_bar(data=??????, aes(x=?????, weight=wins, fill=??????????)) +
  coord_polar()
```
```{r circular-plot3-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bar(data=months, aes(x=month, weight=wins, fill=Conference)) +
  coord_polar()
```
```{r circular-plot3-check}
grade_this_code()
```

Yikes. That looks a lot like a broken pie chart. So months are too thick of a slice. Let's use weeks in our floor date to see what that gives us.

### Exercise 5: Weeks not months.

Let's make a dataframe called weeks. The rest is similar to what you've done before. 

```{r circular-week, exercise=TRUE, exercise.setup = "circular-load-data", message=FALSE}
p5 <- c("SEC", "Big Ten", "Pac-12", "Big 12", "ACC")

?????? <- winlosslogs %>% 
  mutate(week = floor_date(Date, unit="weeks")) %>% 
  group_by(?????, Conference) %>% 
  summarise(wins=sum(???????)) %>% 
  filter(Conference %in% p5) 
```
```{r circular-week-solution, exercise.reveal_solution = FALSE}
p5 <- c("SEC", "Big Ten", "Pac-12", "Big 12", "ACC")

weeks <- winlosslogs %>% 
  mutate(week = floor_date(Date, unit="weeks")) %>% 
  group_by(week, Conference) %>% 
  summarise(wins=sum(winloss)) %>% 
  filter(Conference %in% p5) 
```
```{r circular-week-check}
grade_this_code()
```

And now we'll make a chart.

### Exercise 6: One last chart

```{r circular-plot4, exercise=TRUE, exercise.setup = "circular-load-data", message=FALSE}
ggplot() + 
  geom_bar(data=??????, aes(x=????, weight=wins, fill=??????????)) +
  coord_polar()
```
```{r circular-plot4-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bar(data=weeks, aes(x=week, weight=wins, fill=Conference)) +
  coord_polar()
```
```{r circular-plot4-check}
grade_this_code()
```

That looks better. But what does it say? Does November basketball matter? What this is saying is ... yeah, it kinda does. The reason? Lots of wins get piled up in November and December, during non-conference play. So if you are a team with NCAA tournament dreams, you need to win games in November to make sure your tournament resume is where it needs to be come March. Does an individual win or loss matter? Probably not. But your record in November does.