---
title: "Sports Data Lesson 25: Faceting"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to make a lot of graphics with little code.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

The goal of this lesson is to introduce you to faceting, a powerful technique in ggplot2 for creating multiple related charts in a single visualization. By the end of this tutorial, you'll understand how to use facet_wrap() and facet_grid() to create small multiples, allowing you to compare trends across different categories or groups. You'll learn when to use faceting, how to implement it with various chart types, and how to customize faceted plots for better readability. We'll apply these concepts to real college football data, analyzing offensive performance across Big Ten teams. This lesson will enhance your ability to present complex, multi-dimensional data in a clear and intuitive format, enabling viewers to easily compare patterns and trends across different subsets of your data.

## The Basics

Sometimes the easiest way to spot a trend is to chart a bunch of small things side by side. Edward Tufte, one of the most well known data visualization thinkers on the planet, calls this "small multiples" where ggplot calls this a facet wrap or a facet grid, depending. 

Who had the best offense in the Big Ten last season? We could answer this a number of ways, but the best way to show people would be visually. Let's use Small Multiples.

As always, we start with libraries. 

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
```
```{r load-tidyverse-solution}
library(tidyverse)
```
```{r load-tidyverse-check}
grade_this_code()
```

We're going to use the logs of college football games last season. 

```{r facet-load-data, message=FALSE, warning=FALSE}
logs <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballlogs23.csv")

big10 <- logs |> filter(Conference == "Big Ten Conference")
```
```{r facet-load-data-exercise, exercise = TRUE}
logs <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballlogs23.csv")
```
```{r facet-load-data-exercise-solution}
logs <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballlogs23.csv")
```
```{r facet-load-data-exercise-check}
grade_this_code()
```

Let's narrow our pile and look just at the Big Ten.

```{r facet1, exercise=TRUE, exercise.setup = "facet-load-data", message=FALSE}
big10 <- logs |> filter(Conference == "Big Ten Conference")
```
```{r facet1-solution, exercise.reveal_solution = FALSE}
big10 <- logs |> filter(Conference == "Big Ten Conference")
```
```{r facet1-check}
grade_this_code()
```

The first thing we can do is look at a line chart, like we have done in previous lessons.

### Exercise 1: Make a line chart

We'll do something simple like a line chart of OffenseAvg -- the yards per play on offense -- and we're going to group on Team.

```{r facet2, exercise=TRUE, exercise.setup = "facet-load-data", message=FALSE}
ggplot() + 
  geom_____(data=____, aes(x=Date, y=____, group=____))
```
```{r facet2-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_line(data=big10, aes(x=Date, y=OffenseAvg, group=Team))
```
```{r facet2-check}
grade_this_code()
```

And, not surprisingly, we get a hairball. Who is the best offense week in and out in the Big Ten? We could color certain lines, but that would limit us to focus on one team. **What if we did all of them at once?** We do that with a `facet_wrap`. 

Will that give us an answer?

### Exercise 2: The facet wrap

The only thing we MUST pass into a `facet_wrap` is what thing we're going to separate them out by. In this case, we precede that field with a tilde, so in our case we want the Team field.

```{r facet3, exercise=TRUE, exercise.setup = "facet-load-data", message=FALSE}
ggplot() + 
  geom_____(data=____, aes(x=Date, y=____, group=____)) + 
  facet_wrap(~____)
```
```{r facet3-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_line(data=big10, aes(x=Date, y=OffenseAvg, group=Team)) + 
  facet_wrap(~Team)
```
```{r facet3-check}
grade_this_code()
```

Answer: Not immediately clear, but we can look at this and analyze it. We could add a piece of annotation to help us out.

### Exercise 3: Put everyone in context

Let's add a line to every chart that's the conference average shooting percentage. 

```{r facet4, exercise=TRUE, exercise.setup = "facet-load-data", message=FALSE}
big10 |> summarise(mean(OffenseAvg))
```
```{r facet4-solution, exercise.reveal_solution = FALSE}
big10 |> summarise(mean(OffenseAvg))
```
```{r facet4-check}
grade_this_code()
```

Copy that number, and we're going to add it into the facet chart with a `geom_hline`.

```{r facet5, exercise=TRUE, exercise.setup = "facet-load-data", message=FALSE}
ggplot() + 
  geom_hline(yintercept=____, color="blue") + 
  geom_____(data=____, aes(x=Date, y=____, group=____)) + 
  facet_wrap(~____)
```
```{r facet5-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_hline(yintercept=5.239205, color="blue") + 
  geom_line(data=big10, aes(x=Date, y=OffenseAvg, group=Team)) + 
  facet_wrap(~Team)
```
```{r facet5-check}
grade_this_code()
```

What do you see here? How do teams compare? How do they change over time? I'm not asking you these questions because they're an assignment -- I'm asking because that's exactly what this chart helps answer. Your brain will immediately start making those connections. 

## Facet grid vs facet wraps

Facet grids allow us to put teams on the same plane, versus just repeating them. And we can specify that plane as vertical or horizontal. For example, here's our chart from above, but using facet_grid to stack them.

### Exercise 4: Facet grids

```{r facet6, exercise=TRUE, exercise.setup = "facet-load-data", message=FALSE}
ggplot() + 
  geom_hline(yintercept=____, color="blue") + 
  geom_____(data=____, aes(x=Date, y=____, group=____)) + 
  facet_grid(Team ~ .)
```
```{r facet6-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_hline(yintercept=5.239205, color="blue") + 
  geom_line(data=big10, aes(x=Date, y=OffenseAvg, group=Team)) + 
  facet_grid(Team ~ .)
```
```{r facet6-check}
grade_this_code()
```

And here they are next to each other:

```{r facet7, exercise=TRUE, exercise.setup = "facet-load-data", message=FALSE}
ggplot() + 
  geom_hline(yintercept=____, color="blue") + 
  geom_____(data=____, aes(x=Date, y=____, group=____)) + 
  facet_grid(. ~ Team)
```
```{r facet7-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_hline(yintercept=5.239205, color="blue") + 
  geom_line(data=big10, aes(x=Date, y=OffenseAvg, group=Team)) + 
  facet_grid(. ~ Team)
```
```{r facet7-check}
grade_this_code()
```

Two notes:

1. We'd have some work to do with the labeling on this -- we'll get to that -- but you can see where this is valuable comparing a group of things. One warning: Don't go too crazy with this or it loses it's visual power.
2. There's nothing special about a line chart with faceting. You can facet any kind of ggplot chart. 

## The Recap

In this lesson, we've explored the power of faceting in ggplot2 to create small multiples, a technique that allows for easy comparison across multiple categories or groups. We started by creating a basic line chart of offensive performance for Big Ten football teams, then transformed it into a faceted plot using facet_wrap(). We learned how to add context to our faceted plots by including a reference line for the conference average. We then explored the difference between facet_wrap() and facet_grid(), understanding how each can be used to arrange our small multiples either in a grid or in rows/columns. We saw how faceting can reveal patterns and trends that might be obscured in a single, overlapping chart. Throughout the lesson, we applied these techniques to real college football data, demonstrating how faceting can provide insights into team performance over time. Remember, while faceting is a powerful tool, it's important to use it judiciously to maintain clarity in your visualizations. As we progressed, we also touched on the importance of proper labeling and the versatility of faceting across different chart types.

## Terms to Know

- Faceting: A technique in ggplot2 for creating multiple related charts in a single visualization, allowing for easy comparison across categories or groups.
- Small Multiples: A series of similar graphs or charts using the same scale and axes, allowing for comparison across different categories or time periods.
- `facet_wrap()`: A ggplot2 function that wraps a series of plots into a rectangular layout.
- `facet_grid()`: A ggplot2 function that creates a grid of plots with fixed horizontal and vertical axes.
- Tilde (~): In the context of faceting, used to specify the variable(s) by which to facet the plot.