---
title: "Sports Data Lesson 13: Bar charts"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to start turning data into graphics.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(scales)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

The goal of this lesson is to introduce you to creating bar charts using ggplot2, one of the most powerful and flexible data visualization tools in R. By the end of this tutorial, you'll understand the basic structure of a ggplot command, how to create simple bar charts, and how to customize them to better represent your data. You'll learn how to filter data to focus on specific subsets, reorder bars for clearer presentation, use layering to highlight particular data points, and adjust the orientation of your charts. These skills will form the foundation for creating more complex and insightful visualizations as you progress in your sports data analysis journey. Remember, effective data visualization is key to communicating your findings clearly and compellingly in sports analytics.

## The Basics

With `ggplot2`, we dive into the world of programmatic data visualization. The `ggplot2` library implements something called the grammar of graphics. The main concepts are: 

* aesthetics - which in this case means the data which we are going to plot
* geometries - which means the shape the data is going to take
* scales - which means any transformations we might make on the data
* facets - which means how we might graph many elements of the same dataset in the same space
* layers - which means how we might lay multiple geometries over top of each other to reveal new information.

Hadley Wickham, who is behind all of the libraries we have used in this course to date, wrote about his layered grammar of graphics in [this 2009 paper that is worth your time to read](https://byrneslab.net/classes/biol607/readings/wickham_layered-grammar.pdf). 

Let's dive in using data on college softball in 2023. This workflow will represent a clear picture of what your work in this class will be like for much of the rest of the semester. One way to think of this workflow is that your R Notebook is now your digital sketchbook, where you will try different types of visualizations to find ones that work. Then, you will write the code that adds necessary and required parts to finish it. 

**For purposes of this exercise, you don't need to download this. The data is included here if you want to try this in your own notebook.**

```{r echo=FALSE, class.output="bg-info", results="asis",  message=FALSE,  warning=FALSE}
library(downloadthis)
library(glue)

dllink <- download_link(
  link = "https://mattwaite.github.io/sportsdatafiles/ncaa_softball_2023.csv",
  button_label = "Download csv file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

glue("<pre><p><strong>For this walkthrough:</strong></p><p>{dllink}</p></pre>")

```

Now load the tidyverse.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
```
```{r load-tidyverse-solution}
library(tidyverse)
```
```{r load-tidyverse-check}
grade_this_code()
```

And the data.

```{r ggplot-load-data, message=FALSE, warning=FALSE}
softball <- read_csv("https://mattwaite.github.io/sportsdatafiles/ncaa_softball_2023.csv")

huskers <- softball |>
  filter(
    Institution == "Nebraska"
  )

big10 <- softball |> 
  filter(
    Conference == "Big Ten"
  )
```
```{r ggplot-load-data-exercise, exercise = TRUE}
softball <- read_csv("https://mattwaite.github.io/sportsdatafiles/ncaa_softball_2023.csv")
```
```{r ggplot-load-data-exercise-solution}
softball <- read_csv("https://mattwaite.github.io/sportsdatafiles/ncaa_softball_2023.csv")
```
```{r ggplot-load-data-exercise-check}
grade_this_code()
```

And let's get a glimpse of the data to remind us what is all there. 

Let's `glimpse` the data because it's wide and has a lot of columns. 

```{r glimpse-data, exercise=TRUE, exercise.setup = "ggplot-load-data"}
glimpse(softball)
```
```{r glimpse-data-solution}
glimpse(softball)
```
```{r glimpse-data-check}
grade_this_code()
```

Now that we're set up, when making graphics, you're rarely ever going straight from data to graphic. Normally, you'll have way too much data to visualize, or you need to do something to that data in order to visualize it. In our case, we have way too many schools in our data to make a reasonable bar chart. Our first challenge, then, is to narrow the pile. 

### Exercise 1: Set up your data. 

To do this, we're going to do a few things we've done before. We're going to filter all the softball teams in Division I down to just the Big Ten. We'll put that into a new dataframe called `big10`.

```{r ggplot-top, exercise=TRUE, exercise.setup = "ggplot-load-data", message=FALSE}
____ <- ____ |> 
  filter(
    Conference == "____"
  )
```
```{r ggplot-top-solution, exercise.reveal_solution = FALSE}
big10 <- softball |> 
  filter(
    Conference == "Big Ten"
  )
```
```{r ggplot-top-check}
grade_this_code()
```

Now, we should have 14 teams that we can look at. That's far more manageable than 306 -- the number of softball teams in Division I. 

## The bar chart

The easiest thing we can do is create a simple bar chart of our data. **Bar charts show magnitude. They invite you to compare how much more or less one thing is compared to others.**

We could, for instance, create a bar chart of home runs each team hit. We might be asking the question: Which Big Ten team relies on the long ball to score runs? To do that, we simply tell `ggplot2` what our dataset is, what element of the data we want to make the bar chart out of (which is the aesthetic), and the geometry type (which is the geom). It looks like this:

`ggplot() + geom_bar(data=big10, aes(x=Institution))` 

We start with `ggplot()` which is creating a blank canvas for us to work in. The `geom_bar()` is the geometry -- the form the data will take. We will learn many geometries over the next several lessons. `huskers` is our data, `aes` means aesthetics, `x=Season` explicitly tells `ggplot2` that our x value -- our horizontal value -- is the Season field from the data. Why season? We want one bar per season, no? Put the season on the x axis is what that is saying. And what do we get when we run that? 

```{r ggplot-bar1, exercise=TRUE, exercise.setup = "ggplot-load-data", message=FALSE}
ggplot() + geom_bar(data=big10, aes(x=Institution))
```
```{r ggplot-bar1-solution, exercise.reveal_solution = FALSE}
ggplot() + geom_bar(data=big10, aes(x=Institution))
```
```{r ggplot-bar1-check}
grade_this_code()
```

We get ... weirdness. We expected to see bars of different sizes, but we get all with a count of 1. What gives? Well, this is the default behavior. What we have here is something called a histogram, where `ggplot2` helpfully counted up the number of times the Institution appears and made bars as long as the count. Since we only have one record per Institution, the count is always 1. How do we fix this? By adding `weight` to our aesthetic. 

### Exercise 2: The weight of responsibility

You saw how it was done before. This just adds weight to the aes. The weight is always going to be a number, and our number is the column for home runs.

```{r ggplot-bar2, exercise=TRUE, exercise.setup = "ggplot-load-data", message=FALSE}
ggplot() + 
  geom_bar(data=big10, aes(x=Institution, weight=____))
```
```{r ggplot-bar2-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bar(data=big10, aes(x=Institution, weight=HR))
```
```{r ggplot-bar2-check}
grade_this_code()
```

Now we get bars of different sizes, and you can clearly see a difference between teams. 

Let's work on a different version of this to see if we can find a clearer story.

### Exercise 3: Reordering

`ggplot2`'s default behavior is to sort the data by the x axis variable. It's in alphabetical order. Rarely ever is that useful. 

To change the order of the bars, we have to `reorder` it. With `reorder`, we first have to tell `ggplot` what we are reordering, and then we have to tell it HOW we are reordering it. Nearly 99 times out of 100, that HOW we're reordering it is going to be the number. So it's reorder(FIELD, SORTFIELD). So let's put the bars in order of how many home runs they hit.

```{r ggplot-bar3, exercise=TRUE, exercise.setup = "ggplot-load-data", message=FALSE}
ggplot() + 
  geom_bar(data=big10, aes(x=reorder(Institution, ____), weight=____))
```
```{r ggplot-bar3-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bar(data=big10, aes(x=reorder(Institution, HR), weight=HR))
```
```{r ggplot-bar3-check}
grade_this_code()
```

There, now our bars are in some kind of order.

### Exercise 4: Layering

Our data has 14 teams in it but what if we wanted to focus on one team? We can use layering for that. 
First, we need to make a dataframe of one. Let's focus on Nebraska for ... reasons. The, we just add another geom_bar to our chart.

To our new geom, we're going to add one new thing -- a fill. Let's make the color red. Again. For reasons. 

```{r ggplot-bar4, exercise=TRUE, exercise.setup = "ggplot-load-data", message=FALSE}
huskers <- softball |>
  filter(
    Institution == "Nebraska"
  )

ggplot() + 
  geom_bar(data=big10, aes(x=reorder(____, ____), weight=____)) + 
  geom_bar(data=huskers, aes(x=reorder(____, ____), weight=____), fill="____")   
```
```{r ggplot-bar4-solution, exercise.reveal_solution = FALSE}
huskers <- softball |>
  filter(
    Institution == "Nebraska"
  )

ggplot() + 
  geom_bar(data=big10, aes(x=reorder(Institution, HR), weight=HR)) +
  geom_bar(data=huskers, aes(x=reorder(Institution, HR), weight=HR), fill="red")
```
```{r ggplot-bar4-check}
grade_this_code()
```

Where Nebraska ranks is pretty clear now, eh?

## One last trick: coord flip

Sometimes, we don't want vertical bars. Maybe we think this would look better horizontal. Maybe our x-axis labels will never fit on the x-axis. How do we do that? By adding `coord_flip()` to our code. It does what it says -- it inverts the coordinates of the figures.

### Exercise 5: Flip the coordinates

```{r ggplot-bar5, exercise=TRUE, exercise.setup = "ggplot-load-data", message=FALSE}
ggplot() + 
  geom_bar(data=big10, aes(x=reorder(Institution, HR), weight=HR)) +
  geom_bar(data=huskers, aes(x=reorder(Institution, HR), weight=HR), fill="red") +
  ____()
```
```{r ggplot-bar5-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bar(data=big10, aes(x=reorder(Institution, HR), weight=HR)) +
  geom_bar(data=huskers, aes(x=reorder(Institution, HR), weight=HR), fill="red") + 
  coord_flip()
```
```{r ggplot-bar5-check}
grade_this_code()
```

Now we've got something working. There's a lot of work left to do to make this publishable, but we'll get to that in due time.

## The Recap

In this lesson, we've taken our first steps into the world of data visualization with ggplot2, focusing on bar charts. We've learned how to create basic bar charts using geom_bar(), and how to customize them to better represent our data. We explored key concepts such as filtering data to focus on specific subsets (like the Big Ten conference), using the weight aesthetic to represent values other than count, and reordering bars to improve readability. We also discovered how to use layering to highlight specific data points (like Nebraska's performance) and how to flip coordinates for horizontal bar charts. These foundational skills in creating and customizing bar charts will serve as building blocks for more complex visualizations in future lessons. Remember, effective data visualization is an iterative process - don't be afraid to experiment with different approaches to find the most clear and compelling way to present your sports data insights.

## Terms to Know

- ggplot2: A data visualization package for R that implements the grammar of graphics, allowing for the creation of complex plots from data in a layered fashion.
- `geom_bar()`: A ggplot2 function used to create bar charts, which represent data with rectangular bars proportional to the values they represent.
- Aesthetics (aes): In ggplot2, aesthetics define how variables in the data are mapped to visual properties of the plot, such as x and y positions, colors, sizes, etc.
weight: An aesthetic in geom_bar() that allows you to represent values other than simple counts in your bar chart.
- `reorder()`: A function used to change the order of categorical variables in a plot, often used to sort bars by their values for better readability.
- `coord_flip()`: A ggplot2 function that switches the x and y axes, useful for creating horizontal bar charts.
- Layering: The process of adding multiple geometric objects (geoms) to a plot, allowing for more complex and informative visualizations.
- `fill`: An aesthetic that controls the interior color of geometric shapes in a plot, such as the color inside bars in a bar chart.
- `data`: The argument in ggplot2 functions that specifies which dataset to use for creating the visualization.
