---
title: "Sports Data Lesson 27: Shapes as storytelling"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to draw attention to some points.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(ggalt)
library(ggrepel)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

The goal of this lesson is to introduce you to the technique of encircling points on a scatterplot using the ggalt package in R and annotation with rectangles and `geom_rect`. By the end of this tutorial, you'll understand how to use the `geom_encircle()` and `annotate` functions to draw attention to specific data points or groups of points in your visualizations. You'll learn how to create basic scatterplots, layer additional data points, and then highlight key information using circles and text labels. We'll apply these concepts to real college basketball data, analyzing player performance in terms of shooting percentage and rebounds. This lesson will enhance your ability to create more informative and visually compelling scatterplots, allowing you to guide your audience's attention to the most important aspects of your data story in sports analytics.

## The Basics

One thing we've talked about all semester is drawing attention to the thing you want to draw attention to. We've used color and labels to do that so far. Let's add another layer to it -- a shape around the point or points you want to highlight. 

Remember: The point of all of this is to draw the eye to what you are trying to show your reader. You want people to see the story you are trying to tell. 

It's not hard to draw a shape in ggplot -- it is a challenge to put it in the right place. But, there is a library to the rescue that makes this super easy -- `ggalt`. 

Like usual, you already installed this if you followed the install instructions at the beginning of the course. But if you're having trouble getting this to work, go to your console in R Studio and install it with `install.packages("ggalt")`

There's a bunch of things that `ggalt` does, but one of the most useful for us is the function `encircle`. Let's dive in.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
library(ggalt)
library(ggrepel)
```
```{r load-tidyverse-solution}
library(tidyverse)
library(ggalt)
library(ggrepel)
```
```{r load-tidyverse-check}
grade_this_code()
```

Let's say we want to highlight Brice Williams and his place in college basketball last season. Let's take a look at last season's player data. 

And while we're loading it, let's filter out anyone who hasn't played 100 minutes. 

```{r encircle-load-data, message=FALSE, warning=FALSE}
players <- read_csv("https://mattwaite.github.io/sportsdatafiles/players25.csv") |> filter(MP > 100)

nu <- players |> filter(Team == "Nebraska Cornhuskers Men's")

bw <- nu |> filter(Player == "Brice Williams")
```
```{r encircle-load-data-exercise, exercise = TRUE}
players <- read_csv("https://mattwaite.github.io/sportsdatafiles/players25.csv") |>
  filter(MP > 100)
```
```{r encircle-load-data-exercise-solution}
players <- read_csv("https://mattwaite.github.io/sportsdatafiles/players25.csv") |>
  filter(MP > 100)
```
```{r encircle-load-data-exercise-check}
grade_this_code()
```

Let's make a standard scatterplot of points and player efficiency rating. 

### Exercise 1: Making the first scatterplot

Who is hitting shots, and who is using his chances to score efficiently is the question we're after here. Our data is players, and our two fields are PTS and PER.

```{r encircle1, exercise=TRUE, exercise.setup = "encircle-load-data", message=FALSE}
ggplot() + geom_point(data=____, aes(x=____, y=____))
```
```{r encircle1-solution, exercise.reveal_solution = FALSE}
ggplot() + geom_point(data=players, aes(x=PTS, y=PER))
```
```{r encircle1-check}
grade_this_code()
```

We can see right away that there are some dots at the edges -- lots of points but maybe they aren't contributing elsewhere. Where are the Huskers in this? We can get that with filtering and layering.

### Exercise 2: Adding the Huskers as a layer

Like we have done in the past, let's make a dataframe of Nebraska players and add a layer to our graph that puts Nebraska's dots in red. While we're at it, we'll make everyone else grey. And let's label the Nebraska players.

```{r encircle2, exercise=TRUE, exercise.setup = "encircle-load-data", message=FALSE}
nu <- players |> filter(Team == "Nebraska Cornhuskers Men's")

ggplot() + 
  geom_point(data=____, aes(x=____, y=____), color="grey") +
  geom_point(data=____, aes(x=____, y=____), color="red") +
  geom_text_repel(data=____, aes(x=PTS, y=PER, label=Player))
```
```{r encircle2-solution, exercise.reveal_solution = FALSE}
nu <- players |> filter(Team == "Nebraska Cornhuskers Men's")

ggplot() + 
  geom_point(data=players, aes(x=PTS, y=PER), color="grey") + 
  geom_point(data=nu, aes(x=PTS, y=PER), color="red") +
  geom_text_repel(data=nu, aes(x=PTS, y=PER, label=Player))
```
```{r encircle2-check}
grade_this_code()
```

Given that this chart is about Brice Williams, I'll give you three guesses who that dot is scoring a lot of points really efficiently. Your first two guesses don't count and his name rhymes with rice illiams.

But just this quickly, we're on the path to something publishable. We'll need to label a dot and we'll need to drop the default grey and add some headlines and all that. And, for the most part, we've got a solid chart. 

But what if we could really draw the eye to a dot or group of them? In `ggalt`, there is a geom called `geom_encircle`, which ... does what you think it does. It encircles all the dots in a dataset. 

### Exercise 3: geom_encircle

Now we'll add geom_encircle and we'll just copy the data and the aes from our nu geom_point. Then, we need to give the encirclement a shape using s_shape -- which is a number between 0 and 1 -- and then how far away from the dots to draw the circle using expand, which is another number between 0 and 1.

Let's start with `s_shape` .03 and `expand` .03. Getting to the right number involves a lot of trial and error messing with that number. Sometimes you'll want a bigger circle -- closer to one -- and sometimes you want small. 

```{r encircle3, exercise=TRUE, exercise.setup = "encircle-load-data", message=FALSE}
bw <- nu |> filter(Player == "Brice Williams")

ggplot() + 
  geom_point(data=____, aes(x=____, y=____), color="grey") +
  geom_point(data=____, aes(x=____, y=____), color="red") +
  geom_text_repel(data=nu, aes(x=PTS, y=PER, label=Player)) + 
  geom_encircle(data=bw, aes(x=____, y=____), s_shape=.03, expand=.03, colour="red")
```
```{r encircle3-solution, exercise.reveal_solution = FALSE}
bw <- nu |> filter(Player == "Brice Williams")

ggplot() + 
  geom_point(data=players, aes(x=PTS, y=PER), color="grey") + 
  geom_point(data=nu, aes(x=PTS, y=PER), color="red") +
  geom_text_repel(data=nu, aes(x=PTS, y=PER, label=Player)) +
  geom_encircle(data=bw, aes(x=PTS, y=PER), s_shape=.03, expand=.03, colour="red")
```
```{r encircle3-check}
grade_this_code()
```

That's not bad. If you disagree, now you begin the process of fiddling with `s_shape` and `expand` until you get it just right. 

## Using annotate to draw rectangles

What if, instead of a circle, you wanted to add a rectangle to highlight the zone of good players (or something like that). Rectangles work great to highlight zones or, if you have time as your x axis, events and time periods. Think of it this way -- if you had a line chart of Nebraska's wins over time, you could add rectangles over the years for certain coaches you wanted to highlight. 

### Exercise 4: Using a rectangle instead of a circle

So how do you make rectangles? With annotate and some interpolation. What's interpolation? It's looking at the x and y axis and interpreting where to draw the lines.

Looking at our previous work, what if we use 200 points and a 20 PER rating as the floor for our good zone. The end ceiling is the top of the chart -- which ggplot can conveniently just use infinity for that. 

One more thing? We should label our zone. It looks like there's a spot at 300 points and 35 PER to fit some text. 

```{r encircle4, exercise=TRUE, exercise.setup = "encircle-load-data", message=FALSE}
ggplot() + 
  geom_point(data=players, aes(x=PTS, y=PER), color="grey") +
  geom_point(data=nu, aes(x=PTS, y=PER), color="red") +
  geom_text_repel(data=nu, aes(x=PTS, y=PER, label=Player)) +
  annotate("rect", fill = "blue", alpha = 0.1, 
        xmin = ____, xmax = Inf,
        ymin = ____, ymax = Inf) +
  geom_text(aes(x=____, y=____, label="Good player zone"), size=3)
```
```{r encircle4-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(data=players, aes(x=PTS, y=PER), color="grey") +
  geom_point(data=nu, aes(x=PTS, y=PER), color="red") + 
  geom_text_repel(data=nu, aes(x=PTS, y=PER, label=Player)) +
  annotate("rect", fill = "blue", alpha = 0.1, 
        xmin = 200, xmax = Inf,
        ymin = 20, ymax = Inf) +
  geom_text(aes(x=300, y=35, label="Good player zone"), size=3)
```
```{r encircle4-check}
grade_this_code()
```

Now let's clean this up and make it presentable with labels, a theme and some alterations to that theme.

```{r encircle5, exercise=TRUE, exercise.setup = "encircle-load-data", message=FALSE}
ggplot() + 
  geom_point(data=players, aes(x=PTS, y=PER), color="grey") +
  geom_point(data=nu, aes(x=PTS, y=PER), color="red") +
  geom_text_repel(data=nu, aes(x=PTS, y=PER, label=Player)) +
  annotate("rect", fill = "blue", alpha = 0.1, 
        xmin = ____, xmax = Inf,
        ymin = ____, ymax = Inf) +
  geom_text(aes(x=300, y=35, label="Good player zone"), size=3) +
  labs(title="Brice Williams is Nebraska's MVP", subtitle="Not just Nebraska's best, but one of the best in the Big Ten.", x="Points scored", y="Player Efficiency Rating") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )
```
```{r encircle5-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(data=players, aes(x=PTS, y=PER), color="grey") +
  geom_point(data=nu, aes(x=PTS, y=PER), color="red") + 
  geom_text_repel(data=nu, aes(x=PTS, y=PER, label=Player)) +
  annotate("rect", fill = "blue", alpha = 0.1, 
        xmin = 200, xmax = Inf,
        ymin = 20, ymax = Inf) +
  geom_text(aes(x=300, y=35, label="Good player zone"), size=3) + 
  labs(title="Brice Williams is Nebraska's MVP", subtitle="Not just Nebraska's best, but one of the best in the Big Ten.", x="Points scored", y="Player Efficiency Rating") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )
```
```{r encircle5-check}
grade_this_code()
```

## The Recap

In this lesson, we've explored the powerful technique of encircling points on a scatterplot using the ggalt package in R and making rectangles with annotate. We started by creating a basic scatterplot of college basketball player performance, focusing on points and efficiency. We then learned how to layer additional data points to highlight specific teams or players. The key skill we developed was using geom_encircle() and annotate to draw attention to particular data points, in this case, showcasing Brice Williams exceptional performance for Nebraska. Throughout the lesson, we emphasized the importance of visual hierarchy and guiding the viewer's eye to the most crucial information. We also touched on how to make our visualizations more polished and publication-ready by adding appropriate titles, labels, and theming. Remember, the goal of these techniques is not just to present data, but to tell a compelling story with your visualizations.

## Terms to Know

- `ggalt`: An extension package for ggplot2 that provides additional geoms and statistical transformations.
- `geom_encircle()`: A function from ggalt that allows you to draw a circle or shape around specified data points on a plot.
- `s_shape`: A parameter in geom_encircle() that controls the shape of the encircling line, with values between 0 and 1.
- `expand`: A parameter in geom_encircle() that determines how far the encircling line extends from the data points.