---
title: "Sports Data Lesson 17: Step charts"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to show change over time, but now with magnitude.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

The goal of this lesson is to introduce you to step charts, a powerful visualization tool for displaying cumulative change over time. By the end of this tutorial, you'll understand when to use step charts, how to create them using ggplot2, and how to customize them for clearer data representation. You'll learn to prepare cumulative data, plot multiple step lines, use color effectively to highlight specific data series, and add annotations to enhance your chart's narrative. We'll apply these concepts to real college basketball data, analyzing team performance across a season using cumulative point differentials. This lesson will enhance your ability to visualize and interpret cumulative data trends, a crucial skill in sports analytics for understanding team performance trajectories and making meaningful comparisons between teams over time.

## The Basics

Step charts are **a method of showing progress** toward a total number They combine showing change over time -- **cumulative change over time** -- with magnitude. They're good at inviting comparison. 

There's great examples out there. First is the Washington Post looking at [Lebron passing Jordan's career point total](https://www.washingtonpost.com/graphics/sports/lebron-james-michael-jordan-nba-scoring-list/?utm_term=.481074150849). Another is John Burn-Murdoch's work at the Financial Times (which is paywalled) about soccer stars. [Here's an example of his work outside the paywall](https://johnburnmurdoch.github.io/projects/goal-lines/CL/).

To replicate this, we need cumulative data -- data that is the running total of data at a given point. So think of it this way -- Nebraska scores 50 points in a basketball game and then 50 more the next, their cumulative total at two games is 100 points. 

Step charts can be used for all kinds of things -- showing how a player's career has evolved over time, how a team fares over a season, or franchise history. Let's walk through an example. 

Let's look at Fred Hoiberg's most recent season. 

We'll need the tidyverse.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
```
```{r load-tidyverse-solution}
library(tidyverse)
```
```{r load-tidyverse-check}
grade_this_code()
```

And the data.

```{r step-load-data, message=FALSE, warning=FALSE}
logs <- read_csv("https://mattwaite.github.io/sportsdatafiles/logs26.csv")

difflogs <- logs |> 
  mutate(differential = team_score - opponent_score) |> 
  group_by(team) |> 
  mutate(cum_diff = cumsum(differential))

bigdiff <- difflogs |> filter(conference == "Big Ten")
nu <- bigdiff |> filter(team == "Nebraska Cornhuskers")
pu <- bigdiff |> filter(team == "Purdue Boilermakers")
mi <- bigdiff |> filter(team == "Michigan Wolverines")
```
```{r step-load-data-exercise, exercise = TRUE}
logs <- read_csv("https://mattwaite.github.io/sportsdatafiles/logs26.csv")
```
```{r step-load-data-exercise-solution}
logs <- read_csv("https://mattwaite.github.io/sportsdatafiles/logs26.csv")
```
```{r step-load-data-exercise-check}
grade_this_code()
```

Here we're going to look at the scoring differential of teams. If you score more than your opponent, you win. So it stands to reason that if you score a lot more than your opponent over the course of a season, you should be very good, right? Let's see.

### Exercise 1: Setting up the data

The first step to making a step chart is that we need to calculate our rolling cumulative value. To do that, we'll create a new dataframe called difflogs -- for differential logs, since we have game log data. Then, we'll group it by the team. After that, we're going to *mutate* using a new function called `cumsum` or cumulative sum -- the sum for each game as we go forward. So game 1's cumsum is the differential of that game. Game 2's cumsum is Game 1 + Game 2. Game 3 is Game 1 + 2 + 3 and so on. When you group by and *don't* summarize, it's like putting all your data into little piles -- your groups -- and doing something to each thing in the pile. Then, when you're done, you put the things back. Where group by and summarize creates one row per group, group by *without* summarize leaves you with the same number you started with. It's a subtle thing, but important. 

```{r step1, exercise=TRUE, exercise.setup = "step-load-data", message=FALSE}
____ <- logs |> 
  mutate(differential = team_score - opponent_score) |> 
  group_by(____) |> 
  mutate(cum_diff = ____(differential))
```
```{r step1-solution, exercise.reveal_solution = FALSE}
difflogs <- logs |> 
  mutate(differential = team_score - opponent_score) |> 
  group_by(team) |> 
  mutate(cum_diff = cumsum(differential))
```
```{r step1-check}
grade_this_code()
```

And now we have the same data, but each team has a rolling cumulative value for how many points they scored versus there opponent.

### Exercise 2: Filtering

Now that we have the cumulative sum for each, let's filter it down to just Big Ten teams. Let's create a new dataframe called bigdiff -- for Big Ten differential. 

```{r step2, exercise=TRUE, exercise.setup = "step-load-data", message=FALSE}
____ <- difflogs |> filter(conference == "____")
```
```{r step2-solution, exercise.reveal_solution = FALSE}
bigdiff <- difflogs |> filter(conference == "Big Ten")
```
```{r step2-check}
grade_this_code()
```

Now we've got what we need to make a graphic.

### Exercise 3: The step chart

The step chart is it's own geom -- `geom_step` -- so we can employ it just like we have the others. It works almost exactly the same as a line chart, but it uses the cumulative number instead of a regular value and, as the name implies, creates a step like shape to the line instead of a curve.

```{r step3, exercise=TRUE, exercise.setup = "step-load-data", message=FALSE}
ggplot() + ____(data=____, aes(x=game_date, y=____, group=team))
```
```{r step3-solution, exercise.reveal_solution = FALSE}
ggplot() + geom_step(data=bigdiff, aes(x=game_date, y=cum_diff, group=team))
```
```{r step3-check}
grade_this_code()
```

### Exercise 4: Using color ... badly.

Let's try a different element of the aesthetic: color, but this time *inside the aesthetic*. Last time, we did the color outside. When you put it inside, you pass it a column name and ggplot will color each line based on what thing that is, and it will create a legend that labels each line that thing. 

```{r step4, exercise=TRUE, exercise.setup = "step-load-data", message=FALSE}
ggplot() + ____(data=____, aes(x=game_date, y=____, group=team, color=team))
```
```{r step4-solution, exercise.reveal_solution = FALSE}
ggplot() + geom_step(data=bigdiff, aes(x=game_date, y=cum_diff, group=team, color=team))
```
```{r step4-check}
grade_this_code()
```

From this, we can see a handful of teams in the Big Ten aren't doing so great. But which is which? And which one is Nebraska? Too many colors and it's too hard to tell. This is a lesson: *We use color to highlight.* When everything is a color, nothing is highlighted. We use color sparingly.

### Exercise 5: Using color the right way

How do we sort this out? Layering and annotations. We've used layering before. Annotations, while new, will make sense in a minute. First, let's pick three teams to highlight -- Nebraska, Purdue and Indiana. 

```{r step5, exercise=TRUE, exercise.setup = "step-load-data", message=FALSE}
nu <- bigdiff |> filter(team == "Nebraska Cornhuskers")
pu <- bigdiff |> filter(team == "____ ____")
mi <- bigdiff |> filter(team == "____ ____")
```
```{r step5-solution, exercise.reveal_solution = FALSE}
nu <- bigdiff |> filter(team == "Nebraska Cornhuskers")
pu <- bigdiff |> filter(team == "Purdue Boilermakers")
mi <- bigdiff |> filter(team == "Michigan Wolverines")
```
```{r step5-check}
grade_this_code()
```

Let's introduce a couple of new things here. First, note when I take the color OUT of the aesthetic, the legend disappears. 

Our chart is going to layer bigdiff, nu, pu and id in that order. We put bigdiff first because order matters and the first geom goes on the bottom.

The second thing I'm going to add is the annotation layer. In this case, I am adding a text annotation layer, and I can specify where by adding in a x and a y value where I want to put it. This takes some finesse -- trial and error. You use the x and y axis values, guesstimate where you want the center of the words to be and enter those numbers into the x and y values. 

After that, I'm going to add labels and a theme. These too are new. Eventually, every chart we make will include labels and theming. Because step charts are so similar to line charts, we can add new stuff here and your brain won't break (I promise). To get credit for this, you'll need to fill in the `labs` part where there's all caps. If you try to check your code against the solution, you'll probably get an error because mine will be different from yours. Don't worry about it. If you get a chart and it looks like a step chart, you're good.

```{r step6, exercise=TRUE, exercise.setup = "step-load-data", message=FALSE}
ggplot() + 
  geom_step(data=bigdiff, aes(x=game_date, y=____, group=team), color="lightgrey") +
  geom_step(data=____, aes(x=game_date, y=cum_diff, group=team), color="red") + 
  geom_step(data=____, aes(x=game_date, y=cum_diff, group=team), color="black") +
  geom_step(data=____, aes(x=game_date, y=cum_diff, group=team), color="darkblue") +
  annotate("text", x=(as.Date("2025-12-15")), y=150, label="Nebraska") +
  annotate("text", x=(as.Date("2025-12-15")), y=210, label="Purdue") +
  annotate("text", x=(as.Date("2025-12-15")), y=290, label="Michigan") +
  labs(
    x="YOU SHOULD CHANGE THIS TO WHAT THE X AXIS IS", 
    y="Cumulative Point Differential", 
    title="GIVE THIS A HEADLINE: WHAT DOES THE GRAPH SAY?", 
    subtitle="THE SUBTITLE IS THE TEXT UNDER THE HEADLINE: CHANGE THIS TO HELP EXPLAIN YOUR GRAPH", 
    caption="Source: Sports-Reference.com | By YOUR NAME"
    ) + 
  theme_minimal()
```
```{r step6-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_step(data=bigdiff, aes(x=game_date, y=cum_diff, group=team), color="lightgrey") +
  geom_step(data=nu, aes(x=game_date, y=cum_diff, group=team), color="red") + 
  geom_step(data=pu, aes(x=game_date, y=cum_diff, group=team), color="black") +
  geom_step(data=mi, aes(x=game_date, y=cum_diff, group=team), color="darkblue") +
  annotate("text", x=(as.Date("2025-12-15")), y=150, label="Nebraska") +
  annotate("text", x=(as.Date("2025-12-15")), y=210, label="Purdue") +
  annotate("text", x=(as.Date("2025-12-15")), y=290, label="Michigan") +
  labs(
    x="STOP USING SUBMIT ANSWER AND LABEL THIS YOURSELF", 
    y="STOP USING SUBMIT ANSWER AND LABEL THIS YOURSELF", 
    title="STOP USING SUMBIT ANSWER AND MAKE YOUR OWN HEADLINE", 
    subtitle="STOP USING SUBMIT ANSWER AND MAKE YOUR OWN SUBTITLE", 
    caption="Source: Sports-Reference.com | By YOUR NAME"
    ) + 
  theme_minimal()
```
```{r step6-check}
grade_this_code()
```

This is a decent looking chart.

## The Recap

In this lesson, we've explored the power of step charts for visualizing cumulative change over time in sports data. We started by learning how to prepare cumulative data using group_by() and cumsum() functions, a crucial step in creating meaningful step charts. We then progressed to creating basic step charts using geom_step(), and advanced to more complex visualizations comparing multiple teams' performances. We covered important techniques such as effective use of color to highlight specific teams, layering to provide context, and adding annotations to clarify data points. We also introduced labeling and theming to enhance the professional appearance of our charts. Throughout the lesson, we've seen how step charts can effectively show the progression of team performance over a season, allowing for easy comparisons between teams and revealing patterns that might be missed in other chart types. Remember, while step charts are excellent for showing cumulative trends, they're most effective when used thoughtfully - consider your color choices, annotations, and overall design to ensure your chart tells a clear and compelling story. 

## Terms to Know

- Step Chart: A type of line chart where the line moves in discrete steps, typically used to show cumulative change over time.
- `geom_step()`: A ggplot2 function used to create step charts, connecting data points with horizontal and vertical lines.
- Cumulative Data: Data that represents a running total, where each new data point is added to the sum of all previous data points.
- `cumsum()`: A function in R that calculates the cumulative sum of a vector, essential for creating step charts.
- Annotation: Additional text or graphical elements added to a chart to provide context, explanation, or highlight specific data points.
- `annotate()`: A ggplot2 function used to add annotations to a chart, such as text labels or shapes.
- `labs()`: A ggplot2 function used to add labels to a chart, including titles, subtitles, axis labels, and captions.
- `theme_minimal()`: A ggplot2 theme that provides a clean, minimalist appearance to charts.
- `Point Differential`: In sports, the difference between points scored by a team and points scored by their opponents, often used as a measure of team performance.