---
title: "Sports Data Lesson 29: Text cleaning"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to clear out some junk that data providers give you.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

The goal of this lesson is to introduce you to essential text cleaning techniques for sports data analysis. By the end of this tutorial, you'll understand how to handle common data inconsistencies and formatting issues that often arise when working with sports statistics. You'll learn how to use regular expressions, the gsub() function, and tidyr's separate() function to clean player names, split compound data fields, and convert text-based measurements into numeric values. These skills will enable you to transform messy, raw data into clean, analysis-ready datasets. By mastering these techniques, you'll be able to produce more accurate analyses and create more professional-looking visualizations, enhancing the quality and credibility of your sports analytics blog posts.

## The Basics

On occasion, you'll get some data from someone that ... isn't quite what you need it to be. There's something flawed in it. Some extra text, some choice that the data provider made that you just don't agree with. 

There's a ton of tools in the tidyverse to fix this, and you already have some tools in your toolboxt. Let's take a look at a couple. 

First, you know what you need.

```{r load-tidyverse, exercise=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
```
```{r load-tidyverse-solution}
library(tidyverse)
```
```{r load-tidyverse-check}
grade_this_code()
```

Second: The trick to this is thinking logically. If I could do X, I could get this where I want it to be. How do I do X? If you can get to a point where you ask that question, text cleaning gets a lot easier.

Now, two examples.

## Stripping out text

Throughout this class, we've used data from Sports Reference. If you've used their Share > CSV method to copy data from a table, you may have noticed some extra cruft in the player name field. If you haven't seen it, I'll give you an example -- a dataset of NBA players and their advanced metrics. Per usual: **For purposes of this exercise, you don't need to do this. The data is included here if you want to try this in your own notebook.** 

```{r echo=FALSE, class.output="bg-info", results="asis",  message=FALSE,  warning=FALSE}
library(downloadthis)
library(glue)

dllink <- download_link(
  link = "https://mattwaite.github.io/sportsdatafiles/nbaplayers.csv",
  button_label = "Download csv file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

glue("<pre><p><strong>For this walkthrough:</strong></p><p>{dllink}</p></pre>")
```

Now load it.

```{r cleaning-load-data, message=FALSE, warning=FALSE}
nbaplayers <- read_csv("https://mattwaite.github.io/sportsdatafiles/nbaplayers.csv")
```
```{r cleaning-load-data-exercise, exercise = TRUE}
nbaplayers <- read_csv("https://mattwaite.github.io/sportsdatafiles/nbaplayers.csv")
```
```{r cleaning-load-data-exercise-solution}
nbaplayers <- read_csv("https://mattwaite.github.io/sportsdatafiles/nbaplayers.csv")
```
```{r cleaning-load-data-exercise-check}
grade_this_code()
```

Let's take a look:

```{r head-data, exercise=TRUE, exercise.setup = "cleaning-load-data"}
head(nbaplayers)
```
```{r head-data-solution}
head(nbaplayers)
```
```{r head-data-check}
grade_this_code()
```

You can see that every player's name is their name, then two backslashes, then some version of their name that must have meaning to Sports Reference, but not to us. We need to get rid of that, or our graphics will look like garbage. 

To do this, we're going to use a little regular expression magic. Regular expressions are a programmatic way to find any pattern in text. What we're looking for is that `\\` business. But, that presents a problem, because the `\` is a special character. It's called an escape character. That escape character means what comes next is potentially special. For instance, if you see `\n`, that's a newline character. So normally, if you see that, it would add a return. 

So for us to get rid of the `\` we're going to have to escape the escape character with an escape character. And we have two of them. So we have to do it twice. 

Yes. Really.

So if we wanted to find two backslashes, we need `\\\\`. Then, using regular expressions, we can say "and then everything else after this" with this: .* 

No really. That's it. So we're looking for `\\\\.*`. That'll find two backslashes and then everything after it. If you think this is hard ... you're right. Regular expressions are an entire month of a programming course by themselves. They are EXTREMELY powerful. 

### Exercise 1: Using regular expressions to clean out junk

To find something in text, we'll use a function called `gsub`. The pattern in `gsub` is `pattern, what we want to replace it with, what column this can all be found in`. So in our example, the pattern is `\\\\.*`, what we want to replace it with is ... nothing, and this is all in the Player column. 

```{r cleaning1, exercise=TRUE, exercise.setup = "cleaning-load-data", message=FALSE}
nbaplayers |> mutate(Player=gsub("____","",____))
```
```{r cleaning1-solution, exercise.reveal_solution = FALSE}
nbaplayers |> mutate(Player=gsub("\\\\.*","",Player))
```
```{r cleaning1-check}
grade_this_code()
```

Just like that, the trash is gone.

## Another example: splitting columns

Text cleaning is really just a set of logic puzzles. What do I need to do? How can I get there step by step?

The NCAA does some very interesting things with data, making it pretty useless. Here's some volleyball data from a few seasons ago.

```{r echo=FALSE, class.output="bg-info", results="asis",  message=FALSE,  warning=FALSE}
library(downloadthis)
library(glue)

dllink <- download_link(
  link = "https://mattwaite.github.io/sportsdatafiles/killsperset.csv",
  button_label = "Download csv file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

glue("<pre><p><strong>For this walkthrough:</strong></p><p>{dllink}</p></pre>")
```

Let's import it and take a look.

```{r cleaning2-load-data, message=FALSE, warning=FALSE}
kills <- read_csv("https://mattwaite.github.io/sportsdatafiles/killsperset.csv")
```
```{r cleaning2-load-data-exercise, exercise = TRUE}
kills <- read_csv("https://mattwaite.github.io/sportsdatafiles/killsperset.csv")
```
```{r cleaning2-load-data-exercise-solution}
kills <- read_csv("https://mattwaite.github.io/sportsdatafiles/killsperset.csv")
```
```{r cleaning2-load-data-exercise-check}
grade_this_code()
```

And we'll take a look.

```{r head2-data, exercise=TRUE, exercise.setup = "cleaning2-load-data"}
head(kills)
```
```{r head2-data-solution}
head(kills)
```
```{r head2-data-check}
grade_this_code()
```

First things first, Player isn't just player, it's player, school and conference, all in one. And Ht is a character field -- and in feet and inches.

So ... this is a mess. But there is a pattern. See it? A comma after the player's name. The Conference is in parens. We can use that. 

For this, we're going to use a `tidyr` function called `separate` to split columns into multiple columns based on a character. We'll do this step by step.

### Exercise 2: Separating on a character

Let's use that comma to split the player and the rest. With separate, the pattern is separate WHAT, into WHAT, by WHAT. So we're going to separate Player into Player and School, and we're going to separate it by the comma.

```{r cleaning2, exercise=TRUE, exercise.setup = "cleaning2-load-data", message=FALSE}
kills |> separate(Player, into=c("____", "____"), sep=",") 
```
```{r cleaning2-solution, exercise.reveal_solution = FALSE}
kills |> separate(Player, into=c("Player", "School"), sep=",") 
```
```{r cleaning2-check}
grade_this_code()
```

Good start.

Now, let's get the conference separated. 

### Exericse 3: Separating again, this time with a special character

A problem is going to crop up here -- the paren is a special character, so we have to escape it with the `\\`. 

```{r cleaning3, exercise=TRUE, exercise.setup = "cleaning2-load-data", message=FALSE}
kills |> 
  separate(Player, into=c("____", "____"), sep=",") |>
  separate(School, into=c("School", "Conference"), sep="____(") 
```
```{r cleaning3-solution, exercise.reveal_solution = FALSE}
kills |> 
  separate(Player, into=c("Player", "School"), sep=",") |>
  separate(School, into=c("School", "Conference"), sep="\\(")
```
```{r cleaning3-check}
grade_this_code()
```

Uh oh. Says we have problems in rows 15, 42 and 83. What are they? The NCAA has decided to put (FL), (NY) and (PA) into three teams to tell you they're in Florida, New York and Pennsylvania respectively. 

### Exercise 4: I hate the NCAA 

Well, we can fix that with some gsub and we'll use a switch called `fixed`, which when set to TRUE it means this literal string, no special characters.

```{r cleaning4, exercise=TRUE, exercise.setup = "cleaning2-load-data", message=FALSE}
kills |> 
  separate(Player, into=c("____", "____"), sep=",") |> 
  mutate(School = gsub("(FL)", "FL", School, fixed=TRUE)) |>
  mutate(School = gsub("(NY)", "NY", School, fixed=TRUE)) |>
  mutate(School = gsub("(PA)", "PA", School, fixed=TRUE)) |>
  separate(School, into=c("School", "Conference"), sep="____(") 
```
```{r cleaning4-solution, exercise.reveal_solution = FALSE}
kills |> 
  separate(Player, into=c("Player", "School"), sep=",") |> 
  mutate(School = gsub("(FL)", "FL", School, fixed=TRUE)) |>
  mutate(School = gsub("(NY)", "NY", School, fixed=TRUE)) |>
  mutate(School = gsub("(PA)", "PA", School, fixed=TRUE)) |>
  separate(School, into=c("School", "Conference"), sep="\\(")
```
```{r cleaning4-check}
grade_this_code()
```

One last thing: see the trailing paren? 

### Exercise 5: One last bit of junk

Believe it or not the trailing paren is NOT a special character. We can use mutate and gsub to replace ) with nothing. 

```{r cleaning5, exercise=TRUE, exercise.setup = "cleaning2-load-data", message=FALSE}
kills |> 
  separate(Player, into=c("____", "____"), sep=",") |> 
  mutate(School = gsub("(FL)", "FL", School, fixed=TRUE)) |>
  mutate(School = gsub("(NY)", "NY", School, fixed=TRUE)) |>
  mutate(School = gsub("(PA)", "PA", School, fixed=TRUE)) |>
  separate(School, into=c("School", "Conference"), sep="____(") |> 
  mutate(Conference=gsub("____", "", Conference))
```
```{r cleaning5-solution, exercise.reveal_solution = FALSE}
kills |> 
  separate(Player, into=c("Player", "School"), sep=",") |> 
  mutate(School = gsub("(FL)", "FL", School, fixed=TRUE)) |>
  mutate(School = gsub("(NY)", "NY", School, fixed=TRUE)) |>
  mutate(School = gsub("(PA)", "PA", School, fixed=TRUE)) |>
  separate(School, into=c("School", "Conference"), sep="\\(") |> 
  mutate(Conference=gsub(")", "", Conference))
```
```{r cleaning5-check}
grade_this_code()
```

Looking good, no errors.

## Working with numbers

Now, what should we do about Ht? 6-2 is not going to tell me much when I want to run a regression of height to kills per set. And it's a character field. So we need to convert it to numbers.

Separate again comes to the rescue.

```{r cleaning6, exercise=TRUE, exercise.setup = "cleaning2-load-data", message=FALSE}
kills |> 
  separate(Player, into=c("Player", "School"), sep=",") |> 
  mutate(School = gsub("(FL)", "FL", School, fixed=TRUE)) |>
  mutate(School = gsub("(NY)", "NY", School, fixed=TRUE)) |>
  mutate(School = gsub("(PA)", "PA", School, fixed=TRUE)) |>
  separate(School, into=c("School", "Conference"), sep="\\(") |> 
  mutate(Conference=gsub(")", "", Conference)) |>
  separate(Ht, into=c("Feet", "Inches"), sep="-") |>
  mutate(Feet = as.numeric(Feet), Inches = as.numeric(Inches))
```
```{r cleaning6-solution, exercise.reveal_solution = FALSE}
kills |> 
  separate(Player, into=c("Player", "School"), sep=",") |> 
  mutate(School = gsub("(FL)", "FL", School, fixed=TRUE)) |>
  mutate(School = gsub("(NY)", "NY", School, fixed=TRUE)) |>
  mutate(School = gsub("(PA)", "PA", School, fixed=TRUE)) |>
  separate(School, into=c("School", "Conference"), sep="\\(") |> 
  mutate(Conference=gsub(")", "", Conference)) |>
  separate(Ht, into=c("Feet", "Inches"), sep="-") |>
  mutate(Feet = as.numeric(Feet), Inches = as.numeric(Inches))
```
```{r cleaning6-check}
grade_this_code()
```

But how do we turn that into a height? Math! Check the last mutate.

```{r cleaning7, exercise=TRUE, exercise.setup = "cleaning2-load-data", message=FALSE}
kills |> 
  separate(Player, into=c("Player", "School"), sep=",") |> 
  mutate(School = gsub("(FL)", "FL", School, fixed=TRUE)) |>
  mutate(School = gsub("(NY)", "NY", School, fixed=TRUE)) |>
  mutate(School = gsub("(PA)", "PA", School, fixed=TRUE)) |>
  separate(School, into=c("School", "Conference"), sep="\\(") |> 
  mutate(Conference=gsub(")", "", Conference)) |>
  separate(Ht, into=c("Feet", "Inches"), sep="-") |>
  mutate(Feet = as.numeric(Feet), Inches = as.numeric(Inches)) |>
  mutate(Height = (Feet*12)+Inches)
```
```{r cleaning7-solution, exercise.reveal_solution = FALSE}
kills |> 
  separate(Player, into=c("Player", "School"), sep=",") |> 
  mutate(School = gsub("(FL)", "FL", School, fixed=TRUE)) |>
  mutate(School = gsub("(NY)", "NY", School, fixed=TRUE)) |>
  mutate(School = gsub("(PA)", "PA", School, fixed=TRUE)) |>
  separate(School, into=c("School", "Conference"), sep="\\(") |> 
  mutate(Conference=gsub(")", "", Conference)) |>
  separate(Ht, into=c("Feet", "Inches"), sep="-") |>
  mutate(Feet = as.numeric(Feet), Inches = as.numeric(Inches)) |>
  mutate(Height = (Feet*12)+Inches)
```
```{r cleaning7-check}
grade_this_code()
```

And now, in 10 lines of code, using separate, mutate and gsub, we've turned the mess that is the NCAA's data into actually useful data we can analyze. 

These patterns of thought come in handy when facing messed up data.

## The Recap

In this lesson, we've explored various techniques for cleaning and refining sports data to make it more suitable for analysis and visualization. We started by addressing common issues in data from sources like Sports Reference, learning how to use regular expressions with gsub() to remove unwanted text from player names. We then tackled more complex data cleaning challenges using NCAA volleyball statistics, where we applied the separate() function to split compound fields into distinct columns for player names, schools, and conferences. We also learned how to handle special characters and inconsistencies in data formatting, such as state abbreviations in parentheses. Finally, we converted text-based height measurements into numeric values, demonstrating how to transform data types for quantitative analysis. These skills are crucial for preparing clean, consistent datasets that will yield accurate insights and professional-looking visualizations in your sports analytics work. Remember, effective data cleaning is often the first and most important step in any data analysis project, setting the foundation for all subsequent analyses and visualizations.

## Terms to Know

- Regular Expression (Regex): A sequence of characters that defines a search pattern, used for pattern matching with strings.
- `gsub()`: A function in R used for finding and replacing patterns in strings.
- `separate()`: A function from the tidyr package used to split a single column into multiple columns.
- Escape Character: A character (usually a backslash ) used to indicate that the following character should be interpreted literally.
- `as.numeric()`: A function used to convert other data types to numeric values.
- `fixed = TRUE`: An argument in gsub() that treats the pattern as a fixed string rather than a regular expression.