---
title: "Sports Data Lesson 24: Cowplots"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to make multiple charts into one graphic.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(cowplot)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

# Arranging multiple plots together

## The basics

Sometimes you have two or three (or more) charts that by themselves aren't very exciting and are really just one chart that you need to merge together. It would be nice to be able to arrange them programmatically and not have to mess with it in Adobe Illustrator.

Good news.

There is.

It's called `cowplot`, and it's pretty easy to use. You installed cowplots way back in the beginning, but if you're having problems just go to the console in R Studio and run `install.packages("cowplot")`. 

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
library(cowplot)

```
```{r load-tidyverse-solution}
library(tidyverse)
library(cowplot)
```
```{r load-tidyverse-check}
grade_this_code()
```

We'll use the college football game data we've used before. Per usual: **For purposes of this exercise, you don't need to do this. The data is included here if you want to try this in your own notebook.** 

```{r echo=FALSE, class.output="bg-info", results="asis",  message=FALSE,  warning=FALSE}
library(downloadthis)
library(glue)

dllink <- download_link(
  link = "http://mattwaite.github.io/sportsdatafiles/footballlogs22.csv",
  button_label = "Download csv file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

glue("<pre><p><strong>For this walkthrough:</strong></p><p>{dllink}</p></pre>")
```

And load it.

```{r cowplot-load-data, message=FALSE, warning=FALSE}
games <- read_csv("http://mattwaite.github.io/sportsdatafiles/footballlogs22.csv")

totals <- games |> 
  filter(Conference == "Big Ten Conference") |> 
  group_by(Team) |> 
  summarize(
    TotalRushes = sum(RushingAtt),
    TotalRushYds = sum(RushingYds),
    TotalPasses = sum(PassingAtt),
    TotalPassYds = sum(PassingYds)
  ) |>
  mutate(
    RushingAvg = TotalRushYds/TotalRushes,
    PassingAvg = TotalPassYds/TotalPasses
  ) 

toppass <- totals |> top_n(5, wt=PassingAvg)
toprush <- totals |> top_n(5, wt=RushingAvg)

nu <- totals |> filter(Team == "Nebraska")

bar1 <- ggplot() + 
  geom_bar(data=toppass, aes(x=reorder(Team, PassingAvg), weight=PassingAvg)) +
  coord_flip()

bar2 <- ggplot() + 
  geom_bar(data=toprush, aes(x=reorder(Team, RushingAvg), weight=RushingAvg), color="grey") +
  coord_flip()
```
```{r cowplot-load-data-exercise, exercise = TRUE}
games <- read_csv("http://mattwaite.github.io/sportsdatafiles/footballlogs22.csv")
```
```{r cowplot-load-data-exercise-solution}
games <- read_csv("http://mattwaite.github.io/sportsdatafiles/footballlogs22.csv")
```
```{r cowplot-load-data-exercise-check}
grade_this_code()
```

For this, we're going to total up the rushing yards and attempts for each game, as well as the passing yards and attempts, to get a top 5 list of the best rushing and passing offenseses in the Big Ten last season.

```{r cowolot1, exercise=TRUE, exercise.setup = "cowplot-load-data", message=FALSE}
totals <- games |> 
  filter(Conference == "Big Ten Conference") |> 
  group_by(Team) |> 
  summarize(
    TotalRushes = sum(RushingAtt),
    TotalRushYds = sum(RushingYds),
    TotalPasses = sum(PassingAtt),
    TotalPassYds = sum(PassingYds)
  ) |>
  mutate(
    RushingAvg = TotalRushYds/TotalRushes,
    PassingAvg = TotalPassYds/TotalPasses
  ) 

toppass <- totals |> top_n(5, wt=PassingAvg)
toprush <- totals |> top_n(5, wt=RushingAvg)
```
```{r cowolot1-solution, exercise.reveal_solution = FALSE}
totals <- games |> 
  filter(Conference == "Big Ten Conference") |> 
  group_by(Team) |> 
  summarize(
    TotalRushes = sum(RushingAtt),
    TotalRushYds = sum(RushingYds),
    TotalPasses = sum(PassingAtt),
    TotalPassYds = sum(PassingYds)
  ) |>
  mutate(
    RushingAvg = TotalRushYds/TotalRushes,
    PassingAvg = TotalPassYds/TotalPasses
  ) 

toppass <- totals |> top_n(5, wt=PassingAvg)
toprush <- totals |> top_n(5, wt=RushingAvg)
```
```{r cowolot1-check}
grade_this_code()
```


### Exercise 1: Making bar charts into objects

Okay, now to do this I need to **save my plots to an object**. We do this the same way we save things to a dataframe -- with the arrow aka the assignment operator. We'll make two identical bar charts, one with the top passers and one with the top rushers. We'll call them bar1 and bar2.

```{r cowolot2, exercise=TRUE, exercise.setup = "cowplot-load-data", message=FALSE}
???? <- ggplot() + 
  geom_bar(data=???????, aes(x=reorder(Team, PassingAvg), weight=PassingAvg)) + 
  coord_flip()

???? <- ggplot() + 
  geom_bar(data=???????, aes(x=reorder(Team, RushingAvg), weight=RushingAvg)) + 
  coord_flip()
```
```{r cowolot2-solution, exercise.reveal_solution = FALSE}
bar1 <- ggplot() + 
  geom_bar(data=toppass, aes(x=reorder(Team, PassingAvg), weight=PassingAvg)) +
  coord_flip()

bar2 <- ggplot() + 
  geom_bar(data=toprush, aes(x=reorder(Team, RushingAvg), weight=RushingAvg)) +
  coord_flip()
```
```{r cowolot2-check}
grade_this_code()
```

### Exercise 2: Making a plot grid

With cowplot, we can use a function called `plot_grid` to arrange the charts. All we need to do is pass the names of the charts we just made into it.

```{r cowolot3, exercise=TRUE, exercise.setup = "cowplot-load-data", message=FALSE}
plot_grid(????, ????) 
```
```{r cowolot3-solution, exercise.reveal_solution = FALSE}
plot_grid(bar1, bar2) 
```
```{r cowolot3-check}
grade_this_code()
```

We can also stack them on top of each other with ncol=1, meaning number of columns equals 1.

```{r cowolot4, exercise=TRUE, exercise.setup = "cowplot-load-data", message=FALSE}
plot_grid(????, ????, ncol=1) 
```
```{r cowolot4-solution, exercise.reveal_solution = FALSE}
plot_grid(bar1, bar2, ncol=1) 
```
```{r cowolot4-check}
grade_this_code()
```

## Moving toward publication

To make these publishable, we should add headlines, chatter, decent labels, credit lines, etc. But to do this, we'll have to figure out which labels go on which charts, so we can make it look decent. For example -- both charts don't need x or y labels. If you don't have a title and subtitle on both, the spacing is off, so you need to leave one blank or the other blank. You'll just have to fiddle with it until you get it looking right. 

```{r cowolot5, exercise=TRUE, exercise.setup = "cowplot-load-data", message=FALSE}
bar1 <- ggplot() + 
  geom_bar(data=toppass, aes(x=reorder(Team, PassingAvg), weight=PassingAvg)) +
  coord_flip() + 
  labs(title="The B1G's top passing and rushing offenses", subtitle = "Ohio State, Michigan, Penn State and Minnesota appear on both lists.",  x="", y="Percent change", caption = "") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )

bar2 <- ggplot() + 
  geom_bar(data=toprush, aes(x=reorder(Team, RushingAvg), weight=RushingAvg), color="grey") +
  coord_flip() +  
  labs(title = "", subtitle= "", x="", y="",  caption="Source: NCAA | By Matt Waite") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )

plot_grid(bar1, bar2) 
```
```{r cowolot5-solution, exercise.reveal_solution = FALSE}
bar1 <- ggplot() + 
  geom_bar(data=toppass, aes(x=reorder(Team, PassingAvg), weight=PassingAvg)) +
  coord_flip() + 
  labs(title="The B1G's top passing and rushing offenses", subtitle = "Ohio State, Michigan, Penn State and Minnesota appear on both lists.",  x="", y="Percent change", caption = "") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )

bar2 <- ggplot() + 
  geom_bar(data=toprush, aes(x=reorder(Team, RushingAvg), weight=RushingAvg), color="grey") +
  coord_flip() +  
  labs(title = "", subtitle= "", x="", y="",  caption="Source: NCAA | By Matt Waite") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )

plot_grid(bar1, bar2) 
```
```{r cowolot5-check}
grade_this_code()
```

What's missing here? Color. Our eyes aren't drawn to anything (except maybe the top and bottom). So we need to help that. **A bar chart without context or color to draw attention to something isn't much of a bar chart**. Same with a line chart -- if your line chart has one line, no context, no color, it's going to fare poorly. To get some color and context in this, we're going to focus on ... a local team ... and add another geom_bar to our two bar charts using that data. We can then assign that one bar a color.

```{r cowolot6, exercise=TRUE, exercise.setup = "cowplot-load-data", message=FALSE}
nu <- totals |> filter(Team == "Nebraska")

bar1 <- ggplot() + 
  geom_bar(data=toppass, aes(x=reorder(Team, PassingAvg), weight=PassingAvg), fill="grey") +
  geom_bar(data=nu, aes(x=reorder(Team, PassingAvg), weight=PassingAvg), fill="red") +
  coord_flip() + 
  labs(title="The B1G's top passing and rushing offenses", subtitle = "Ohio State, Michigan, Penn State and Minnesota appear on both lists.",  x="", y="Percent change", caption = "") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank(),
    plot.title.position = "plot"
    )

bar2 <- ggplot() + 
  geom_bar(data=toprush, aes(x=reorder(Team, RushingAvg), weight=RushingAvg), fill="grey") +
  geom_bar(data=nu, aes(x=reorder(Team, RushingAvg), weight=RushingAvg), fill="red") +
  coord_flip() +  
  labs(title = "", subtitle= "", x="", y="",  caption="Source: NCAA | By Matt Waite") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank(),
    plot.title.position = "plot"
    )

plot_grid(bar1, bar2) 
```
```{r cowolot6-solution, exercise.reveal_solution = FALSE}
nu <- totals |> filter(Team == "Nebraska")

bar1 <- ggplot() + 
  geom_bar(data=toppass, aes(x=reorder(Team, PassingAvg), weight=PassingAvg), fill="grey") +
  geom_bar(data=nu, aes(x=reorder(Team, PassingAvg), weight=PassingAvg), fill="red") +
  coord_flip() + 
  labs(title="The B1G's top passing and rushing offenses", subtitle = "Ohio State, Michigan, Penn State and Minnesota appear on both lists.",  x="", y="Percent change", caption = "") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank(),
    plot.title.position = "plot"
    )

bar2 <- ggplot() + 
  geom_bar(data=toprush, aes(x=reorder(Team, RushingAvg), weight=RushingAvg), fill="grey") +
  geom_bar(data=nu, aes(x=reorder(Team, RushingAvg), weight=RushingAvg), fill="red") +
  coord_flip() +  
  labs(title = "", subtitle= "", x="", y="",  caption="Source: NCAA | By Matt Waite") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank(),
    plot.title.position = "plot"
    )

plot_grid(bar1, bar2)  
```
```{r cowolot6-check}
grade_this_code()
```

THREE GAMES. We won only THREE games last season. HOW?
