---
title: "Sports Data Lesson 11: Using packages to get data"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to use a library to import data instead of a csv.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(wehoop)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

# Using packages to get data

## The basics

There is a growing number of packages and repositories of sports data, largely because there's a growing number of people who want to analyze that data. We've done it ourselves with simple Google Sheets tricks. Then there's RVest, which is a method of scraping the data yourself from websites that you'll do later. But with these packages, someone has done the work of gathering the data for you. All you have to learn are the commands to get it.

One very promising collection of libraries is something called the [SportsDataverse](https://sportsdataverse.org/), which has a collection of packages covering specific sports, all of which are in various stages of development. Some are more complete than others, but they are all being actively worked on by developers. Packages of interest in this class are: 

* [cfbfastR, for college football](https://saiemgilani.github.io/cfbfastR/).
* [hoopR, for men's professional and college basketball](https://saiemgilani.github.io/hoopR/).
* [wehoop, for women's professional and college basketball](https://saiemgilani.github.io/wehoop/).
* [baseballr, for professional and college baseball](https://billpetti.github.io/baseballr/).
* [worldfootballR, for soccer data from around the world](https://jaseziv.github.io/worldfootballR/).
* [hockeyR, for NHL hockey data](https://hockeyr.netlify.app/)
* [recruitR, for college sports recruiting](https://saiemgilani.github.io/recruitR/)

Not part of the SportsDataverse, but in the same neighborhood, is [nflfastR](https://www.nflfastr.com/), which can provide NFL play-by-play data. 
Because they're all under development, not all of them can be installed with just a simple `install.packages("something")`. Some require a little work, some require API keys or paid accounts with certain providers.  

The main issue for you is to read the documentation carefully. 

## Using wehoop

wehoop presents us a good view into what using libraries like this are all about. And, as of this writing, Nebraska's women's basketball team is having a helluva season. 

First things first, we need to load our libraries.  

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
library(wehoop)
```
```{r load-tidyverse-solution}
library(tidyverse)
library(wehoop)
```
```{r load-tidyverse-check}
grade_this_code()
```

Normally, you would be reading the documentation for wehoop or any of the libraries in the Sportsdataverse, but some of them are a little thin (and hey, they're free, so I'm not complaining). The best part of [the documentation](https://wehoop.sportsdataverse.org/index.html) for wehoop is in the functional reference, which is a road map to everything it can do. One of the more interesting things to me with wehoop, cfbfastR and hoopR is play-by-play data. But that data gets very large, very fast. Let's sstart with something smaller, like player box scores. 

Here's how to get player box scores for the most recent season (which are numbered by the year they end in). 

```{r packages-load-data, message=FALSE, warning=FALSE, results=FALSE}
playerboxscores <- load_wbb_player_box(seasons = 2023)
```
```{r packages-load-data-exercise, exercise = TRUE}
playerboxscores <- load_wbb_player_box(seasons = 2023)
```
```{r packages-load-data-exercise-solution}
playerboxscores <- load_wbb_player_box(seasons = 2023)
```
```{r packages-load-data-exercise-check}
grade_this_code()
```

Let's `glimpse` the data because it's very wide and has a lot of columns. 

```{r glimpse-data, exercise=TRUE, exercise.setup = "packages-load-data"}
glimpse(playerboxscores)
```
```{r glimpse-data-solution}
glimpse(playerboxscores)
```
```{r glimpse-data-check}
grade_this_code()
```

Without noticing it, you've already seen the magic. You ran that one line of code, and now you have data, just like before. You didn't have to download a CSV. You didn't have to get it yourself. You just have data. The code you ran pulled the data from a remote resource, and now, like all the other data we've used all along, you can answer questions with it. 

### Exercise 1: Who has started the most for the Huskers?

If you look at the data, there is one field that is a TRUE/FALSE field of who started the game. Does Amy Williams tinker with her lineups?

First, since we have box scores for every team, we need to filter to just Nebraska. Then, we're going to use the & sign -- which is a way of saying AND. The filter is the team AND the starting status being true. Then we'll group by the player's name (look at glimpse) and count. This time, we're going to use a shortcut for counting called `tally`. 

```{r starters, exercise=TRUE, exercise.setup = "packages-load-data"}
playerboxscores |> 
  filter(team_short_display_name == "?????????" & starter == ????) |> 
  group_by(???????_???????_????) |> 
  tally(sort=TRUE) 
```
```{r starters-solution, exercise.reveal_solution = FALSE}
playerboxscores |> 
  filter(team_short_display_name == "Nebraska" & starter == TRUE) |> 
  group_by(athlete_display_name) |> 
  tally(sort=TRUE) 
```
```{r starters-check}
grade_this_code()
```

At least at the time of this writing, Coach Williams seems to be sticking with what works. 

## Another example

The wehoop package also has play by play data in it. A warning -- this step will take a beat or two to finish. It's going to pull in hundreds of thousands of plays. 

### Exercise 2: Get play-by-play data

The `load_wbb_pbp` function works very similar to the box scores. You're going to provide the season you want. We want this season.

```{r packages2-load-data, message=FALSE, warning=FALSE, results=FALSE}
playbyplay <- load_wbb_pbp(2023)
```
```{r packages2-load-data-exercise, exercise = TRUE}
playbyplay <- load_wbb_pbp(????)
```
```{r packages2-load-data-exercise-solution, exercise.reveal_solution = FALSE}
playbyplay <- load_wbb_pbp(2023)
```
```{r packages2-load-data-exercise-check}
grade_this_code()
```

And you'll want to glimpse it too. 

```{r glimpse2-data, exercise=TRUE, exercise.setup = "packages2-load-data"}
glimpse(playbyplay)
```
```{r glimpse2-data-solution}
glimpse(playbyplay)
```
```{r glimpse2-data-check}
grade_this_code()
```

There are a *lot* of questions you can ask and answer with play-by-play data if you have some creativity. You can filter down to the last two minutes of a game and seeing who is getting the ball. You can look at who makes the plays when the score is close. There's a lot we can do.

For now, let's try something simple -- what's the most common shooting play for Nebraska this season.

### Exercise 3: Shot's in the air, who is doing what?

Like the box scores, we're going to start with a filter. We're going to use both & and |. The | means OR. So we want shooting plays AND where the home team OR the away time is Nebraska. After that, it's group by and tally. 

```{r playbyplay, exercise=TRUE, exercise.setup = "packages2-load-data"}
?????????? |> 
  filter(shooting_play == ???? & home_team_name == "?????????" | away_team_name == "????????") |> 
  group_by(text) |> 
  tally(sort=TRUE)
```
```{r playbyplay-solution, exercise.reveal_solution = FALSE}
playbyplay |> 
  filter(shooting_play == TRUE & home_team_name == "Nebraska" | away_team_name == "Nebraska") |> 
  group_by(text) |> 
  tally(sort=TRUE)
```
```{r playbyplay-check}
grade_this_code()
```

The most common plays -- defensive rebounds and free throws. But if you look at basketball stats, that makes sense. There's a lot more defensive rebounds than offensive. There's a lot more missed threes than made ones, typically. But as Nebraska goes further on this run, you now have the ability to get very granular data to analyze how Amy Williams and the team are doing it. 
