---
title: "Sports Data Lesson 21: Beeswarm plots"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to make scatterplots grouped by categories on a number line.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(ggrepel)
library(ggbeeswarm)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

# Beeswarm plots

## The basics

A beeswarm plot is sometimes called a column scatterplot. It's an effective way to show how individual things -- teams, players, etc. -- are distributed along a numberline. The column is a grouping -- say positions in basketball -- and the dots are players, and the dots cluster where the numbers are more common. So think of it like a histogram mixed with a scatterplot crossed with a bar chart. 

An example will help.

Let's use the NU Women's Volleyball team as our example. Let's look at a new metric: points. Who is doing the work for NU? 

Here is that data. Per usual: **For purposes of this exercise, you don't need to do this. The data is included here if you want to try this in your own notebook.**

```{r echo=FALSE, class.output="bg-info", results="asis",  message=FALSE,  warning=FALSE}
library(downloadthis)
library(glue)

dllink <- download_link(
  link = "https://mattwaite.github.io/sportsdatafiles/ncaa_womens_volleyball_playerstats_2023.csv",
  button_label = "Download csv file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

glue("<pre><p><strong>For this walkthrough:</strong></p><p>{dllink}</p></pre>")
```

To make beeswarm plots, you need a library that adds some geoms to ggplot. In this cases it's called ggbeeswarm, and you installed it way back at the beginning. But any time you need a library, and it's on CRAN, you can go to your console and install it with `install.packages("ggbeeswarm")`

We'll need to load ggbeeswarm, the tidyverse and, for later, ggrepel.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
library(ggbeeswarm)
library(ggrepel)
```
```{r load-tidyverse-solution}
library(tidyverse)
library(ggbeeswarm)
library(ggrepel)
```
```{r load-tidyverse-check}
grade_this_code()
```

And the data.

```{r beeswarm-load-data, message=FALSE, warning=FALSE}
players <- read_csv("https://mattwaite.github.io/sportsdatafiles/ncaa_womens_volleyball_playerstats_2023.csv")

set.seed(1234)

activeplayers <- players |> filter(s > 75 & kills > 10 & is.na(pos) == FALSE)

nu <- activeplayers |> 
  filter(team == "Nebraska Huskers")
```
```{r beeswarm-load-data-exercise, exercise = TRUE}
players <- read_csv("https://mattwaite.github.io/sportsdatafiles/ncaa_womens_volleyball_playerstats_2023.csv")
```
```{r beeswarm-load-data-exercise-solution}
players <- read_csv("https://mattwaite.github.io/sportsdatafiles/ncaa_womens_volleyball_playerstats_2023.csv")
```
```{r beeswarm-load-data-exercise-check}
grade_this_code()
```

Another bit of setup: we need to set the seed for the random number generator. The library "jitters" the dots in the beeswarm randomly. If we don't set the seed, we'll get different results each time. Setting the seed means we get the same look. You *can* use any number you want. I use 1234, so that's what we'll use here.

```{r beeswarm1, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
set.seed(1234)
```
```{r beeswarm1-solution, exercise.reveal_solution = FALSE}
set.seed(1234)
```
```{r beeswarm1-check}
grade_this_code()
```

We know this data has a lot of players who didn't play, so let's get rid of them.

### Exercise 1: Active players only please.

Recall in an earlier lesson, we used `filter` to get rid of players who don't play much. Let's do that again, dumping everyone who hasn't played 75 sets -- `s` in the data -- and we're going to get rid of every player with 10 or fewer kills. Believe it or not, there are players who have appeared in 75 sets who didn't rack up a single kill. We'll name our dataframe activeplayers.

```{r beeswarm2, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
____ <- ____ |> 
  filter(
    s > ____ & 
      kills > ____ & 
      is.na(pos) == FALSE)
```
```{r beeswarm2-solution, exercise.reveal_solution = FALSE}
activeplayers <- players |> 
  filter(
    s > 75 & 
      kills > 10 & 
      is.na(pos) == FALSE)
```
```{r beeswarm2-check}
grade_this_code()
```

### Exercise 2: Your first beeswarm

It works very much like you would expect, if you think about it. The group value -- the columns that you put your dots in -- is the x, the number is the y. We're going to beeswarm by position -- `pos` -- and the dots will be `kills`. 

```{r beeswarm5, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
ggplot() + 
  geom_beeswarm(data=____, aes(x=____, y=____), color="grey")
```
```{r beeswarm5-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_beeswarm(data=activeplayers, aes(x=pos, y=kills), color="grey")
```
```{r beeswarm5-check}
grade_this_code()
```

You can see that there's a lot of setters who are getting kills, but not a lot. It's not their job. Same with Defensive Specialists and Liberos. You'll also see there's some creativity with naming positions -- OPP for opposite side hitter, which is the same thing as an RS, or right side hitter, which are really all Outside Hitters, or OH. A middle hitter -- MH -- is just another name for MB or middle blocker. If we were going to publish this, we'd have to fix that. 

So where are the Nebraska players in that mix? 

### Exercise 3: Filtering for Nebraska

We'll filter players on Nebraska who meet our criteria. Weirdly, this is the only dataset we have that calls Nebraska the Huskers, so be mindful of that when you filter. 

```{r beeswarm6, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
nu <- activeplayers |> 
  filter(team == "____ ____")
```
```{r beeswarm6-solution, exercise.reveal_solution = FALSE}
nu <- activeplayers |> 
  filter(team == "Nebraska Huskers")
```
```{r beeswarm6-check}
grade_this_code()
```

There's six Huskers who have played in more than 75 sets and who have more than 10 kills. 

But how good are they as killers compared to the rest of college volleyball?  

### Exercise 4: Layering in Nebraska

When you add another beeswarm, we need to pass another element in -- we need to tell it if we're grouping on the x value. Not sure why -- it's a requirement of the library -- but you'll get a warning if you don't. 

```{r beeswarm7, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
ggplot() + 
  geom_beeswarm(
    data=____, 
    aes(x=____, y=____), color="grey") + 
  geom_beeswarm(
    data=____, 
    aes(x=____, y=____), color="red")
```
```{r beeswarm7-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_beeswarm(
    data=activeplayers, 
    aes(x=pos, y=kills), color="grey") + 
  geom_beeswarm(
    data=nu, 
    aes(x=pos, y=kills), color="red")
```
```{r beeswarm7-check}
grade_this_code()
```

What does this say about Nebraska? The number of players, where they rank, by position?

### Exercise 5: Labeling 

This is where we can use ggrepel. Let's add a text layer and label the dots. 

```{r beeswarm8, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
ggplot() + 
  geom_beeswarm(
    data=____, 
    aes(x=____, y=____), color="grey") + 
  geom_beeswarm(
    data=____, 
    aes(x=____, y=____), color="red") + 
  geom_text_repel(
    data=nu, 
    aes(x=____, y=____, label=full_name))
```
```{r beeswarm8-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_beeswarm(
    data=activeplayers, 
    aes(x=pos, y=kills), color="grey") + 
  geom_beeswarm(
    data=nu, 
    aes(x=pos, y=kills), color="red") + 
  geom_text_repel(
    data=nu, 
    aes(x=pos, y=kills, label=full_name))
```
```{r beeswarm8-check}
grade_this_code()
```

What stories do you see? 

## A few other options

The ggbeeswarm library has a couple of variations on the geom_beeswarm that may work better for your application. They are `geom_quasirandom` and `geom_jitter`. 

There's not a lot to change from our example to see what they do. 

```{r beeswarm9, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
ggplot() + 
  geom_quasirandom(
    data=activeplayers, 
    aes(x=pos, y=kills), color="grey") + 
  geom_quasirandom(
    data=nu, 
    aes(x=pos, y=kills), color="red") + 
  geom_text_repel(
    data=nu, 
    aes(x=pos, y=kills, label=full_name))
```
```{r beeswarm9-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_quasirandom(
    data=activeplayers, 
    aes(x=pos, y=kills), color="grey") + 
  geom_quasirandom(
    data=nu, 
    aes(x=pos, y=kills), color="red") + 
  geom_text_repel(
    data=nu, 
    aes(x=pos, y=kills, label=full_name))
```
```{r beeswarm9-check}
grade_this_code()
```

Quasirandom spreads out the dots you see in beeswarm using -- you guessed it -- quasirandom spacing. 

```{r beeswarm10, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
ggplot() + 
  geom_jitter(
    data=activeplayers, 
    aes(x=pos, y=kills), color="grey") + 
  geom_jitter(
    data=nu, 
    aes(x=pos, y=kills), color="red") + 
  geom_text_repel(
    data=nu, 
    aes(x=pos, y=kills, label=full_name))
```
```{r beeswarm10-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_jitter(
    data=activeplayers, 
    aes(x=pos, y=kills), color="grey") + 
  geom_jitter(
    data=nu, 
    aes(x=pos, y=kills), color="red") + 
  geom_text_repel(
    data=nu, 
    aes(x=pos, y=kills, label=full_name))
```
```{r beeswarm10-check}
grade_this_code()
```

`geom_jitter` spreads out the dots evenly across the width of the column, randomly deciding where in the line of the kills they appear. 

Which one is right for you? You're going to have to experiment and decide. This is the art in the art and a science. 
