---
title: "Sports Data Lesson 21: Beeswarm plots"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to make scatterplots grouped by categories on a number line.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(ggrepel)
library(ggbeeswarm)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

The goal of this lesson is to introduce you to beeswarm plots, an innovative visualization technique that combines elements of scatterplots, histograms, and bar charts. By the end of this tutorial, you'll understand when to use beeswarm plots, how to create them using the ggbeeswarm package in R, and how to interpret the distributions they reveal. You'll learn to prepare data for beeswarm plots, create basic plots, layer data to highlight specific groups, and add labels using ggrepel. We'll apply these concepts to real college volleyball data, analyzing player performance across different positions. You'll also explore variations like quasirandom and jitter plots. This lesson will enhance your ability to visualize the distribution of individual data points within categories, a crucial skill for understanding player performance patterns and making nuanced comparisons in sports analytics.

## The Basics

A beeswarm plot is sometimes called a column scatterplot. It's an effective way to show how individual things -- teams, players, etc. -- are distributed along a numberline. The column is a grouping -- say positions in basketball -- and the dots are players, and the dots cluster where the numbers are more common. So think of it like a histogram mixed with a scatterplot crossed with a bar chart. 

An example will help.

Let's use the NU Women's Volleyball team as our example. Let's look at a new metric: points. Who is doing the work for NU? 

To make beeswarm plots, you need a library that adds some geoms to ggplot. In this cases it's called ggbeeswarm, and you installed it way back at the beginning. But any time you need a library, and it's on CRAN, you can go to your console and install it with `install.packages("ggbeeswarm")`

We'll need to load ggbeeswarm, the tidyverse and, for later, ggrepel.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
library(ggbeeswarm)
library(ggrepel)
```
```{r load-tidyverse-solution}
library(tidyverse)
library(ggbeeswarm)
library(ggrepel)
```
```{r load-tidyverse-check}
grade_this_code()
```

And the data.

```{r beeswarm-load-data, message=FALSE, warning=FALSE}
players <- read_csv("https://mattwaite.github.io/sportsdatafiles/ncaa_womens_volleyball_playerstats_2023.csv")

set.seed(1234)

activeplayers <- players |> filter(s > 75 & kills > 10 & is.na(pos) == FALSE)

nu <- activeplayers |> 
  filter(team == "Nebraska Huskers")
```
```{r beeswarm-load-data-exercise, exercise = TRUE}
players <- read_csv("https://mattwaite.github.io/sportsdatafiles/ncaa_womens_volleyball_playerstats_2023.csv")
```
```{r beeswarm-load-data-exercise-solution}
players <- read_csv("https://mattwaite.github.io/sportsdatafiles/ncaa_womens_volleyball_playerstats_2023.csv")
```
```{r beeswarm-load-data-exercise-check}
grade_this_code()
```

Another bit of setup: we need to set the seed for the random number generator. The library "jitters" the dots in the beeswarm randomly. If we don't set the seed, we'll get different results each time. Setting the seed means we get the same look. You *can* use any number you want. I use 1234, so that's what we'll use here.

```{r beeswarm1, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
set.seed(1234)
```
```{r beeswarm1-solution, exercise.reveal_solution = FALSE}
set.seed(1234)
```
```{r beeswarm1-check}
grade_this_code()
```

We know this data has a lot of players who didn't play, so let's get rid of them.

### Exercise 1: Active players only please.

Recall in an earlier lesson, we used `filter` to get rid of players who don't play much. Let's do that again, dumping everyone who hasn't played 75 sets -- `s` in the data -- and we're going to get rid of every player with 10 or fewer kills. Believe it or not, there are players who have appeared in 75 sets who didn't rack up a single kill. We'll name our dataframe activeplayers.

```{r beeswarm2, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
____ <- ____ |> 
  filter(
    s > ____ & 
      kills > ____ & 
      is.na(pos) == FALSE)
```
```{r beeswarm2-solution, exercise.reveal_solution = FALSE}
activeplayers <- players |> 
  filter(
    s > 75 & 
      kills > 10 & 
      is.na(pos) == FALSE)
```
```{r beeswarm2-check}
grade_this_code()
```

### Exercise 2: Your first beeswarm

It works very much like you would expect, if you think about it. The group value -- the columns that you put your dots in -- is the x, the number is the y. We're going to beeswarm by position -- `pos` -- and the dots will be `kills`. 

```{r beeswarm5, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
ggplot() + 
  geom_beeswarm(data=____, aes(x=____, y=____), color="grey")
```
```{r beeswarm5-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_beeswarm(data=activeplayers, aes(x=pos, y=kills), color="grey")
```
```{r beeswarm5-check}
grade_this_code()
```

You can see that there's a lot of setters who are getting kills, but not a lot. It's not their job. Same with Defensive Specialists and Liberos. You'll also see there's some creativity with naming positions -- OPP for opposite side hitter, which is the same thing as an RS, or right side hitter, which are really all Outside Hitters, or OH. A middle hitter -- MH -- is just another name for MB or middle blocker. If we were going to publish this, we'd have to fix that. 

So where are the Nebraska players in that mix? 

### Exercise 3: Filtering for Nebraska

We'll filter players on Nebraska who meet our criteria. Weirdly, this is the only dataset we have that calls Nebraska the Huskers, so be mindful of that when you filter. 

```{r beeswarm6, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
nu <- activeplayers |> 
  filter(team == "____ ____")
```
```{r beeswarm6-solution, exercise.reveal_solution = FALSE}
nu <- activeplayers |> 
  filter(team == "Nebraska Huskers")
```
```{r beeswarm6-check}
grade_this_code()
```

There's six Huskers who have played in more than 75 sets and who have more than 10 kills. 

But how good are they as killers compared to the rest of college volleyball?  

### Exercise 4: Layering in Nebraska

When you add another beeswarm, we need to pass another element in -- we need to tell it if we're grouping on the x value. Not sure why -- it's a requirement of the library -- but you'll get a warning if you don't. 

```{r beeswarm7, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
ggplot() + 
  geom_beeswarm(
    data=____, 
    aes(x=____, y=____), color="grey") + 
  geom_beeswarm(
    data=____, 
    aes(x=____, y=____), color="red")
```
```{r beeswarm7-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_beeswarm(
    data=activeplayers, 
    aes(x=pos, y=kills), color="grey") + 
  geom_beeswarm(
    data=nu, 
    aes(x=pos, y=kills), color="red")
```
```{r beeswarm7-check}
grade_this_code()
```

What does this say about Nebraska? The number of players, where they rank, by position?

### Exercise 5: Labeling 

This is where we can use ggrepel. Let's add a text layer and label the dots. 

```{r beeswarm8, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
ggplot() + 
  geom_beeswarm(
    data=____, 
    aes(x=____, y=____), color="grey") + 
  geom_beeswarm(
    data=____, 
    aes(x=____, y=____), color="red") + 
  geom_text_repel(
    data=nu, 
    aes(x=____, y=____, label=full_name))
```
```{r beeswarm8-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_beeswarm(
    data=activeplayers, 
    aes(x=pos, y=kills), color="grey") + 
  geom_beeswarm(
    data=nu, 
    aes(x=pos, y=kills), color="red") + 
  geom_text_repel(
    data=nu, 
    aes(x=pos, y=kills, label=full_name))
```
```{r beeswarm8-check}
grade_this_code()
```

What stories do you see? 

## A few other options

The ggbeeswarm library has a couple of variations on the geom_beeswarm that may work better for your application. They are `geom_quasirandom` and `geom_jitter`. 

There's not a lot to change from our example to see what they do. 

```{r beeswarm9, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
ggplot() + 
  geom_quasirandom(
    data=activeplayers, 
    aes(x=pos, y=kills), color="grey") + 
  geom_quasirandom(
    data=nu, 
    aes(x=pos, y=kills), color="red") + 
  geom_text_repel(
    data=nu, 
    aes(x=pos, y=kills, label=full_name))
```
```{r beeswarm9-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_quasirandom(
    data=activeplayers, 
    aes(x=pos, y=kills), color="grey") + 
  geom_quasirandom(
    data=nu, 
    aes(x=pos, y=kills), color="red") + 
  geom_text_repel(
    data=nu, 
    aes(x=pos, y=kills, label=full_name))
```
```{r beeswarm9-check}
grade_this_code()
```

Quasirandom spreads out the dots you see in beeswarm using -- you guessed it -- quasirandom spacing. 

```{r beeswarm10, exercise=TRUE, exercise.setup = "beeswarm-load-data", message=FALSE}
ggplot() + 
  geom_jitter(
    data=activeplayers, 
    aes(x=pos, y=kills), color="grey") + 
  geom_jitter(
    data=nu, 
    aes(x=pos, y=kills), color="red") + 
  geom_text_repel(
    data=nu, 
    aes(x=pos, y=kills, label=full_name))
```
```{r beeswarm10-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_jitter(
    data=activeplayers, 
    aes(x=pos, y=kills), color="grey") + 
  geom_jitter(
    data=nu, 
    aes(x=pos, y=kills), color="red") + 
  geom_text_repel(
    data=nu, 
    aes(x=pos, y=kills, label=full_name))
```
```{r beeswarm10-check}
grade_this_code()
```

`geom_jitter` spreads out the dots evenly across the width of the column, randomly deciding where in the line of the kills they appear. 

Which one is right for you? You're going to have to experiment and decide. This is the art in the art and a science. 

## The Recap

In this lesson, we've explored the power of beeswarm plots for visualizing the distribution of individual data points within categories in sports data. We started by learning how to prepare data, filtering for active players to ensure meaningful visualizations. We then progressed from creating basic beeswarm plots to more complex layered plots, highlighting specific teams within the context of all players. We covered important techniques such as setting a random seed for reproducibility, using color to differentiate groups, and adding non-overlapping labels with ggrepel. We also explored variations of the beeswarm plot, including quasirandom and jitter plots, discussing how each can offer different insights into data distribution. Throughout the lesson, we've seen how beeswarm plots can reveal patterns in player performance across different positions in volleyball, allowing for nuanced comparisons between teams and the broader player population. Remember, while beeswarm plots are powerful tools for visualizing categorical-numerical relationships, the choice between beeswarm, quasirandom, or jitter plots often depends on your specific data and the story you're trying to tell. Always experiment with different approaches to find the most effective visualization for your sports analytics insights.

## Terms to Know

- Beeswarm Plot: A type of plot that displays the distribution of individual data points within categories, combining elements of scatterplots, histograms, and bar charts.
- `ggbeeswarm`: An R package that extends ggplot2 to create beeswarm and related plots.
- `geom_beeswarm()`: A function from the ggbeeswarm package used to create beeswarm plots in ggplot2.
- `set.seed()`: An R function used to set a seed for the random number generator, ensuring reproducibility in plots with random elements.
- `Jittering`: A technique of adding small random offsets to data points to reduce overplotting in visualizations.
- `geom_quasirandom()`: A function from ggbeeswarm that creates a variation of the beeswarm plot with quasirandom point placement.
- `geom_jitter()`: A function that creates a plot similar to a beeswarm plot but with points spread evenly across the width of each category.