---
title: "Sports Data Lesson 4: Aggregates, Part 2"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn more about how to take lots of little things and total them up into bigger things.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(glue)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

The goal of this lesson is to take what we did with the `group_by()` and `summarise()` functions that you learned about in the last lesson and add some new tools to it. Not everything is a count. Sometimes, you need an average. Or a median. Or the biggest number. Or the smallest number. By the end of this lesson, you'll be able to answer questions like "What position averages the most kills?" The examples will use NCAA women's volleyball statistics, but the techniques you learn can be applied to any sport or dataset.

## The Basics

Hopefully, this is starting to feel a little familiar. Remember: It's all a pattern. Just about every single time it's the same start. Load libraries, load data, look at the data, answer the questions. 

Let's get to work.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
```

```{r load-tidyverse-solution}
library(tidyverse)
```

```{r load-tidyverse-check}
grade_this_code()
```

Now, we load the data. 

```{r load-data, message=FALSE, warning=FALSE}
volleyballplayers <- read_csv("https://mattwaite.github.io/sportsdatafiles/ncaa_womens_volleyball_playerstats_2023.csv")

nplayers <- nrow(volleyballplayers)

avgsets <- volleyballplayers |>
  group_by(pos) |>
  summarise(
    count = n(),
    mean_sets = mean(s),
    median_sets = median(s)
  ) |>
  arrange(desc(mean_sets)) |> slice(1)

killsleader <- volleyballplayers |> arrange(desc(kills)) |> slice(1)
```

```{r load-data-exercise, exercise = TRUE}
volleyballplayers <- read_csv("https://mattwaite.github.io/sportsdatafiles/ncaa_womens_volleyball_playerstats_2023.csv")
```

```{r load-data-exercise-solution}
volleyballplayers <- read_csv("https://mattwaite.github.io/sportsdatafiles/ncaa_womens_volleyball_playerstats_2023.csv")
```

```{r load-data-exercise-check}
grade_this_code()
```

Now, let's get a peek at our data. 

```{r head-data, exercise=TRUE, exercise.setup = "load-data"}
head(_____)
```

```{r head-data-solution, exercise.reveal_solution = FALSE}
head(volleyballplayers)
```

```{r head-data-check}
grade_this_code()
```

::: {#head-data-hint}
**Hint:** The thing you need is to the left of a \<- in a block above.
:::

Now that we're set up, we can answer more questions with data. 

### Exercise 1: More in summarize

In the last example, we grouped some data together and counted it up, but there's so much more you can do. 

Sticking with our NCAA volleyball player data, we can calculate any number of measures inside summarize. Here, we'll use R's built in `mean` function to calculate ... well, you get the idea.

Let's look just a the number of sets (s) each position gets. **In summarize, what you do left of the equal sign is giving something a name**. We should name our new column what it is: `mean_sets`. Then let's arrange it by the average number of sets (hint: that wording, average number of sets, is a trick. You arrange by the name of a column created in summarize).

<div class="alert alert-warning">
  <h4 class="alert-heading">Common Mistake</h4>
  <p class="mb-0">What goes inside <code>mean()</code> has to be a column name. You can find those above where you ran <code>head()</code>.</p>
</div>

```{r group-by-4, exercise=TRUE, exercise.setup = "load-data"}
_____ |>
  group_by(___) |>
  summarise(
    ____ = mean(____)
  ) |>
  arrange(desc(____))
```

```{r group-by-4-solution, exercise.reveal_solution = FALSE}
volleyballplayers |>
  group_by(pos) |>
  summarise(
    mean_sets = mean(s)
  ) |>
  arrange(desc(mean_sets))
```

```{r group-by-4-check}
grade_this_code()
```

::: {#group-by-4-hint}
**Hint:** What goes in `group_by` and `mean()` all come from column names you find with `head()`. What goes before the = are names in the previous paragraph.
:::

```{r avgsets, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "load-data", results='asis'}
glue("So there's {nplayers} players in the data. Of them, {avgsets$count} are of the position {avgsets$pos}, which if you know volleyball you can figure out what that is pretty quickly. The average {avgsets$pos} plays in {format(avgsets$mean_sets, digits=1)} sets on average.")
```

### Exercise 2: More than one number

You can do multiple measures in a single `summarize` step as well. There's a simple way to think about this pattern. You might have noticed in the previous exercise, you first give your column a name, then an equal sign, then you do some math stuff. In the last exercise, it was `mean()`. All of that -- `nameofcolumn = math()` creates a column with a name and puts numbers in it. To add another one, just add a comma to the end of that line, and repeat the pattern. 

Let's add median_sets to our data analysis and arrange it by that. 

```{r group-by-5, exercise=TRUE, exercise.setup = "load-data"}
_____ |>
  group_by(___) |>
  summarise(
    ____ = mean(____),
    ____ = median(____),    
  ) |>
  arrange(desc(____))
```

```{r group-by-5-solution, exercise.reveal_solution = FALSE}
volleyballplayers |>
  group_by(pos) |>
  summarise(
    mean_sets = mean(s),
    median_sets = median(s)
  ) |>
  arrange(desc(median_sets))
```

```{r group-by-5-check}
grade_this_code()
```

::: {#group-by-5-hint}
**Hint:** What goes in `group_by` and `mean()` all come from column names you find with `head()`. What goes before the = are names in the previous paragraph.
:::

### Exercise 3: Arranging two ways

One thing to keep in mind is this: You can use any of these verbs -- group_by, summarize, arrange -- on their own without the others. For example, what player had the most kills in 2021? To find this out, we don't need a group_by or summarize, so what goes in arrange is just the field name. Let's put this in order of kills.

First let's get the most kills.

```{r group-by-6, exercise=TRUE, exercise.setup = "load-data"}
volleyballplayers |> arrange(desc(____))
```

```{r group-by-6-solution, exercise.reveal_solution = FALSE}
volleyballplayers |> arrange(desc(kills))
```

```{r group-by-6-check}
grade_this_code()
```

Now we can get the least by removing desc() and just doing arrange by our field name.

<div class="alert alert-info">
  <h4 class="alert-heading">Key Concept</h4>
  <p class="mb-0">The default behavior of arrange is smallest to largest. If you do arrange(touchdowns), you'll get a list of the fewest touchdowns first. `desc()` reverses that.</p>
</div>

```{r group-by-7, exercise=TRUE, exercise.setup = "load-data"}
volleyballplayers |> arrange(____)
```

```{r group-by-7-solution, exercise.reveal_solution = FALSE}
volleyballplayers |> arrange(kills)
```

```{r group-by-7-check}
grade_this_code()
```

```{r killer, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "load-data", results='asis'}
glue("The kill leader in the NCAA? {killsleader$full_name} of the {killsleader$team} with {killsleader$kills} kills. The lowest? A whole lot of people with 0.")
```

That's a huge difference.

So when choosing a measure of the middle, you have to ask yourself -- could I have extremes? Because a median won't be sensitive to extremes. It will be the point at which half the numbers are above and half are below. The average or mean will be a measure of the middle, but if you have a bunch of pine riders and then one iron superstar, the average could be wildly skewed.

We'll work more on that in the future.

### Exercise 4: Even more aggregates

There's a ton of things we can do in summarize -- we'll work with more of them as the course progresses -- but here's a few other questions you can ask.

Which position in the NCAA generates the most points? And what is the highest and lowest point total for that position? And how wide is the spread between points? We can find that with `sum` to add up the points to get the total points, `min` to find the minimum points, `max` to find the maximum points.

```{r group-by-8, exercise=TRUE, exercise.setup = "load-data"}
volleyballplayers |> 
  group_by(____) |> 
  summarise(
    total = sum(pts), 
    avg_points = mean(pts), 
    min_points = min(pts),
    max_points = max(pts),
    stdev_points = sd(pts)) |> 
  arrange(desc(total))
```

```{r group-by-8-solution, exercise.reveal_solution = FALSE}
volleyballplayers |> 
  group_by(pos) |> 
  summarise(
    total = sum(pts), 
    avg_points = mean(pts), 
    min_points = min(pts),
    max_points = max(pts),
    stdev_points = sd(pts)) |> 
  arrange(desc(total))
```

```{r group-by-8-check}
grade_this_code()
```

::: {#group-by-8-hint}
**Hint:** Breathe deep. Slow down. Think about it. What's missing here is stuff you've done already.
:::

<div class="alert alert-warning">
  <h4 class="alert-heading">Common Mistake</h4>
  <p class="mb-0">So many students of this class have struggled with the simple difference between a count (<code>n()</code>) and a sum (<code>sum()</code>). Let's pretend I have 5 players who have each scored 2 points. If I count them, I get 5. If I sum them, I get 10 -- 2+2+2+2+2 = 10.</p>
</div>

No surprise with our data analysis, outside hitters with the most points. Most of the stats here we can intuitively understand -- a minimum of zero means they probably didn't play at all. A large max means that player was good a volleyball. The one that you should give an extra check is the standard deviation. When your standard deviation is large, that's a sign you've got some huge points players and a bunch of bench players and you should consider which measure of the middle you're going to use.

## The Recap

In data analysis, `group_by` and `summarize` are two of the most basic, but most common functions. With these functions, you can take every game played and turn it into season totals, team statistics and many others. You can take every pitch thrown and look at them by type. You can look at a volleyball team in conference and out of conference. In just two lessons, you've learned a *huge* amount of basic data analysis. You'll use this pattern *a lot*. What remains is creative application of what you've learned.

## Terms to Know

- **mean()**: A function that calculates the arithmetic average of a set of numbers.
- **median()**: A function that finds the middle value in a sorted set of numbers.
- **sum()**: A function that adds up all the values in a specified column.
- **arrange()**: A function used to order rows in a dataset based on values in specified columns.
- **desc()**: A function used within arrange() to specify descending order for sorting.
