---
title: "Lesson 31: Rvest"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to scrape tables from websites.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(rvest)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

# Intro to rvest

## The basics

All the way back in Lesson 2, we used Google Sheets and importHTML to get our own data out of a website. For me, that's a lot of pointing and clicking and copying and pasting. R has a library that can automate the harvesting of data from HTML on the internet. It's called `rvest`. 

Let's grab [a simple, basic HTML table from College Football Stats](http://www.cfbstats.com/2022/leader/national/team/offense/split01/category09/sort01.html). There's nothing particularly strange about this table -- it's simply formatted and easy to scrape. 

First we'll need some libraries. We're going to use a library called `rvest`, which you already have installed. But like always, if you are having problems, you can install it by running `install.packages('rvest')` in the console. 

```{r load-tidyverse, exercise=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```
```{r load-tidyverse-solution}
library(tidyverse)
library(rvest)
```
```{r load-tidyverse-check}
grade_this_code()
```

The rvest package has functions that make fetching, reading and parsing HTML simple. The first thing we need to do is specify a url that we're going to scrape.

```{r rvest-load-data, message=FALSE, warning=FALSE}
scoringoffenseurl <- "http://www.cfbstats.com/2022/leader/national/team/offense/split01/category09/sort01.html"

scoringoffense <- scoringoffenseurl %>%
  read_html() %>%
  html_nodes(xpath = '//*[@id="content"]/div[2]/table') %>%
  html_table()

scoring <- scoringoffense[[1]]

scoringoffenseurl21 <- "http://www.cfbstats.com/2022/leader/national/team/offense/split01/category09/sort01.html"

scoringoffense21 <- scoringoffenseurl21 %>%
  read_html() %>%
  html_nodes(xpath = '//*[@id="content"]/div[2]/table') %>%
  html_table()

scoringoffense2021 <- scoringoffense21[[1]]

fixedscoringoffense2021 <- scoringoffense2021 %>% rename(Rank = 1)%>% mutate(YEAR = 2021)
fixedscoringoffense2022 <- scoring %>% rename(Rank = 1) %>% mutate(YEAR = 2022)

combined <- bind_rows(fixedscoringoffense2021, fixedscoringoffense2022)

url <- "https://www.sports-reference.com/cbb/seasons/2021-school-stats.html"

stats <- url %>%
  read_html() %>%
  html_nodes(xpath = '//*[@id="basic_school_stats"]') %>%
  html_table(header=FALSE)

stats <- stats[[1]]

statsrenamed <- stats %>% 
  rename(
    Rank=X1, 
    School=X2, 
    Games=X3, 
    OverallWins=X4, 
    OverallLosses=X5, 
    WinPct=X6, 
    OverallSRS=X7, 
    OverallSOS=X8, 
    Blank1=X9, 
    ConferenceWins=X10, 
    ConferenceLosses=X11, 
    Blank2=X12, 
    HomeWins=X13, 
    HomeLosses=X14, 
    Blank3=X15, 
    AwayWins=X16, 
    AwayLosses=X17, 
    Blank4=X18, 
    ForPoints=X19, 
    OppPoints=X20, 
    Blank5=X21, 
    Minutes=X22, 
    FieldGoalsMade=X23, 
    FieldGoalsAttempted=X24, 
    FieldGoalPCT=X25, 
    ThreePointMade=X26, 
    ThreePointAttempts=X27, 
    ThreePointPct=X28, 
    FreeThrowsMade=X29, 
    FreeThrowsAttempted=X30, 
    FreeThrowPCT=X31, 
    OffensiveRebounds=X32, 
    TotalRebounds=X33, 
    Assists=X34, 
    Steals=X35, 
    Blocks=X36, 
    Turnovers=X37, 
    PersonalFouls=X38)

statswithoutgarbage <- statsrenamed %>% filter(Rank != "Rk" & Games != "Overall") 

statswithnumbers <- statswithoutgarbage %>% mutate_at(vars(-School), as.numeric) %>% select(-starts_with("Blank"))
```
```{r rvest-load-data-exercise, exercise = TRUE}
scoringoffenseurl <- "http://www.cfbstats.com/2022/leader/national/team/offense/split01/category09/sort01.html"
```
```{r rvest-load-data-exercise-solution}
scoringoffenseurl <- "http://www.cfbstats.com/2022/leader/national/team/offense/split01/category09/sort01.html"
```
```{r rvest-load-data-exercise-check}
grade_this_code()
```

Now, the most difficult part of scraping data from any website is knowing what exact HTML tag you need to grab. In this case, we want a `<table>` tag that has all of our data table in it. But how do you tell R which one that is? Well, it's easy, once you know what to do. But it's not simple. So I've made a short video to show you how to find it. 

<iframe width="560" height="315" src="https://www.youtube.com/embed/kYkSE3zWa9Y" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

When you have simple tables, the code is very simple. You create a variable to receive the data, then pass it the url, read the html that was fetched, find the node you need using your XPath value you just copied and you tell rvest that it's a table. 

### Exercise 1: Copy the XPath and scrape

Watch the video, go to the URL we're scraping and copy the XPath for the table. Paste it into the code below.

```{r rvest1, exercise=TRUE, exercise.setup = "rvest-load-data", message=FALSE}
scoringoffense <- scoringoffenseurl %>%
  read_html() %>%
  html_nodes(xpath = '???????????????????????????????') %>%
  html_table()
```
```{r rvest1-solution, exercise.reveal_solution = FALSE}
scoringoffense <- scoringoffenseurl %>%
  read_html() %>%
  html_nodes(xpath = '//*[@id="content"]/div[2]/table') %>%
  html_table()
```
```{r rvest1-check}
grade_this_code()
```

What we get from this is ... not a dataframe. It's a list with one element in it, which just so happens to be our dataframe. When you get this, the solution is simple: just write to a new variable with the first list element.

```{r rvest2, exercise=TRUE, exercise.setup = "rvest-load-data", message=FALSE}
scoring <- scoringoffense[[1]]
```
```{r rvest2-solution, exercise.reveal_solution = FALSE}
scoring <- scoringoffense[[1]]
```
```{r rvest2-check}
grade_this_code()
```

And what do we have? 

```{r head-data, exercise=TRUE, exercise.setup = "rvest-load-data"}
head(scoring)
```
```{r head-data-solution}
head(scoring)
```
```{r head-data-check}
grade_this_code()
```

We have data, ready for analysis. 

## A slightly more complicated example

What if we want more than one year in our dataframe?

This is a common problem. What if we want to look at every scoring offense going back several years? The website has them going back to 2009. How can we combine them? 

First, we should note, that the data does not have anything in it to indicate what year it comes from. So we're going to have to add that. And we're going to have to figure out a way to stack two dataframes on top of each other. 

So let's grab 2021.

```{r rvest-load-data2-exercise, exercise = TRUE}
scoringoffenseurl21 <- "http://www.cfbstats.com/2021/leader/national/team/offense/split01/category09/sort01.html"

scoringoffense21 <- scoringoffenseurl21 %>%
  read_html() %>%
  html_nodes(xpath = '//*[@id="content"]/div[2]/table') %>%
  html_table()

scoringoffense2021 <- scoringoffense21[[1]]
```
```{r rvest-load-data2-exercise-solution}
scoringoffenseurl21 <- "http://www.cfbstats.com/2021/leader/national/team/offense/split01/category09/sort01.html"

scoringoffense21 <- scoringoffenseurl21 %>%
  read_html() %>%
  html_nodes(xpath = '//*[@id="content"]/div[2]/table') %>%
  html_table()

scoringoffense2021 <- scoringoffense21[[1]]
```
```{r rvest-load-data2-exercise-check}
grade_this_code()
```

First, how are we going to know, in the data, which year our data is from? We can use mutate. But we have a problem. Let's look at head again. Notice the first column -- it doesn't have a name. That's bad.

```{r head-data2, exercise=TRUE, exercise.setup = "rvest-load-data"}
head(scoringoffense2021)
```
```{r head-data2-solution}
head(scoringoffense2021)
```
```{r head-data2-check}
grade_this_code()
```

### Exercise 2: Fixing nameless columns

To fix this all of this, we're going to run two functions -- `rename` and `mutate`. The mutate is easy -- were going to mutate a new column that has the year the data comes from. Rename, to refresh your memory, is rename(NAME WE ARE GIVING THE COLUMN = WHICH COLUMN WE ARE RENAMING). But wait. Doesn't our first column not have a name? How are we going to ... you know ... name it? We can call it by it's position. In this case, it's in position 1. But we have to fix the name issue before we can fix the what-year-is-it problem. We'll also fix our 2021 data while we're at it.

```{r rvest3, exercise=TRUE, exercise.setup = "rvest-load-data", message=FALSE}
fixedscoringoffense2021 <- scoringoffense2021 %>% rename(Rank = 1) %>% mutate(YEAR = ????)
fixedscoringoffense2022 <- scoring %>% rename(Rank = 1) %>% mutate(YEAR = ????)
```
```{r rvest3-solution, exercise.reveal_solution = FALSE}
fixedscoringoffense2021 <- scoringoffense2021 %>% rename(Rank = 1)%>% mutate(YEAR = 2021)
fixedscoringoffense2022 <- scoring %>% rename(Rank = 1) %>% mutate(YEAR = 2022)
```
```{r rvest3-check}
grade_this_code()
```

And now, to combine the two tables together length-wise -- we need to make long data -- we'll use a dpylr function called `bind_rows`. The good thing is bind_rows is simple: It's just `bind_rows(dataframe, otherdataframe)` and we can keep putting dataframes in there until we run out. The catch is that they have to have the same columns for it to work. In our case, we have that.

```{r rvest4, exercise=TRUE, exercise.setup = "rvest-load-data", message=FALSE}
combined <- bind_rows(??????????????2021, ??????????????2022)
```
```{r rvest4-solution, exercise.reveal_solution = FALSE}
combined <- bind_rows(fixedscoringoffense2021, fixedscoringoffense2022)
```
```{r rvest4-check}
grade_this_code()
```

Note if we run glimpse on the dataframe called combined that it has 257 rows -- which just so happens to be what 127 from 2020 and 130 from 2021 add up to. 

```{r head-data3, exercise=TRUE, exercise.setup = "rvest-load-data"}
glimpse(combined)
```
```{r head-data3-solution}
glimpse(combined)
```
```{r head-data3-check}
grade_this_code()
```

## An even more complicated example

What do you do when the table has non-standard headers? 

Unfortunately, non-standard means there's no one way to do it -- it's going to depend on the table and the headers. But here's one idea: Don't try to make it work. 

I'll explain.

Let's try to get [college basketball season team stats from Sports Reference](https://www.sports-reference.com/cbb/seasons/2023-school-stats.html). If you look at that page, you'll see the problem right away -- the headers span two rows, and they repeat (i.e. there's conference wins, home wins and away wins, but because the conference, home and away markers are on a different line, R won't know which W is which). That's going to be all kinds of no good. **You can't import that.** Dataframes must have names all in one row. If you have two-line headers, you have a problem you have to fix before you can do anything else with it.

### Exercise 3: No headers please

First we'll grab the page. But this time, in our `html_table` bits, which we've just merrily been using untouched, we're going to add a switch: We're going to tell it that our table does not have headers by saying `header=FALSE`. 

```{r rvest5, exercise=TRUE, exercise.setup = "rvest-load-data", message=FALSE}
url <- "https://www.sports-reference.com/cbb/seasons/2023-school-stats.html"

stats <- url %>%
  read_html() %>%
  html_nodes(xpath = '//*[@id="basic_school_stats"]') %>%
  html_table(header=?????)
```
```{r rvest5-solution, exercise.reveal_solution = FALSE}
url <- "https://www.sports-reference.com/cbb/seasons/2023-school-stats.html"

stats <- url %>%
  read_html() %>%
  html_nodes(xpath = '//*[@id="basic_school_stats"]') %>%
  html_table(header=FALSE)
```
```{r rvest5-check}
grade_this_code()
```

And we'll take a look at what we have. 

```{r head-data4, exercise=TRUE, exercise.setup = "rvest-load-data"}
head(stats)
```
```{r head-data4-solution}
head(stats)
```
```{r head-data4-check}
grade_this_code()
```

So, that's not ideal. We have headers and data mixed together, and our columns are named X1 to X38. Also note: They're all character fields. Because the headers are interspersed with data, it all gets called character data. So we've got to first rename each field. 

Note, this is nothing special, it's just a lot of code and kind of tedious. I'm just going through each field and naming it what it is. While I'm at it, **I am removing any special characters, spaces or anything that would make R throw a fit.** So no dashes, percent signs, no columns that start with number.

```{r rvest6, exercise=TRUE, exercise.setup = "rvest-load-data", message=FALSE}
statsrenamed <- stats %>% 
  rename(
    Rank=X1, 
    School=X2, 
    Games=X3, 
    OverallWins=X4, 
    OverallLosses=X5, 
    WinPct=X6, 
    OverallSRS=X7, 
    OverallSOS=X8, 
    Blank1=X9, 
    ConferenceWins=X10, 
    ConferenceLosses=X11, 
    Blank2=X12, 
    HomeWins=X13, 
    HomeLosses=X14, 
    Blank3=X15, 
    AwayWins=X16, 
    AwayLosses=X17, 
    Blank4=X18, 
    ForPoints=X19, 
    OppPoints=X20, 
    Blank5=X21, 
    Minutes=X22, 
    FieldGoalsMade=X23, 
    FieldGoalsAttempted=X24, 
    FieldGoalPCT=X25, 
    ThreePointMade=X26, 
    ThreePointAttempts=X27, 
    ThreePointPct=X28, 
    FreeThrowsMade=X29, 
    FreeThrowsAttempted=X30, 
    FreeThrowPCT=X31, 
    OffensiveRebounds=X32, 
    TotalRebounds=X33, 
    Assists=X34, 
    Steals=X35, 
    Blocks=X36, 
    Turnovers=X37, 
    PersonalFouls=X38)
```
```{r rvest6-solution, exercise.reveal_solution = FALSE}
statsrenamed <- stats %>% 
  rename(
    Rank=X1, 
    School=X2, 
    Games=X3, 
    OverallWins=X4, 
    OverallLosses=X5, 
    WinPct=X6, 
    OverallSRS=X7, 
    OverallSOS=X8, 
    Blank1=X9, 
    ConferenceWins=X10, 
    ConferenceLosses=X11, 
    Blank2=X12, 
    HomeWins=X13, 
    HomeLosses=X14, 
    Blank3=X15, 
    AwayWins=X16, 
    AwayLosses=X17, 
    Blank4=X18, 
    ForPoints=X19, 
    OppPoints=X20, 
    Blank5=X21, 
    Minutes=X22, 
    FieldGoalsMade=X23, 
    FieldGoalsAttempted=X24, 
    FieldGoalPCT=X25, 
    ThreePointMade=X26, 
    ThreePointAttempts=X27, 
    ThreePointPct=X28, 
    FreeThrowsMade=X29, 
    FreeThrowsAttempted=X30, 
    FreeThrowPCT=X31, 
    OffensiveRebounds=X32, 
    TotalRebounds=X33, 
    Assists=X34, 
    Steals=X35, 
    Blocks=X36, 
    Turnovers=X37, 
    PersonalFouls=X38)
```
```{r rvest6-check}
grade_this_code()
```

Now we have to get rid of those headers interspersed in the data. We can do this with a filter, but we need a pattern to get rid of them. Let's pull up the head of the data and see if we can spot one.

```{r head-data5, exercise=TRUE, exercise.setup = "rvest-load-data"}
head(statsrenamed)
```
```{r head-data5-solution}
head(statsrenamed)
```
```{r head-data5-check}
grade_this_code()
```

From that, I can see that in my Rank column, I have a row that has "Rk" in it. I don't want that. Then, in my Games column, along with several others, I have "Overall" in a row. Don't want that either. With filters so far, we've been using mostly == and greater than/less than. The opposite of == is !=, or does not equal. In other words, keep everything EXCEPT this.

```{r rvest7, exercise=TRUE, exercise.setup = "rvest-load-data", message=FALSE}
statswithoutgarbage <- statsrenamed %>% 
  filter(Rank != "Rk" & Games != "Overall") 
```
```{r rvest7-solution, exercise.reveal_solution = FALSE}
statswithoutgarbage <- statsrenamed %>% 
  filter(Rank != "Rk" & Games != "Overall") 
```
```{r rvest7-check}
grade_this_code()
```

And finally, we need to change the data type of all the fields that need it. We're going to use a clever little trick. We're going to use `mutate_at`, which means mutate these fields. The pattern for `mutate_at` is `mutate_at` these variables and do this thing to them. But instead of specifying which of 38 variables we're going to mutate, we're going to specify the one we don't want to change, which is just the name of the school. And we want to convert them to numeric, which is simple. 

One last thing: Who needs columns called Blank1, Blank2, Blank3, etc? 

Here's what it looks like:

```{r rvest8, exercise=TRUE, exercise.setup = "rvest-load-data", message=FALSE}
statswithnumbers <- statswithoutgarbage %>% 
  mutate_at(vars(-School), as.numeric) %>% 
  select(-starts_with("Blank"))
```
```{r rvest8-solution, exercise.reveal_solution = FALSE}
statswithnumbers <- statswithoutgarbage %>% 
  mutate_at(vars(-School), as.numeric) %>% 
  select(-starts_with("Blank"))
```
```{r rvest8-check}
grade_this_code()
```

And just like that (cough cough), we have a method for getting up to the minute season stats for every team in Division I. 

```{r head-data6, exercise=TRUE, exercise.setup = "rvest-load-data"}
head(statswithnumbers)
```
```{r head-data6-solution}
head(statswithnumbers)
```
```{r head-data6-check}
grade_this_code()
```

While this seems like a lot of work, if your job is to analyze college basketball as the year goes by, and you're going to be pulling this page every few days, the work you just did will save you hours of time down the road. Automating the import of data is an important function for data scientists to save time on drudgery and focus on analysis.
