---
title: "Sports Data Lesson 22: Bump charts"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to make a combination of step charts and subway maps for ranks.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(ggbump)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

The goal of this lesson is to introduce you to bump charts, a powerful visualization technique for displaying changes in rankings over time. By the end of this tutorial, you'll understand when to use bump charts, how to create them using the ggbump package in R, and how to customize them for clearer data representation. You'll learn to prepare data for bump charts, create basic plots, add points and labels to enhance readability, and fine-tune the appearance with custom colors and themes. We'll apply these concepts to real college football playoff ranking data, visualizing how team rankings change across multiple weeks. This lesson will enhance your ability to effectively communicate ranking changes and trends in sports data, a crucial skill for analyzing and presenting competitive standings over a season or multiple time periods.

## The Basics

The point of a bump chart is to show how the ranking of something changed over time -- you could do this with the top 25 in football or basketball. I've seen it done with European soccer league standings over a season. 

The requirements are that you have a row of data for a team, in that week, with their rank.

This is another extension to ggplot. You installed it at the beginning of the semester, but if you need it, you'll install it the usual way: `install.packages("ggbump")`

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
library(ggbump)
```
```{r load-tidyverse-solution}
library(tidyverse)
library(ggbump)
```
```{r load-tidyverse-check}
grade_this_code()
```

Let's use college football playoff rankings, specifically the top 10 in each week of the rankings. Per usual: **For purposes of this exercise, you don't need to do this. The data is included here if you want to try this in your own notebook.** 

```{r echo=FALSE, class.output="bg-info", results="asis",  message=FALSE,  warning=FALSE}
library(downloadthis)
library(glue)

dllink <- download_link(
  link = "https://mattwaite.github.io/sportsdatafiles/cfbranking23.csv",
  button_label = "Download csv file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

glue("<pre><p><strong>For this walkthrough:</strong></p><p>{dllink}</p></pre>")
```

Now let's load it.

```{r bump-load-data, message=FALSE, warning=FALSE}
rankings <- read_csv("https://mattwaite.github.io/sportsdatafiles/cfbranking23.csv")
```
```{r bump-load-data-exercise, exercise = TRUE}
rankings <- read_csv("https://mattwaite.github.io/sportsdatafiles/cfbranking23.csv")
```
```{r bump-load-data-exercise-solution}
rankings <- read_csv("https://mattwaite.github.io/sportsdatafiles/cfbranking23.csv")
```
```{r bump-load-data-exercise-check}
grade_this_code()
```

Given our requirements of a row of data for a team, in that week, with their rank, take a look at the data provided. We have 6 weeks of playoff rankings, so we should see a ranking, the week of the ranking and the team at that rank. You can see the basic look of the data by using head()

```{r head-data, exercise=TRUE, exercise.setup = "bump-load-data"}
head(rankings)
```
```{r head-data-solution}
head(rankings)
```
```{r head-data-check}
grade_this_code()
```

Our data is in the form we need it to be. Now we can make a bump chart. 

### Exercise 1: 

We'll start simple. `ggbump` adds it's own geom, `geom_bump`, and the parts of it are what you think. x is the time element -- Week in our case -- and the Y is the rank. Last, we'll add a color inside the aesthetic. Given that our data only has three columns, that narrows down what the color should be. 

```{r bump1, exercise=TRUE, exercise.setup = "bump-load-data", message=FALSE}
ggplot() + geom_bump(data=rankings, aes(x=____, y=____, color=____))
```
```{r bump1-solution, exercise.reveal_solution = FALSE}
ggplot() + geom_bump(data=rankings, aes(x=Week, y=Rank, color=Team))
```
```{r bump1-check}
grade_this_code()
```

Well, it's a start. 

First things first: I'm immediately annoyed by the top teams being at the bottom. 

### Exercise 2: Reversing one of the axes.

I learned a neat trick from ggbump that's been in ggplot all along -- `scale_y_reverse()`

```{r bump2, exercise=TRUE, exercise.setup = "bump-load-data", message=FALSE}
ggplot() + 
  geom_bump(data=rankings, aes(x=____, y=____, color=____)) +
  scale_y_reverse()
```
```{r bump2-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bump(data=rankings, aes(x=Week, y=Rank, color=Team)) +
  scale_y_reverse()
```
```{r bump2-check}
grade_this_code()
```

Better. But, still not great. 

### Exercise 3: Adding a point for each week

Let's add a point at each week. All along to this point, we've been laying in geoms but mostly they're the same geom. There's no reason they can't be different, though. To make this happen, we add a `geom_point` with the exact same aesthetic.

```{r bump3, exercise=TRUE, exercise.setup = "bump-load-data", message=FALSE}
ggplot() + 
  geom_bump(data=rankings, aes(x=____, y=____, color=____)) +
  geom_point(data=rankings, aes(x=____, y=____, color=____), size = 4) +
  scale_y_reverse() 
```
```{r bump3-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bump(data=rankings, aes(x=Week, y=Rank, color=Team)) + 
  geom_point(data=rankings, aes(x=Week, y=Rank, color=Team), size = 4) +
  scale_y_reverse() 
```
```{r bump3-check}
grade_this_code()
```

## Fine tuning

We've taken another step closer. That makes it more subway-map like. But the colors are all wrong. To fix this, we're going to use `scale_color_manual` and we're going to Google the hex codes for each team. The legend will tell you what order your `scale_color_manual` needs to be. You just need to have an entry for every team in the legend. Googling "Alabama hex code" will give you the info you need to copy and paste into `scale_color_manual`. 

```{r bump4, exercise=TRUE, exercise.setup = "bump-load-data", message=FALSE}
ggplot() + 
  geom_bump(data=rankings, aes(x=Week, y=Rank, color=Team)) + 
  geom_point(data=rankings, aes(x=Week, y=Rank, color=Team), size = 4) + 
  scale_color_manual(values = c("#690014", "#782F40", "#BA0C2F", "#AD0000", "#00274C", "#14213D", "#000000", "#BB0000", "#841617", "#154733", "#041E42", "#BF5700", "#4B2E83")) +
  scale_y_reverse() 
```
```{r bump4-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bump(data=rankings, aes(x=Week, y=Rank, color=Team)) + 
  geom_point(data=rankings, aes(x=Week, y=Rank, color=Team), size = 4) + 
  scale_color_manual(values = c("#690014", "#782F40", "#BA0C2F", "#AD0000", "#00274C", "#14213D", "#000000", "#BB0000", "#841617", "#154733", "#041E42", "#BF5700", "#4B2E83")) +
  scale_y_reverse() 
```
```{r bump4-check}
grade_this_code()
```

Another step. But the legend is annoying. And trying to find which red is Alabama vs Ohio State is hard. So what if we labeled each dot at the beginning **and** end? We can do that with some clever usage of geom_text and a little dplyr filtering inside the data step. We filter out the first and last weeks, then use hjust -- horizontal justification -- to move them left or right. NOTE: A good practice is to put all your geoms together, all your scale work together, an your theme work last. 

```{r bump5, exercise=TRUE, exercise.setup = "bump-load-data", message=FALSE}
ggplot() + 
  geom_bump(data=rankings, aes(x=Week, y=Rank, color=Team)) + 
  geom_point(data=rankings, aes(x=Week, y=Rank, color=Team), size = 4) + 
  geom_text(data = rankings |> filter(Week == min(Week)), aes(x = Week - .2, y=Rank, label = Team), size = 3, hjust = 1) +
  geom_text(data = rankings |> filter(Week == max(Week)), aes(x = Week + .2, y=Rank, label = Team), size = 3, hjust = 0) +
  scale_color_manual(values = c("#690014", "#782F40", "#BA0C2F", "#AD0000", "#00274C", "#14213D", "#000000", "#BB0000", "#841617", "#154733", "#041E42", "#BF5700", "#4B2E83")) +
  scale_y_reverse() 
```
```{r bump5-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bump(data=rankings, aes(x=Week, y=Rank, color=Team)) + 
  geom_point(data=rankings, aes(x=Week, y=Rank, color=Team), size = 4) + 
  geom_text(data = rankings |> filter(Week == min(Week)), aes(x = Week - .2, y=Rank, label = Team), size = 3, hjust = 1) +
  geom_text(data = rankings |> filter(Week == max(Week)), aes(x = Week + .2, y=Rank, label = Team), size = 3, hjust = 0) +
  scale_color_manual(values = c("#690014", "#782F40", "#BA0C2F", "#AD0000", "#00274C", "#14213D", "#000000", "#BB0000", "#841617", "#154733", "#041E42", "#BF5700", "#4B2E83")) +
  scale_y_reverse() 
```
```{r bump5-check}
grade_this_code()
```

Better, but the legend is still there. We can drop it in a theme directive by saying `legend.position = "none"`. We'll also throw a theme_minimal on there to drop the default grey, and we'll add some better labeling.

```{r bump6, exercise=TRUE, exercise.setup = "bump-load-data", message=FALSE}
ggplot() + 
  geom_bump(data=rankings, aes(x=Week, y=Rank, color=Team)) + 
  geom_point(data=rankings, aes(x=Week, y=Rank, color=Team), size = 4) + 
  geom_text(data = rankings |> filter(Week == min(Week)), aes(x = Week - .2, y=Rank, label = Team), size = 3, hjust = 1) +
  geom_text(data = rankings |> filter(Week == max(Week)), aes(x = Week + .2, y=Rank, label = Team), size = 3, hjust = 0) +
  scale_color_manual(values = c("#690014", "#782F40", "#BA0C2F", "#AD0000", "#00274C", "#14213D", "#000000", "#BB0000", "#841617", "#154733", "#041E42", "#BF5700", "#4B2E83")) +
  scale_y_reverse() +
  labs(title="A playoff rankings twist", subtitle="The CFB playoff picture changed more in the last week than it did in the previous four weeks.", y= "Rank", x = "Week") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank()
    ) 
```
```{r bump6-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bump(data=rankings, aes(x=Week, y=Rank, color=Team)) + 
  geom_point(data=rankings, aes(x=Week, y=Rank, color=Team), size = 4) + 
  geom_text(data = rankings |> filter(Week == min(Week)), aes(x = Week - .2, y=Rank, label = Team), size = 3, hjust = 1) +
  geom_text(data = rankings |> filter(Week == max(Week)), aes(x = Week + .2, y=Rank, label = Team), size = 3, hjust = 0) +
  scale_color_manual(values = c("#690014", "#782F40", "#BA0C2F", "#AD0000", "#00274C", "#14213D", "#000000", "#BB0000", "#841617", "#154733", "#041E42", "#BF5700", "#4B2E83")) +
  scale_y_reverse() +
  labs(title="A playoff rankings twist", subtitle="The CFB playoff picture changed more in the last week than it did in the previous four weeks.", y= "Rank", x = "Week") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank()
    ) 
```
```{r bump6-check}
grade_this_code()
```

And the last thing: anyone else annoyed at 7.5th place on the left? We can fix that too by specifying the breaks in scale_y_reverse. We can do that with the x axis as well, but since we haven't reversed it, we do that in `scale_x_continuous` with the same breaks. 

```{r bump7, exercise=TRUE, exercise.setup = "bump-load-data", message=FALSE}
ggplot() + 
  geom_bump(data=rankings, aes(x=Week, y=Rank, color=Team)) + 
  geom_point(data=rankings, aes(x=Week, y=Rank, color=Team), size = 4) + 
  geom_text(data = rankings |> filter(Week == min(Week)), aes(x = Week - .2, y=Rank, label = Team), size = 3, hjust = 1) +
  geom_text(data = rankings |> filter(Week == max(Week)), aes(x = Week + .2, y=Rank, label = Team), size = 3, hjust = 0) +
  scale_color_manual(values = c("#690014", "#782F40", "#BA0C2F", "#AD0000", "#00274C", "#14213D", "#000000", "#BB0000", "#841617", "#154733", "#041E42", "#BF5700", "#4B2E83")) +
  scale_x_continuous(breaks=c(9,10,11,12,13,14), limits = c(8, 15)) + 
  scale_y_reverse(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15)) +
  labs(title="A playoff rankings twist", subtitle="The CFB playoff picture changed more in the last week than it did in the previous four weeks.", y= "Rank", x = "Week") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank()
    ) 
```
```{r bump7-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bump(data=rankings, aes(x=Week, y=Rank, color=Team)) + 
  geom_point(data=rankings, aes(x=Week, y=Rank, color=Team), size = 4) + 
  geom_text(data = rankings |> filter(Week == min(Week)), aes(x = Week - .2, y=Rank, label = Team), size = 3, hjust = 1) +
  geom_text(data = rankings |> filter(Week == max(Week)), aes(x = Week + .2, y=Rank, label = Team), size = 3, hjust = 0) +
  scale_color_manual(values = c("#690014", "#782F40", "#BA0C2F", "#AD0000", "#00274C", "#14213D", "#000000", "#BB0000", "#841617", "#154733", "#041E42", "#BF5700", "#4B2E83")) +
  scale_x_continuous(breaks=c(9,10,11,12,13,14), limits = c(8, 15)) + 
  scale_y_reverse(breaks=c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15)) +
  labs(title="A playoff rankings twist", subtitle="The CFB playoff picture changed more in the last week than it did in the previous four weeks.", y= "Rank", x = "Week") +
  theme_minimal() +
  theme(
    legend.position = "none",
    panel.grid.major = element_blank()
    ) 
```
```{r bump7-check}
grade_this_code()
```
There's still some work to do here -- note the names on the edges are getting cut off. But compare to where we started. This is looking pretty good. 

## The Recap

In this lesson, we've explored the power of bump charts for visualizing changes in rankings over time in sports data. We started by learning how to create basic bump charts using the ggbump package, and then progressively enhanced our visualization through several iterations. We covered important techniques such as reversing axes to place higher ranks at the top, adding points to highlight specific time periods, and incorporating team-specific colors. We also learned how to add labels at the start and end points for better readability, remove unnecessary elements like legends, and fine-tune axis breaks for a cleaner look. Throughout the process, we saw how combining different ggplot2 elements like geom_bump, geom_point, and geom_text can create a more informative and visually appealing chart. We also touched on the importance of thoughtful design choices, such as using team colors and minimizing chart junk. Remember, while bump charts are excellent for showing ranking changes, the effectiveness of your visualization often comes down to these final details and refinements. Always consider your audience and the story you're trying to tell when making these design decisions.

## Terms to Know

- Bump Chart: A type of chart that visualizes changes in rankings over time, often resembling a subway map.
- `ggbump`: An R package that extends ggplot2 to create bump charts and related visualizations.
- `geom_bump()`: A function from the ggbump package used to create the main lines in a bump chart.
- `scale_y_reverse()`: A ggplot2 function used to reverse the y-axis, often used in bump charts to place higher ranks at the top.
- `geom_point()`: A ggplot2 function used to add points to a chart, in this case highlighting specific time periods in the bump chart.
- `scale_color_manual()`: A ggplot2 function used to manually specify colors for different categories, such as team colors in this example.
- `hjust`: Short for "horizontal justification", a parameter in text geoms that controls the horizontal alignment of text labels.
- `legend.position`: A theme parameter in ggplot2 used to control the position of the legend, or remove it entirely.
- `scale_x_continuous()`: A ggplot2 function used to customize the x-axis, including setting specific breaks and limits.