---
title: "Lesson 12: Bar charts"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to start turning data into graphics.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(scales)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

# Intro to ggplot

## The basics

With `ggplot2`, we dive into the world of programmatic data visualization. The `ggplot2` library implements something called the grammar of graphics. The main concepts are: 

* aesthetics - which in this case means the data which we are going to plot
* geometries - which means the shape the data is going to take
* scales - which means any transformations we might make on the data
* facets - which means how we might graph many elements of the same dataset in the same space
* layers - which means how we might lay multiple geometries over top of each other to reveal new information.

Hadley Wickham, who is behind all of the libraries we have used in this course to date, wrote about his layered grammar of graphics in [this 2009 paper that is worth your time to read](http://byrneslab.net/classes/biol607/readings/wickham_layered-grammar.pdf). 

Here are some `ggplot2` resources you'll want to keep handy: 

* [The ggplot documentation](http://ggplot2.tidyverse.org/reference/index.html)
* [The ggplot cookbook](http://www.cookbook-r.com/Graphs/)

Let's dive in using data on college football since 2018. Why 2018? Uh, no local reason. This workflow will represent a clear picture of what your work in this class will be like for much of the rest of the semester. One way to think of this workflow is that your R Notebook is now your digital sketchbook, where you will try different types of visualizations to find ones that work. Then, you will write the code that adds necessary and required parts to finish it. 

**For purposes of this exercise, you don't need to download this. The data is included here if you want to try this in your own notebook.**

```{r echo=FALSE, class.output="bg-info", results="asis",  message=FALSE,  warning=FALSE}
library(downloadthis)
library(glue)

dllink <- download_link(
  link = "http://mattwaite.github.io/sportsdatafiles/cfb1822.csv",
  button_label = "Download csv file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

glue("<pre><p><strong>For this walkthrough:</strong></p><p>{dllink}</p></pre>")

```

Now load the tidyverse.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
```
```{r load-tidyverse-solution}
library(tidyverse)
```
```{r load-tidyverse-check}
grade_this_code()
```

And the data.

```{r ggplot-load-data, message=FALSE, warning=FALSE}
seasons <- read_csv("http://mattwaite.github.io/sportsdatafiles/cfb1822.csv")

huskers <- seasons %>%
  mutate(
    yardspergame = OffensiveYards/Games
  ) %>% 
  filter(
    Team == "Nebraska"
  )

big10 <- seasons %>%
  mutate(
    yardspergame = OffensiveYards/Games
  ) %>% 
  filter(
    Conference == "Big Ten Conference" &
      Season == "2022"
  )
```
```{r ggplot-load-data-exercise, exercise = TRUE}
seasons <- read_csv("http://mattwaite.github.io/sportsdatafiles/cfb1822.csv")
```
```{r ggplot-load-data-exercise-solution}
seasons <- read_csv("http://mattwaite.github.io/sportsdatafiles/cfb1822.csv")
```
```{r ggplot-load-data-exercise-check}
grade_this_code()
```

And let's get a glimpse of the data to remind us what is all there. 

Let's `glimpse` the data because it's wide and has a lot of columns. 

```{r glimpse-data, exercise=TRUE, exercise.setup = "ggplot-load-data"}
glimpse(seasons)
```
```{r glimpse-data-solution}
glimpse(seasons)
```
```{r glimpse-data-check}
grade_this_code()
```

Now that we're set up, when making graphics, you're rarely ever going straight from data to graphic. Normally, you'll have way too much data to visualize, or you need to do something to that data in order to visualize it. In our case, we have way too many schools in our data to make a reasonable bar chart. Our first challenge, then, is to narrow the pile. 

### Exercise 1: Set up your data. 

To do this, we're going to do a few things we've done before -- we're going to mutate a new column calculating total offensive yards per game so we can compare season to season (including the COVID season) and then we'll use filter to get just Nebraska. We'll put that into a new dataframe called `huskers`.

```{r ggplot-top, exercise=TRUE, exercise.setup = "ggplot-load-data", message=FALSE}
??????? <- ??????? %>%
  mutate(
    yardspergame = ????????????/?????
  ) %>% 
  filter(
    ???? == "Nebraska"
  ) 
```
```{r ggplot-top-solution, exercise.reveal_solution = FALSE}
huskers <- seasons %>%
  mutate(
    yardspergame = OffensiveYards/Games
  ) %>% 
  filter(
    Team == "Nebraska"
  )
```
```{r ggplot-top-check}
grade_this_code()
```

And we can check that it all worked by doing glimpse again on on our new dataframe.

```{r glimpse2-data, exercise=TRUE, exercise.setup = "ggplot-load-data"}
glimpse(?????)
```
```{r glimpse2-data-solution}
glimpse(huskers)
```
```{r glimpse2-data-check}
grade_this_code()
```

## The bar chart

The easiest thing we can do is create a simple bar chart of our data. **Bar charts show magnitude. They invite you to compare how much more or less one thing is compared to others.**

We could, for instance, create a bar chart of the yards per game. We might be asking the question: Which Scott Frost offense was the most effective? To do that, we simply tell `ggplot2` what our dataset is, what element of the data we want to make the bar chart out of (which is the aesthetic), and the geometry type (which is the geom). It looks like this:

`ggplot() + geom_bar(data=huskers, aes(x=Season))` 

We start with `ggplot()` which is creating a blank canvas for us to work in. The `geom_bar()` is the geometry -- the form the data will take. We will learn many geometries over the next several lessons. `huskers` is our data, `aes` means aesthetics, `x=Season` explicitly tells `ggplot2` that our x value -- our horizontal value -- is the Season field from the data. Why season? We want one bar per season, no? Put the season on the x axis is what that is saying. And what do we get when we run that? 

```{r ggplot-bar1, exercise=TRUE, exercise.setup = "ggplot-load-data", message=FALSE}
ggplot() + geom_bar(data=huskers, aes(x=Season))
```
```{r ggplot-bar1-solution, exercise.reveal_solution = FALSE}
ggplot() + geom_bar(data=huskers, aes(x=Season))
```
```{r ggplot-bar1-check}
grade_this_code()
```

We get ... weirdness. We expected to see bars of different sizes, but we get all with a count of 1. What gives? Well, this is the default behavior. What we have here is something called a histogram, where `ggplot2` helpfully counted up the number of times the Season appears and made bars as long as the count. Since we only have one record per Season, the count is always 1. How do we fix this? By adding `weight` to our aesthetic. 

### Exercise 2: The weight of responsibility

You saw how it was done before. This just adds weight to the aes. The weight is always going to be a number, and our number is the `digsperset`.

```{r ggplot-bar2, exercise=TRUE, exercise.setup = "ggplot-load-data", message=FALSE}
ggplot() + 
  geom_bar(data=huskers, aes(x=Season, weight=??????????))
```
```{r ggplot-bar2-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bar(data=huskers, aes(x=Season, weight=yardspergame))
```
```{r ggplot-bar2-check}
grade_this_code()
```

Now we get bars of different sizes, and you can clearly see a difference between seasons. 

Let's work on a different version of this to see if we can find a clearer story.

### Exercise 3: New data

`ggplot2`'s default behavior is to sort the data by the x axis variable. It's in alphabetical order -- that's ... weirdly ... why our seasons were in the correct order. 

But what happens when that isn't quite the case. 

To change the order of the bars, we have to `reorder` it. With `reorder`, we first have to tell `ggplot` what we are reordering, and then we have to tell it HOW we are reordering it. So it's reorder(FIELD, SORTFIELD). 

First, let's get a dataset of every Big Ten team in 2022 instead of just Nebraska. 

```{r ggplot-bar3, exercise=TRUE, exercise.setup = "ggplot-load-data", message=FALSE}
big10 <- seasons %>%
  mutate(
    yardspergame = ????????????/Games
  ) %>% 
  filter(
    Conference == "??? ??? ?????????" &
      Season == "????"
  )
```
```{r ggplot-bar3-solution, exercise.reveal_solution = FALSE}
big10 <- seasons %>%
  mutate(
    yardspergame = OffensiveYards/Games
  ) %>% 
  filter(
    Conference == "Big Ten Conference" &
      Season == "2022"
  )
```
```{r ggplot-bar3-check}
grade_this_code()
```

There, now we can use reorder.

### Exercise 4: Reordering

Our new data has 14 teams in it and only one season, so we can no longer use Season as our x axis label. And we want those institutions not in alphabetical order, but in order of each team's yardspergame 

```{r ggplot-bar4, exercise=TRUE, exercise.setup = "ggplot-load-data", message=FALSE}
ggplot() + 
  geom_bar(data=big10, aes(x=reorder(????, ???????????), weight=???????????))
```
```{r ggplot-bar4-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bar(data=big10, aes(x=reorder(Team, yardspergame), weight=yardspergame)) 
```
```{r ggplot-bar4-check}
grade_this_code()
```

Better. And look who is dead last in the conference in offensive yards. Good. Rot there. 

## One last trick: coord flip

Sometimes, we don't want vertical bars. Maybe we think this would look better horizontal. Maybe our x-axis labels will never fit on the x-axis. How do we do that? By adding `coord_flip()` to our code. It does what it says -- it inverts the coordinates of the figures.

### Exercise 5: Flip the coordinates

```{r ggplot-bar5, exercise=TRUE, exercise.setup = "ggplot-load-data", message=FALSE}
ggplot() + 
  geom_bar(data=big10, aes(x=reorder(Team, ??????????), weight=??????????)) +
  ?????_????()
```
```{r ggplot-bar5-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_bar(data=big10, aes(x=reorder(Team, yardspergame), weight=yardspergame)) +
  coord_flip()
```
```{r ggplot-bar5-check}
grade_this_code()
```

Now we've got something working. There's a lot of work left to do to make this publishable, but we'll get to that in due time.