---
title: "Sports Data Lesson 16: Line charts"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to show change over time.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

The goal of this lesson is to introduce you to line charts, a powerful visualization tool for displaying trends over time. By the end of this tutorial, you'll understand when to use line charts, how to create them using ggplot2, and how to customize them for clearer data representation. You'll learn to plot single and multiple lines, adjust axes for better context, and add layers to compare different datasets. We'll apply these concepts to real college basketball data, analyzing team performance across a season. This lesson will enhance your ability to visualize and interpret time-series data, a crucial skill in sports analytics for understanding performance trends, seasonal patterns, and comparative team dynamics.

## The Basics

So far, we've talked about bar charts -- stacked or otherwise -- are good for showing relative size of a thing compared to another thing. Stacked Bars and Waffle charts are good at showing proportions of a whole. 

**Line charts are good for showing change over time.** 

Let's look at how we can answer this question: How is this season's Nebraska's basketball season progressing?

We'll use the logs of every game in college basketball. 

Let's start getting all that we need. We can use the tidyverse shortcut. 

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
```
```{r load-tidyverse-solution}
library(tidyverse)
```
```{r load-tidyverse-check}
grade_this_code()
```

And the data.

```{r line-load-data, message=FALSE, warning=FALSE}
logs <- read_csv("https://mattwaite.github.io/sportsdatafiles/logs26.csv")

nu <- logs |> filter(team == "Nebraska Cornhuskers")
pu <- logs |> filter(team == "Purdue Boilermakers")
big <- logs |> filter(conference == "Big Ten")
average <- logs |> group_by(game_date) |> summarise(mean_shooting=mean(team_fgpct))
```
```{r line-load-data-exercise, exercise = TRUE}
logs <- read_csv("https://mattwaite.github.io/sportsdatafiles/logs26.csv")
```
```{r line-load-data-exercise-solution}
logs <- read_csv("https://mattwaite.github.io/sportsdatafiles/logs26.csv")
```
```{r line-load-data-exercise-check}
grade_this_code()
```

### Exercise 1: Prepare your data

This data has every game from every team in it, so we need to use filtering to limit it, because we just want to look at Nebraska. If you don't remember, flip back to lesson 5. Let's make a new dataframe called nu for Nebraska's games. In this data, the team name is the whole name -- Nebraska Cornhuskers. 

```{r nu, exercise=TRUE, exercise.setup = "line-load-data", message=FALSE}
____ <- logs |> filter(team == "____ ____")
```
```{r nu-solution, exercise.reveal_solution = FALSE}
nu <- logs |> filter(team == "Nebraska Cornhuskers")
```
```{r nu-check}
grade_this_code()
```

We can get a peek at it to make sure it all worked.

```{r head-data, exercise=TRUE, exercise.setup = "line-load-data"}
head(nu)
```
```{r head-data-solution}
head(nu)
```
```{r head-data-check}
grade_this_code()
```

Because this data has just Nebraska data in it, the dates are formatted correctly, and the data is long data (instead of wide), we have what we need to make line charts.

### Exercise 2: Your first line chart

Line charts, unlike bar charts, do have a y-axis. So in our ggplot step, we have to define what our x and y axes are. In this case, the x axis is our `game_date` -- *the most common x axis in line charts is going to be a date of some variety* -- and y in this case is up to us. How well a team shoots the ball has a lot to do with how well a team does in a season, so let's chart that using the team field goal percentage. 

```{r line1, exercise=TRUE, exercise.setup = "line-load-data", message=FALSE}
ggplot() + geom_line(data=____, aes(x=____, y=____fgpct))
```
```{r line1-solution, exercise.reveal_solution = FALSE}
ggplot() + geom_line(data=nu, aes(x=game_date, y=team_fgpct))
```
```{r line1-check}
grade_this_code()
```

See some problems here? Whatever you came up with, the real problem is that the Y axis doesn't start with zero. 

### Exercise 3: Zeroing your axes

Using a non-zero axis makes a line chart look more dramatic than it is. To make the axis what you want, you can use `scale_x_continuous` or `scale_y_continuous` and pass in a list with the bottom and top value you want. The bottom value is zero. The upper value is up to you. Let'ss try .7 or 70 percent.

```{r line2, exercise=TRUE, exercise.setup = "line-load-data", message=FALSE}
ggplot() + 
  geom_line(data=____, aes(x=____, y=____fgpct)) +
  scale_y_continuous(limits = c(0, .7))
```
```{r line2-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_line(data=nu, aes(x=game_date, y=team_fgpct)) +
  scale_y_continuous(limits = c(0, .7))
```
```{r line2-check}
grade_this_code()
```

Note also that our X axis labels are automated. It knows it's a date and it just labels it by month.

## One line line charts are bad 

With datasets, we want to invite comparison. So let's answer the question visually. Let's put two lines on the same chart. How does Nebraska compare to Purdue, for example? 

### Exercise 4: Adding another line

When you have multiple datasets from the same data, you can add multiple geoms. All you have to do is swap out the data and add a + between them. 

```{r line3, exercise=TRUE, exercise.setup = "line-load-data", message=FALSE}
pu <- logs |> filter(team == "____ ____")

ggplot() + 
  geom_line(data=____, aes(x=game_date, y=team_fgpct), color="red") + 
  geom_line(data=____, aes(x=game_date, y=team_fgpct), color="gold") +
  scale_y_continuous(limits = c(0, .7))
```
```{r line3-solution, exercise.reveal_solution = FALSE}
pu <- logs |> filter(team == "Purdue Boilermakers")

ggplot() + 
  geom_line(data=nu, aes(x=game_date, y=team_fgpct), color="red") + 
  geom_line(data=pu, aes(x=game_date, y=team_fgpct), color="gold") +
  scale_y_continuous(limits = c(0, .7))
```
```{r line3-check}
grade_this_code()
```

When you're doing this on your own in a notebook, REMEMBER COPY AND PASTE IS A THING. Nothing changes except what data you are using and maybe the color. The less typing you're doing, the better.

So visually speaking, what is the difference between Nebraska and Purdue's season? Does it tell a story?

## A lot of lines = context. 

There are some line charts where a small number are good. A good way to add more context? Add more lines. But unlike what we're doing above, some lines don't need attention. I'll explain.

Let's add all of the Big Ten to our line chart, but this time, we're going to fade them into the background by making them grey. Then, our red line for Nebraska and the gold line of Purdue will really stand out.

### Exercise 5: Adding the B1G

An important thing to know is that geoms render in the order that they appear. The first one is on the bottom. The next one is the second layer. The third layer of three is on top. If you put the Big Ten layer last, the lines will cover everything in grey. If you put it first, then the Nebraska and Purdue's lines will be the only ones covered ... by Nebraska and Purdue.

NOTE: When you have more than one thing in a dataframe -- teams, players, whatever -- and you're making a line chart out of it, you need to add the `group` to the aesthetic. In this case, the group is the team.

```{r line4, exercise=TRUE, exercise.setup = "line-load-data", message=FALSE}
big <- logs |> filter(conference == "Big Ten")

ggplot() + 
  geom_line(data=____, aes(x=game_date, y=team_fgpct, group=team), color="____") +
  geom_line(data=____, aes(x=game_date, y=team_fgpct), color="red") + 
  geom_line(data=____, aes(x=game_date, y=team_fgpct), color="gold") +
  scale_y_continuous(limits = c(0, .7))
```
```{r line4-solution, exercise.reveal_solution = FALSE}
big <- logs |> filter(conference == "Big Ten")

ggplot() + 
  geom_line(data=big, aes(x=game_date, y=team_fgpct, group=team), color="grey") +
  geom_line(data=nu, aes(x=game_date, y=team_fgpct), color="red") + 
  geom_line(data=pu, aes(x=game_date, y=team_fgpct), color="gold") +
  scale_y_continuous(limits = c(0, .7))
```
```{r line4-check}
grade_this_code()
```

What do we see here? How has Nebraska and Purdue's season evolved against all the rest of the teams in the Big Ten?

### Exercise 6: Adding the average

But how does that compare to the average? We can add that pretty easily by creating a new dataframe with it and add another geom_line. 

```{r average, exercise=TRUE, exercise.setup = "line-load-data"}
average <- logs |> group_by(game_date) |> summarise(mean_shooting=mean(team_fgpct))
```
```{r average-solution}
average <- logs |> group_by(game_date) |> summarise(mean_shooting=mean(team_fgpct))
```
```{r average-check}
grade_this_code()
```

And now we have another geom_line to add. 

```{r line5, exercise=TRUE, exercise.setup = "line-load-data", message=FALSE}
big <- logs |> filter(conference == "Big Ten")

ggplot() + 
  geom_line(data=____, aes(x=game_date, y=team_fgpct, group=team), color="____") +
  geom_line(data=____, aes(x=game_date, y=team_fgpct), color="red") + 
  geom_line(data=____, aes(x=game_date, y=team_fgpct), color="gold") +
  geom_line(data=____, aes(x=game_date, y=mean_shooting), color="black") + 
  scale_y_continuous(limits = c(0, .7))
```
```{r line5-solution, exercise.reveal_solution = FALSE}
big <- logs |> filter(conference == "Big Ten")

ggplot() + 
  geom_line(data=big, aes(x=game_date, y=team_fgpct, group=team), color="grey") +
  geom_line(data=nu, aes(x=game_date, y=team_fgpct), color="red") + 
  geom_line(data=pu, aes(x=game_date, y=team_fgpct), color="gold") +
  geom_line(data=average, aes(x=game_date, y=mean_shooting), color="black") + 
  scale_y_continuous(limits = c(0, .7))
```
```{r line5-check}
grade_this_code()
```


## The Recap

In this lesson, we've explored the power of line charts for visualizing trends over time in sports data. We started with creating basic line charts using ggplot2, learning how to plot a single team's performance throughout a season. We then advanced to more complex visualizations, comparing multiple teams on the same chart and adding context with conference-wide data. We covered important techniques such as zeroing axes for honest representation, using color to highlight specific data series, and incorporating average lines for broader context. Throughout the lesson, we've seen how line charts can effectively show performance fluctuations, allow for easy comparisons between teams, and reveal patterns that might be missed in tabular data. Remember, while line charts are excellent for showing trends over time, they're most effective when used thoughtfully - consider your audience and the story you're trying to tell when deciding how many lines to include and how to style them. This skill in creating and interpreting line charts will be invaluable as you continue to analyze and present sports performance data.

## Terms to Know

- Line Chart: A type of graph that displays data points connected by straight line segments, typically used to show trends over time.
- `geom_line()`: A ggplot2 function used to create line charts by connecting data points with lines.
- Time Series: A sequence of data points indexed in time order, often visualized using line charts.
- x-axis: The horizontal axis in a chart, typically representing time in line charts.
- y-axis: The vertical axis in a chart, typically representing the measured variable in line charts.
- `scale_y_continuous()` or `scale_x_continuous()`: A ggplot2 function used to customize the y-axis or x-axis, including setting limits and breaks.
- Layering: The technique of adding multiple geom_line() functions to create a chart with multiple lines.
- Context Line: A line added to a chart to provide additional context, such as an average or benchmark, often distinguished by color or style.