---
title: "Sports Data Lesson 6: Filters"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to narrow in on what's important and remove what isn't.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(glue)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

The goal of this lesson is to introduce you to filtering and selecting data, two essential skills in data analysis. You'll learn how to use the `filter()` function to create subsets of your data based on specific criteria, allowing you to focus on the most relevant information for your analysis. We'll also cover the `select()` function, which enables you to choose specific columns from your dataset, streamlining your work by removing unnecessary information. By the end of this lesson, you'll be able to narrow down large datasets to exactly the information you need, whether it's looking at a specific team's performance, analyzing games with high rushing yards, or focusing on particular conferences. These skills are crucial for efficient data manipulation and analysis in sports analytics and beyond.

## The Basics

More often than not, we have more data than we want. Sometimes we need to be rid of that data. In `dplyr`, there's two ways to go about this: filtering and selecting.

**Filtering creates a subset of the data based on criteria**. All records where the count is greater than 10. All records that match "Nebraska". Something like that. 

**Selecting simply returns only the fields named**. So if you only want to see Team and TeamFGA, you select those fields. When you look at your data again, you'll have two columns. If you try to use one of your columns that you had before you used select, you'll get an error.  

<div class="alert alert-info">
  <h4 class="alert-heading">Key Concept</h4>
  <p class="mb-0"><code>filter()</code> limits rows. If you have 1000 rows and limit it to a single team, you'll have fewer than 1000 rows. <code>select()</code> limits columns. If you have 1000 rows and 25 columns, but you only need three columns, you'll still have 1000 rows when you are done using select.</p>
</div>

Let's work on some examples using college football game data from last season. First we'll need the tidyverse. Your first step is always loading libraries and you'll need to run this step in nearly every single thing you do.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
```
```{r load-tidyverse-solution}
library(tidyverse)
```
```{r load-tidyverse-check}
grade_this_code()
```
Now import the data.

```{r filtering-load-data, message=FALSE, warning=FALSE}
games <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballlogs25.csv")

rushinggames <- nrow(games |> filter(rushing_yds > 250))

bigrush <- games |> filter(rushing_yds > 250) |> filter(conference == "Big Ten Conference")

powerfour <- c("Southeastern Conference", "Big Ten Conference", "Big 12 Conference", "Atlantic Coast Conference")

toprush <- games |> 
  filter(conference %in% powerfour) |> 
  select(team, conference, rushing_yds, opponent) |>
  top_n(1, wt=rushing_yds)
```
```{r filtering-load-data-exercise, exercise = TRUE}
games <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballlogs25.csv")
```
```{r filtering-load-data-exercise-solution}
games <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballlogs25.csv")
```
```{r filtering-load-data-exercise-check}
grade_this_code()
```

So, first things first, let's say we don't care about all this Air Force, Akron, Alabama crap and just want to see Dear Old Nebraska U. We do that with `filter` and then we pass it a condition. 

Before we do anything, a note about conditions. Most of the conditional operators you'll understand -- greater than and less than are > and <. The tough one to remember is equal to. In conditional statements, equal to is == not =. If you haven't noticed, = is a variable assignment operator, not a conditional statement. **So equal is == and NOT equal is !=.**

<div class="alert alert-info">
  <h4 class="alert-heading">Key Concept</h4>
  <p class="mb-0">Conditional operators to remember:</p>
  <ul>
    <li>`==` is equal to.</li>
    <li>`!=` is does not equal.</li>
    <li>`<` is greater than.</li>
    <li>`<=` is greater than or equal to.</li>
    <li>`>` is less than.</li>
    <li>`>=` is less than or equal to.</li>
  </ul>
</div>

So if you want to see Nebraska, what would you do?

First, let's use head to give us a peek at the data. This data's layout is ... not great ... so you're going to want to scroll all the way to the right to see the most useful columns for this exercise.

```{r head-data, exercise=TRUE, exercise.setup = "filtering-load-data"}
head(____)
```
```{r head-data-solution, exercise.reveal_solution = FALSE}
head(games)
```
```{r head-data-check}
grade_this_code()
```

Now that we can see the column names, we now know Team is where the names of the universities are. Replace the all caps and question marks here and see if you can get only Nebraska.

### Exercise 1: Dear Old Nebraska U 

```{r filtering-nebraska, exercise=TRUE, exercise.setup = "filtering-load-data"}
____ |> filter(____ == "____")
```
```{r filtering-nebraska-solution, exercise.reveal_solution = FALSE}
games |> filter(team == "Nebraska")
```
```{r filtering-nebraska-check}
grade_this_code()
```

And what do you see? All of Nebraska's games last season. As much as we want to forget most of them, there they are.

### Exercise 2: Run the damn ball

There's a certain strain of football fan around these parts that is absolutely obsessed with the past (cough cough cough) I mean running the ball. To this fan, run the ball = win. So what if we wanted to see games where teams ran for more than 250 yards?

**To do this, we need to know which column is rushing yards** and we need to think about operators. We want more than 250 yards. Another name for more than, which you learned in math class in the third grade, is greater than. There's a symbol for that.

```{r filtering-rushers, exercise=TRUE, exercise.setup = "filtering-load-data"}
____ |> filter(____ > ___)
```
```{r filtering-rushers-solution, exercise.reveal_solution = FALSE}
games |> filter(rushing_yds > 250)
```
```{r filtering-rushers-check}
grade_this_code()
```

```{r rushers, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "filtering-load-data", results='asis'}
glue("Look there at the bottom of that window, below the data. See how many rows there are? Since each row is a game, that means there were were {rushinggames} games last season where a team ran for more than 250 yards.")
```

But how many of those were in the Big Ten? To do this we're going to *chain* filters together, which is a fancy way of just saying we're going to add another.

Note: The conference name *must* be exact. B1G or Big Ten won't work. You have to spell it and capitalize it *exactly* like it is in the data. 

### Exercise 3: B1G rushers

```{r filtering-big, exercise=TRUE, exercise.setup = "filtering-load-data"}
____ |> 
  filter(____ > ____) |> 
  filter(____ == "____")
```
```{r filtering-big-solution, exercise.reveal_solution = FALSE}
games |> 
  filter(rushing_yds > 250) |> 
  filter(conference == "Big Ten Conference")
```
```{r filtering-big-check}
grade_this_code()
```

```{r bigrushers, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "filtering-load-data", results='asis'}
glue("In the Big Ten last year, there were {nrow(bigrush)} games last season where a team ran for more than 250 yards. The next questions an analyst would ask are how many times did those teams win their games and does this hold consistently across seasons? Alas, another time. We have more filtering to do.")
```

### Exercise 4: Filtering with a list

But what if we want to see all of the Power Four conferences (RIP Pac-12)? We *could* use conditional logic in our filter. The conditional logic operators are `|` for OR and `&` for AND. **NOTE: AND means all conditions have to be met. OR means any of the conditions work.** So be careful about boolean logic. 

Here's what that looks like. It's a lot of extra code that you've got to be careful with.

```{r filtering-power5, exercise=TRUE, exercise.setup = "filtering-load-data"}
games |> 
  filter(
    conference == "Big Ten Conference" | 
      conference == "Southeastern Conference" | 
      conference == "Atlantic Coast Conference" | 
      conference == "Big 12 Conference")
```
```{r filtering-power5-solution}
games |> 
  filter(
    conference == "Big Ten Conference" | 
      conference == "Southeastern Conference" | 
      conference == "Atlantic Coast Conference" | 
      conference == "Big 12 Conference")
```
```{r filtering-power5-check}
grade_this_code()
```

But that's a lot of repetitive code. And a lot of typing. And typing is the devil. So what if we could create a list of conferences and pass it into the filter just once? It's pretty simple.

We can create a new variable -- remember variables can represent just about anything -- and create a list. To do that we use the `c` operator, which stands for concatenate. That just means take all the stuff in the parenthesis after the c and bunch it into a list. 

Note here: text is in quotes. If they were numbers, we wouldn't need the quotes. And `c()` gives us a list that we can use. 

With a list, we can use the %in% operator. It does what you think it does -- it gives you data that matches things IN the list you give it. 

```{r filtering-power, exercise=TRUE, exercise.setup = "filtering-load-data"}
powerfour <- c("Southeastern Conference", "Big Ten Conference", "Big 12 Conference", "Atlantic Coast Conference")

____ |> filter(____ %in% ____)
```
```{r filtering-power-solution}
powerfour <- c("Southeastern Conference", "Big Ten Conference", "Big 12 Conference", "Atlantic Coast Conference")

games |> filter(conference %in% powerfour)
```
```{r filtering-power-check}
grade_this_code()
```

## Selecting data to make it easier to read

So now we have our Power Four list. What if we just wanted to see team names, conferences and passing yards from the most recent season and ignore all the rest? Select to the rescue. With select, we just spell out exactly which columns we want to see. In our case, we've got three we're interested in seeing: Team, Conference and PassingYds. 

### Exercise 5: Select to simplify 

```{r selecting-power, exercise=TRUE, exercise.setup = "filtering-load-data"}
powerfour <- c("Southeastern Conference", "Big Ten Conference", "Big 12 Conference", "Atlantic Coast Conference")

____ |> 
  filter(____ %in% ____) |> 
  select(____, ____, ____)
```
```{r selecting-power-solution}
powerfour <- c("Southeastern Conference", "Big Ten Conference", "Big 12 Conference", "Atlantic Coast Conference")

games |> 
  filter(conference %in% powerfour) |> 
  select(team, conference, passing_yds)
```
```{r selecting-power-check}
grade_this_code()
```

If you have truly massive data, `select` has tools to help you select fields that start_with the same things or ends with a certain word. [The documentation will guide you](https://dplyr.tidyverse.org/reference/select.html) if you need those someday. For 90 plus percent of what we do, just naming the fields will be sufficient. 

## Top list

One last little dplyr trick that's nice to have in the toolbox is a shortcut for selecting only the top values for your dataset. Want to make a Top 10 List? Or Top 25? Or Top Whatever You Want? It's easy. 

### Exercise 6: Top N lists

So what are the top rushing games by Power Five schools? All we're doing here is chaining commands together with what we've already got. We're *filtering* by our list of Power Five conferences, we're *selecting* the three fields we need -- Team, Conference and the rushing yards column. Now we'll introduce the new function: `top_n`. The `top_n` function just takes a number and a weight, abbreviated wt. So we want a top 10 list? The number is pretty obvious. The weight is the column we're ranking on -- the rushing yards.

```{r selecting-top-power, exercise=TRUE, exercise.setup = "filtering-load-data"}
powerfour <- c("Southeastern Conference", "Big Ten Conference", "Big 12 Conference", "Atlantic Coast Conference")

____ |> 
  filter(____ %in% ____) |> 
  select(____, ____, ____) |>
  arrange(desc(____)) |> 
  top_n(____, wt=____)
```
```{r selecting-top-power-solution, exercise.reveal_solution = FALSE}
powerfour <- c("Southeastern Conference", "Big Ten Conference", "Big 12 Conference", "Atlantic Coast Conference")

games |> 
  filter(conference %in% powerfour) |> 
  select(team, conference, rushing_yds) |>
  arrange(desc(rushing_yds)) |> 
  top_n(10, wt=rushing_yds)
```
```{r selecting-top-power-check}
grade_this_code()
```

```{r teamrushers, exercise=FALSE, exercise.eval=TRUE, exercise.setup = "filtering-load-data", results='asis'}
glue("Last season, it was {toprush$Team} that went off for {toprush$RushingYds} yards in a single game. Impressive, until you realize it was against {toprush$Opponent}.")
```

## The Recap

In this lesson, we've explored two powerful tools for data manipulation: filtering and selecting. We learned how to use filter() to narrow down our dataset based on specific conditions, such as team names or performance metrics. We also discovered how to chain multiple filters together for more complex queries, and how to use lists with the %in% operator for efficient filtering of multiple categories. The select() function allowed us to focus on specific columns, simplifying our dataset for analysis. We combined these tools with previously learned functions like arrange() and introduced top_n() for creating ranked lists. These skills are fundamental in sports analytics, allowing you to drill down into the data that matters most for your analysis. Whether you're isolating performances of a specific team, comparing conferences, or identifying top performers, the ability to filter and select data efficiently will be crucial in your sports data analysis toolkit. Remember, effective data analysis often starts with getting the right subset of data, and the skills you've learned here will help you do just that.

## Terms to Know

- `filter()`: A function used to subset data by rows based on specified conditions.
- `select()`: A function used to choose specific columns from a dataset, reducing the number of variables.
- Conditional operators: Symbols used in filtering to create conditions (e.g., ==, >, <, >=, <=, !=).
- Boolean operators: Logical operators used to combine multiple conditions in filtering (e.g., & for AND, | for OR).
- `%in%`: An operator used to check if values are present in a specified set or list.
- `c()`: A function used to combine values into a vector or list.
- `top_n()`: A function used to select the top n rows based on a specified column.