---
title: "Sports Data Lesson 24: Tables"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn how to make a table with visual interestingness.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(gt)
library(wehoop)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

The goal of this lesson is to introduce you to creating informative and visually appealing tables using the gt package in R. By the end of this tutorial, you'll understand when to use tables for data presentation, how to create basic tables, and how to enhance them with various styling and formatting options. You'll learn to prepare data for tables, add headers and labels, style different elements, and use color to highlight specific information. We'll apply these concepts to real women's basketball data, analyzing shooting percentages across Big Ten teams. This lesson will enhance your ability to present complex data in a clear, readable format, allowing viewers to make their own comparisons while still highlighting key points. You'll see how tables can be an effective alternative to graphs in certain situations, especially when you want to provide detailed information that allows for multiple comparisons.

## The Basics

A table is the most simple form of data presentation. Tables have been around for millenia. We've improved in them since, so we're not making a table. We're making a table with features. 

Sometimes, the best way to show your data is with a table -- simple rows and columns. It allows a reader to compare whatever they want to compare a little easier than a graph where you've chosen what to highlight. The folks that made R Studio and the tidyverse have a neat package called `gt`.

For this assignment, we'll need `gt` and `wehoop`.

What does this library do? We need the tidyverse and gt to find out.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
library(gt)
library(wehoop)
```
```{r load-tidyverse-solution}
library(tidyverse)
library(gt)
library(wehoop)
```
```{r load-tidyverse-check}
grade_this_code()
```

For the data, I want to look at Nebraska women's basketball. I want to look at where they rank in the Big Ten in shooting metrics -- both field goal shooting and three point shooting percentages. To get that, I'm going to use wehoop to get team box scores. To start, I need to make a list with Big Ten teams in it, because wehoop doesn't include a conference column to make filtering easy. Then, I need to do some formatting work that we'll get into later in the class, but the short version is I'm splitting the way they handle made shots and attempted shots, turning them into numbers, totaling them up, calculating the season shooting percentages and then using select to just end with three columns: Team and my two shooting metrics. 

```{r table-load-data, message=FALSE, warning=FALSE}
big <- c("Nebraska", "Iowa", "Minnesota", "Illinois", "Wisconsin", "Michigan", "Michigan St", "Ohio State", "Purdue", "Indiana", "Maryland", "Rutgers", "Penn State", "Northwestern")

bigwbb <- load_wbb_team_box(seasons = 2026) |> 
  filter(team_short_display_name %in% big) |>
  select(team_short_display_name, field_goals_made, field_goals_attempted, three_point_field_goals_made, three_point_field_goals_attempted) |>
  mutate_at(vars(-team_short_display_name), as.numeric) |>
  group_by(team_short_display_name) |>
  summarize(
    TotalFG = sum(field_goals_made),
    TotalFGA = sum(field_goals_attempted),
    Total3P = sum(three_point_field_goals_made),
    Total3PA = sum(three_point_field_goals_attempted)
  ) |>
  mutate(FGPct = TotalFG/TotalFGA, ThreePct = Total3P/Total3PA) |>
  rename(Team = team_short_display_name) |>
  select(Team, FGPct, ThreePct) |>
  arrange(desc(FGPct))
```
```{r table-load-data-exercise, exercise = TRUE}
big <- c("Nebraska", "Iowa", "Minnesota", "Illinois", "Wisconsin", "Michigan", "Michigan St", "Ohio State", "Purdue", "Indiana", "Maryland", "Rutgers", "Penn State", "Northwestern")

bigwbb <- load_wbb_team_box(seasons = 2026) |> 
  filter(team_short_display_name %in% big) |>
  select(team_short_display_name, field_goals_made, field_goals_attempted, three_point_field_goals_made, three_point_field_goals_attempted) |>
  mutate_at(vars(-team_short_display_name), as.numeric) |>
  group_by(team_short_display_name) |>
  summarize(
    TotalFG = sum(field_goals_made),
    TotalFGA = sum(field_goals_attempted),
    Total3P = sum(three_point_field_goals_made),
    Total3PA = sum(three_point_field_goals_attempted)
  ) |>
  mutate(FGPct = TotalFG/TotalFGA, ThreePct = Total3P/Total3PA) |>
  rename(Team = team_short_display_name) |>
  select(Team, FGPct, ThreePct) |>
  arrange(desc(FGPct))
```
```{r table-load-data-exercise-solution}
big <- c("Nebraska", "Iowa", "Minnesota", "Illinois", "Wisconsin", "Michigan", "Michigan St", "Ohio State", "Purdue", "Indiana", "Maryland", "Rutgers", "Penn State", "Northwestern")

bigwbb <- load_wbb_team_box(seasons = 2026) |> 
  filter(team_short_display_name %in% big) |>
  select(team_short_display_name, field_goals_made, field_goals_attempted, three_point_field_goals_made, three_point_field_goals_attempted) |>
  mutate_at(vars(-team_short_display_name), as.numeric) |>
  group_by(team_short_display_name) |>
  summarize(
    TotalFG = sum(field_goals_made),
    TotalFGA = sum(field_goals_attempted),
    Total3P = sum(three_point_field_goals_made),
    Total3PA = sum(three_point_field_goals_attempted)
  ) |>
  mutate(FGPct = TotalFG/TotalFGA, ThreePct = Total3P/Total3PA) |>
  rename(Team = team_short_display_name) |>
  select(Team, FGPct, ThreePct) |>
  arrange(desc(FGPct))
```
```{r table-load-data-exercise-check}
grade_this_code()
```

We've used head a bunch of times in this series to see data, but gt makes this super easy too. 

```{r gt-data, exercise=TRUE, exercise.setup = "table-load-data"}
bigwbb |> gt()
```
```{r gt-data-solution}
bigwbb |> gt()
```
```{r gt-data-check}
grade_this_code()
```

So there you have it. You made a table. You're done, right? 

`gt` has a mountain of customization options. The good news is that it works in a very familiar pattern -- we start with data and we use |> to pipe to the next thing.

### Exercise 1: Making headers into real words

We'll start with fixing headers. What we have isn't bad, but they aren't good either. Let's fix that. The way this works is we need to tell `cols_label` what columns we want to change and what we want to change them into.

```{r table1, exercise=TRUE, exercise.setup = "table-load-data", message=FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    ____ = "Shooting Percentage",
    ____ = "Three Point Percentage"
  )
```
```{r table1-solution, exercise.reveal_solution = FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    FGPct = "Shooting Percentage",
    ThreePct = "Three Point Percentage"
  )
```
```{r table1-check}
grade_this_code()
```

Better. Note the pattern: Actual header name = "What we want to see". So if we wanted to change Team to School, we'd do this: `Team = "School"` inside the `cols_label` bits.

Now we can start working with styling. 

### Exercise 2: Styling, part 1

The truth is most of your code in tables is going to be dedicated to styling specific things. The first thing we need: A headline and some chatter. They're required parts of a graphic -- we'll talk more about that soon -- so they're a good place to start. We do that with `tab_header` and in there we need a title and subtitle.

```{r table2, exercise=TRUE, exercise.setup = "table-load-data", message=FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    ____ = "Shooting Percentage",
    ____ = "Three Point Percentage"
  ) |>
  tab_header(
    ____ = "Does Nebraska have a chance to be special this season?",
    ____ = "Amy Williams' team is near the middle of a tough conference where it matters most."
  )
```
```{r table2-solution, exercise.reveal_solution = FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    FGPct = "Shooting Percentage",
    ThreePct = "Three Point Percentage"
  ) |>
  tab_header(
    title = "Does Nebraska have a chance to be special this season?",
    subtitle = "Amy Williams' team is near the middle of a tough conference where it matters most."
  )
```
```{r table2-check}
grade_this_code()
```

We have a headline and some chatter, but ... gross. Centered? The extra lines? No real difference in font weight? We can do better. 

### Exercise 3: Changing typography

We can style individual elements using `tab_style`. First, let's make the main headline -- the `title` -- bold and left aligned. Then, we can left align the `subtitle`, but leave it normal weight (i.e. not bold).

```{r table3, exercise=TRUE, exercise.setup = "table-load-data", message=FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    ____ = "Shooting Percentage",
    ____ = "Three Point Percentage"
  ) |>
  tab_header(
    ____ = "Does Nebraska have a chance to be special this season?",
    ____ = "Amy Williams' team is near the middle of a tough conference where it matters most."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  )
```
```{r table3-solution, exercise.reveal_solution = FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    FGPct = "Shooting Percentage",
    ThreePct = "Three Point Percentage"
  ) |>
  tab_header(
    title = "Does Nebraska have a chance to be special this season?",
    subtitle = "Amy Williams' team is near the middle of a tough conference where it matters most."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  )
```
```{r table3-check}
grade_this_code()
```

### Exercise 4: Adding source and credit lines

The next item on the required elements list: Source and credit lines. In `gt`, those are called a `tab_source_note`:

```{r table4, exercise=TRUE, exercise.setup = "table-load-data", message=FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    ____ = "Shooting Percentage",
    ____ = "Three Point Percentage"
  ) |>
  tab_header(
    ____ = "Does Nebraska have a chance to be special this season?",
    ____ = "Amy Williams' team is near the middle of a tough conference where it matters most."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  ______________(
    source_note = md("**By:** Matt Waite  |  **Source:** wehoop: The SportsDataverse's R Package for Women's Basketball Data")
  )
```
```{r table4-solution, exercise.reveal_solution = FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    FGPct = "Shooting Percentage",
    ThreePct = "Three Point Percentage"
  ) |>
  tab_header(
    title = "Does Nebraska have a chance to be special this season?",
    subtitle = "Amy Williams' team is near the middle of a tough conference where it matters most."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Matt Waite  |  **Source:** wehoop: The SportsDataverse's R Package for Women's Basketball Data")
  )
```
```{r table4-check}
grade_this_code()
```

## More styling

We can do a lot with `tab_style`. For instance, we can make the headers bold and reduce the size a bit to reduce font congestion in the area.

```{r table5, exercise=TRUE, exercise.setup = "table-load-data", message=FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    FGPct = "Shooting Percentage",
    ThreePct = "Three Point Percentage"
  ) |>
  tab_header(
    title = "Does Nebraska have a chance to be special this season?",
    subtitle = "Amy Williams' team is near the middle of a tough conference where it matters most."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Matt Waite  |  **Source:** wehoop: The SportsDataverse's R Package for Women's Basketball Data")
  ) |>
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   ) 
```
```{r table5-solution, exercise.reveal_solution = FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    FGPct = "Shooting Percentage",
    ThreePct = "Three Point Percentage"
  ) |>
  tab_header(
    title = "Does Nebraska have a chance to be special this season?",
    subtitle = "Amy Williams' team is near the middle of a tough conference where it matters most."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Matt Waite  |  **Source:** wehoop: The SportsDataverse's R Package for Women's Basketball Data")
  ) |>
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   ) 
```
```{r table5-check}
grade_this_code()
```

Next up: There's a lot of lines in this that don't need to be there. `gt` has some tools to get rid of them easily and add in some other readability improvements: `opt_row_striping` and `opt_table_lines`.

```{r table6, exercise=TRUE, exercise.setup = "table-load-data", message=FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    FGPct = "Shooting Percentage",
    ThreePct = "Three Point Percentage"
  ) |>
  tab_header(
    title = "Does Nebraska have a chance to be special this season?",
    subtitle = "Amy Williams' team is near the middle of a tough conference where it matters most."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Matt Waite  |  **Source:** wehoop: The SportsDataverse's R Package for Women's Basketball Data")
  ) |>
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   ) |>
  opt_row_striping() |> 
  opt_table_lines("none")
```
```{r table6-solution, exercise.reveal_solution = FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    FGPct = "Shooting Percentage",
    ThreePct = "Three Point Percentage"
  ) |>
  tab_header(
    title = "Does Nebraska have a chance to be special this season?",
    subtitle = "Amy Williams' team is near the middle of a tough conference where it matters most."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Matt Waite  |  **Source:** wehoop: The SportsDataverse's R Package for Women's Basketball Data")
  ) |>
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   ) |>
  opt_row_striping() |> 
  opt_table_lines("none")
```
```{r table6-check}
grade_this_code()
```

We're in pretty good shape here, but look closer. What else makes this table sub-par? How about the formatting of the percent change? We can fix that with a formatter.

```{r table7, exercise=TRUE, exercise.setup = "table-load-data", message=FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    FGPct = "Shooting Percentage",
    ThreePct = "Three Point Percentage"
  ) |>
  tab_header(
    title = "Does Nebraska have a chance to be special this season?",
    subtitle = "Amy Williams' team is near the middle of a tough conference where it matters most."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Matt Waite  |  **Source:** wehoop: The SportsDataverse's R Package for Women's Basketball Data")
  ) |>
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   ) |>
  opt_row_striping() |> 
  opt_table_lines("none") |>
    fmt_percent(
    columns = c(FGPct, ThreePct),
    decimals = 1
  )
```
```{r table7-solution, exercise.reveal_solution = FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    FGPct = "Shooting Percentage",
    ThreePct = "Three Point Percentage"
  ) |>
  tab_header(
    title = "Does Nebraska have a chance to be special this season?",
    subtitle = "Amy Williams' team is near the middle of a tough conference where it matters most."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Matt Waite  |  **Source:** wehoop: The SportsDataverse's R Package for Women's Basketball Data")
  ) |>
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   ) |>
  opt_row_striping() |> 
  opt_table_lines("none") |>
    fmt_percent(
    columns = c(FGPct, ThreePct),
    decimals = 1
  )
```
```{r table7-check}
grade_this_code()
```

Throughout this series, we've been using color and other signals to highlight things. With a little `tab_style` magic, we can change individual rows and add color. The last `tab_style` block here will first pass off the styles we want to use -- we're going to make the rows red and the text white -- and then for locations we specify where with a simple filter. What that means is that any rows we can address with logic -- all rows with a value greater than X, for example -- we can change the styling. 

```{r table8, exercise=TRUE, exercise.setup = "table-load-data", message=FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    FGPct = "Shooting Percentage",
    ThreePct = "Three Point Percentage"
  ) |>
  tab_header(
    title = "Does Nebraska have a chance to be special this season?",
    subtitle = "Amy Williams' team is near the middle of a tough conference where it matters most."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Matt Waite  |  **Source:** wehoop: The SportsDataverse's R Package for Women's Basketball Data")
  ) |>
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   ) |>
  opt_row_striping() |> 
  opt_table_lines("none") |>
    fmt_percent(
    columns = c(FGPct, ThreePct),
    decimals = 1
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "red"),
      cell_text(color = "white")
      ),
    locations = cells_body(
      rows = Team == "Nebraska")
  )
```
```{r table8-solution, exercise.reveal_solution = FALSE}
bigwbb |> 
  gt() |> 
  cols_label(
    FGPct = "Shooting Percentage",
    ThreePct = "Three Point Percentage"
  ) |>
  tab_header(
    title = "Does Nebraska have a chance to be special this season?",
    subtitle = "Amy Williams' team is near the middle of a tough conference where it matters most."
  ) |> tab_style(
    style = cell_text(color = "black", weight = "bold", align = "left"),
    locations = cells_title("title")
  ) |> tab_style(
    style = cell_text(color = "black", align = "left"),
    locations = cells_title("subtitle")
  ) |>
  tab_source_note(
    source_note = md("**By:** Matt Waite  |  **Source:** wehoop: The SportsDataverse's R Package for Women's Basketball Data")
  ) |>
  tab_style(
     locations = cells_column_labels(columns = everything()),
     style = list(
       cell_borders(sides = "bottom", weight = px(3)),
       cell_text(weight = "bold", size=12)
     )
   ) |>
  opt_row_striping() |> 
  opt_table_lines("none") |>
    fmt_percent(
    columns = c(FGPct, ThreePct),
    decimals = 1
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "red"),
      cell_text(color = "white")
      ),
    locations = cells_body(
      rows = Team == "Nebraska")
  )
```
```{r table8-check}
grade_this_code()
```

We've arrived where we want to be: We've created a clear table that allows a reader to compare schools at will while also using color to draw attention to the thing we want to draw attention to. **We've kept it simple so the color has impact**.

## The Recap

In this lesson, we've explored the power of tables for presenting sports data using the gt package in R. We started by learning how to create basic tables from our dataset, then progressively enhanced them with various features. We covered important techniques such as adding meaningful headers and labels, styling text elements for better readability, and incorporating source information. We learned how to use tab_style() to customize specific parts of our table, including making headers bold and adding bottom borders. We also explored how to improve the table's overall appearance by using row striping and removing unnecessary lines. Importantly, we practiced formatting numerical data appropriately, converting raw numbers into percentages. Finally, we learned how to use color strategically to highlight specific data points, enhancing the table's ability to tell a story. Throughout the process, we saw how tables can effectively present detailed information while still guiding the viewer's attention to key points. Remember, while tables may seem simple, thoughtful design and formatting can make them powerful tools for communicating sports analytics insights.

## Terms to Know

- `gt`: An R package used for creating customizable tables in data analysis and presentation.
- `tab_header()`: A function in gt used to add a title and subtitle to a table.
- `cols_label()`: A function in gt used to rename column headers in a table.
- `tab_style()`: A function in gt used to apply custom styles to specific parts of a table.
- `tab_source_note()`: A function in gt used to add source information or notes at the bottom of a table.