---
title: "Sports Data Lesson 18: Slope charts"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
    theme: united
    ace_theme: github
runtime: shiny_prerendered
description: >
  Learn another way to show changes between two time periods.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

## The Goal

The goal of this lesson is to introduce you to slope charts, a powerful visualization tool for comparing data between two time periods. By the end of this tutorial, you'll understand when to use slope charts, how to create them using ggplot2, and how to customize them for clearer data representation. You'll learn to prepare data for slope charts, plot lines and points to create the slope effect, use color effectively to highlight specific teams, and add labels to enhance your chart's readability. We'll apply these concepts to real college football data, analyzing team performance across two seasons using yards per play statistics.

## The Basics

A line chart, as we learned, shows change over time, with the date on the x-axis and each line representing a team or player. But sometimes, you have just two seasons -- a line chart of two seasons is ... not a line chart. 

But with some fiddling with lines and points, you can create a new chart type that does show change over two time periods quite well: A slope chart. 

Think of a slope chart like the meme you see online: How is started vs how it's going. 

Nebraska, in 2024, started a five-star freshman for the first time in school history. What happened next? I don't remember details. Something about a bowl game for the first time since the Truman administration or something like that. Anywho. What does that look like on offense?

You're going to need a dataset I created that has two seasons of Yards Per Play data on offense. This is the total yards gained divided by the total number of plays. For each team, there's two rows -- one in 2024 and one in 2025. 

We'll need the tidyverse.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
```
```{r load-tidyverse-solution}
library(tidyverse)
```
```{r load-tidyverse-check}
grade_this_code()
```

And the data.

```{r slope-load-data, message=FALSE, warning=FALSE}
yards_per_play <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballslope2425.csv")

nu <- yards_per_play |> filter(team == "Nebraska")
oh <- yards_per_play |> filter(team == "Ohio State")
id <- yards_per_play |> filter(team == "Indiana")

big <- yards_per_play |> 
  filter(conference == "Big Ten Conference")
```
```{r slope-load-data-exercise, exercise = TRUE}
yards_per_play <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballslope2425.csv")
```
```{r slope-load-data-exercise-solution}
yards_per_play <- read_csv("https://mattwaite.github.io/sportsdatafiles/footballslope2425.csv")
```
```{r slope-load-data-exercise-check}
grade_this_code()
```


### Exercise 1: Setting up the data

Our data is in pretty good shape already. The problem is, there's just too much of it. So let's first filter our data so we only have the `Big Ten Conference`, and we're going to put that in a dataframe called `big`. 
```{r slope1, exercise=TRUE, exercise.setup = "slope-load-data", message=FALSE}
____ <- yards_per_play |> 
  filter(conference == "____")
```
```{r slope1-solution, exercise.reveal_solution = FALSE}
big <- yards_per_play |> 
  filter(conference == "Big Ten Conference")
```
```{r slope1-check}
grade_this_code()
```

And we can glimpse that quick to see what we've got. 

```{r glimpse-data, exercise=TRUE, exercise.setup = "slope-load-data"}
glimpse(____)
```
```{r glimpse-data-solution}
glimpse(big)
```
```{r glimpse-data-check}
grade_this_code()
```

Better. 

### Exercise 2: Making our first slope chart. 

A slope chart is really made up of two separate geoms that combine to make a form -- a geom_point and a geom_line. The form the two geoms take is identical. In this case, we're going to have the x be the `season`, the y is our yards per play number and we're going to add a group value, which is the `team`. I'm also going to go ahead and set the color for these to grey, which you'll see why in a moment. 

```{r slope2, exercise=TRUE, exercise.setup = "slope-load-data", message=FALSE}
ggplot() + 
  geom_line(data=____, aes(x=season, y=____, group=team), color="grey") +
  geom_point(data=____, aes(x=season, y=____, group=team), color="grey")
```
```{r slope2-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_line(data=big, aes(x=season, y=yards_per_play, group=team), color="grey") +
  geom_point(data=big, aes(x=season, y=yards_per_play, group=team), color="grey")
```
```{r slope2-check}
grade_this_code()
```

Well, we got something. But where is Nebraska? And who is that so far ahead of everyone else? 

### Exercise 3: More filtering

Because of where we are right now, we should use filtering to find Nebraska first. Then, I have to believe Ohio State is always near the top in this, so let's find them too. Let's also do Indiana, because ... well ... they're good. 

```{r slope3, exercise=TRUE, exercise.setup = "slope-load-data", message=FALSE}
nu <- yards_per_play |> filter(team == "____")
oh <- yards_per_play |> filter(team == "____")
id <- yards_per_play |> filter(team == "____")
```
```{r slope3-solution, exercise.reveal_solution = FALSE}
nu <- yards_per_play |> filter(team == "Nebraska")
oh <- yards_per_play |> filter(team == "Ohio State")
id <- yards_per_play |> filter(team == "Indiana")
```
```{r slope3-check}
grade_this_code()
```

Now that we have these schools isolated, we can layer them in. 

### Exercise 4: Layering with color

Adding in layers in ggplot is easy, we just have to be aware of ordering and color. Putting a black line on a black line will not do much. So we're going to add our Big Ten points and lines first, then, we'll layer in the other schools. For each school, we'll add two geoms -- the point and the line. The aes will be the same as the Big Ten with one exception -- the color. We'll use black for Ohio State and blue for Indiana because we're red, screw them. 

```{r slope4, exercise=TRUE, exercise.setup = "slope-load-data", message=FALSE}
ggplot() + 
  geom_line(data=big, aes(x=season, y=yards_per_play, group=team), color="grey") +
  geom_point(data=big, aes(x=season, y=yards_per_play, group=team), color="grey") +
  geom_line(data=____, aes(x=season, y=yards_per_play, group=team), color="red") +
  geom_point(data=____, aes(x=season, y=yards_per_play, group=team), color="red") +
  geom_line(data=____, aes(x=season, y=yards_per_play, group=team), color="black") +
  geom_point(data=____, aes(x=season, y=yards_per_play, group=team), color="black") +
  geom_line(data=____, aes(x=season, y=yards_per_play, group=team), color="blue") +
  geom_point(data=____, aes(x=season, y=yards_per_play, group=team), color="blue")
```
```{r slope4-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_line(data=big, aes(x=season, y=yards_per_play, group=team), color="grey") +
  geom_point(data=big, aes(x=season, y=yards_per_play, group=team), color="grey") +
  geom_line(data=nu, aes(x=season, y=yards_per_play, group=team), color="red") +
  geom_point(data=nu, aes(x=season, y=yards_per_play, group=team), color="red") +
  geom_line(data=oh, aes(x=season, y=yards_per_play, group=team), color="black") +
  geom_point(data=oh, aes(x=season, y=yards_per_play, group=team), color="black") +
  geom_line(data=id, aes(x=season, y=yards_per_play, group=team), color="blue") +
  geom_point(data=id, aes(x=season, y=yards_per_play, group=team), color="blue")
```
```{r slope4-check}
grade_this_code()
```

What story does this tell? Follow the Ohio State line. Follow the Indiana line. Surprising? How about Nebraska?

### Exercise 5: Getting creative with spacing and breaks

Now that we have something to work with, it's time to start improving it. It would be nice if the years at the bottom didn't have a lot of meaningless decimals on them -- as in we only have two years -- and also, it would be good to label each team. Maybe we label them on each end of the graph. Given how much space the default ggplot behavior has left us to do that, we're going to need to make some space.

We can do that with some clever manipulations of the x axis. In ggplot, there are all kinds of ways to work with the x and y axis scales. In this case, we have a continuous number on the x axis, so we'll use `scale_x_continuous` and we're going to first set the line breaks manually so there's just 2024 and 2025. Then, we're going to set the limits of our x axis so that it doesn't just default to 2024-2025, but is a little wider so we have extra space. We're going to add a half a year on each side to pad it out. 

```{r slope5, exercise=TRUE, exercise.setup = "slope-load-data", message=FALSE}
ggplot() + 
  geom_line(data=big, aes(x=season, y=yards_per_play, group=team), color="grey") +
  geom_point(data=big, aes(x=season, y=yards_per_play, group=team), color="grey") +
  geom_line(data=nu, aes(x=season, y=yards_per_play, group=team), color="red") +
  geom_point(data=nu, aes(x=season, y=yards_per_play, group=team), color="red") +
  geom_line(data=oh, aes(x=season, y=yards_per_play, group=team), color="black") +
  geom_point(data=oh, aes(x=season, y=yards_per_play, group=team), color="black") +
  geom_line(data=id, aes(x=season, y=yards_per_play, group=team), color="blue") +
  geom_point(data=id, aes(x=season, y=yards_per_play, group=team), color="blue") + 
  scale_x_continuous(breaks=c(____, ____), limits=c(2023.5,2025.5)) 
```
```{r slope5-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_line(data=big, aes(x=season, y=yards_per_play, group=team), color="grey") +
  geom_point(data=big, aes(x=season, y=yards_per_play, group=team), color="grey") +
  geom_line(data=nu, aes(x=season, y=yards_per_play, group=team), color="red") +
  geom_point(data=nu, aes(x=season, y=yards_per_play, group=team), color="red") +
  geom_line(data=oh, aes(x=season, y=yards_per_play, group=team), color="black") +
  geom_point(data=oh, aes(x=season, y=yards_per_play, group=team), color="black") +
  geom_line(data=id, aes(x=season, y=yards_per_play, group=team), color="blue") +
  geom_point(data=id, aes(x=season, y=yards_per_play, group=team), color="blue") + 
  scale_x_continuous(breaks=c(2024, 2025), limits=c(2023.5,2025.5)) 
```
```{r slope5-check}
grade_this_code()
```

A ha! Better. We now just have the two years and we have some space to put our team names in there to help people.

### Exercise 6: adding labels

To add the labels, we again need to be a bit clever. We want to label each dot *twice* -- once for 2024 and once for 2025. But our data has both years in it, so we're going to have to do something we don't do a lot of: put a `dplyr` filter into our data= bits. 

We're going to use a `geom_text`, which will just put text anywhere on your graph. The aes will be the same as the geom_point, except it will have a label in it that we'll set to the team. Then, for data, we're going to filter our dataframes for the `min(season)` and the `max(season)`, which will give us 2024 and 2025 respectively. Then, one last little trick -- we're going to nudge our labels over a little. For our min, we're going to subtract .2 years off our season, which will have the effect of moving the text to the left just a bit. Then, on the max side, we'll nudge it .2 to the right by adding it to the season. 

To cap it off, we'll add `theme_minimal()` at the end because we want to remove the grey background. 

```{r slope6, exercise=TRUE, exercise.setup = "slope-load-data", message=FALSE}
ggplot() + 
  geom_line(data=big, aes(x=season, y=yards_per_play, group=team), color="grey") +
  geom_point(data=big, aes(x=season, y=yards_per_play, group=team), color="grey") +
  geom_line(data=nu, aes(x=season, y=yards_per_play, group=team), color="red") +
  geom_point(data=nu, aes(x=season, y=yards_per_play, group=team), color="red") +
  geom_line(data=oh, aes(x=season, y=yards_per_play, group=team), color="black") +
  geom_point(data=oh, aes(x=season, y=yards_per_play, group=team), color="black") +
  geom_line(data=id, aes(x=season, y=yards_per_play, group=team), color="blue") +
  geom_point(data=id, aes(x=season, y=yards_per_play, group=team), color="blue") +
  geom_text(data=nu |> filter(season == max(____)), aes(x=season + .2, y=yards_per_play+.1, group=team, label=____)) + 
  geom_text(data=nu |> filter(season == min(____)), aes(x=season - .2, y=yards_per_play, group=team, label=____)) + 
  geom_text(data=id |> filter(season == max(season)), aes(x=season + .2, y=yards_per_play, group=team, label=team)) + 
  geom_text(data=id |> filter(season == min(season)), aes(x=season - .2, y=yards_per_play, group=team, label=team)) + 
  geom_text(data=oh |> filter(season == max(season)), aes(x=season + .2, y=yards_per_play, group=team, label=team)) + 
  geom_text(data=oh |> filter(season == min(season)), aes(x=season - .2, y=yards_per_play, group=team, label=team)) + 
  scale_x_continuous(breaks=c(2024, 2025), limits=c(2023.5,2025.5))  +
  theme_minimal()
```
```{r slope6-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_line(data=big, aes(x=season, y=yards_per_play, group=team), color="grey") +
  geom_point(data=big, aes(x=season, y=yards_per_play, group=team), color="grey") +
  geom_line(data=nu, aes(x=season, y=yards_per_play, group=team), color="red") +
  geom_point(data=nu, aes(x=season, y=yards_per_play, group=team), color="red") +
  geom_line(data=oh, aes(x=season, y=yards_per_play, group=team), color="black") +
  geom_point(data=oh, aes(x=season, y=yards_per_play, group=team), color="black") +
  geom_line(data=id, aes(x=season, y=yards_per_play, group=team), color="blue") +
  geom_point(data=id, aes(x=season, y=yards_per_play, group=team), color="blue") +
  geom_text(data=nu |> filter(season == max(season)), aes(x=season + .2, y=yards_per_play+.1, group=team, label=team)) + 
  geom_text(data=nu |> filter(season == min(season)), aes(x=season - .2, y=yards_per_play, group=team, label=team)) + 
  geom_text(data=id |> filter(season == max(season)), aes(x=season + .2, y=yards_per_play, group=team, label=team)) + 
  geom_text(data=id |> filter(season == min(season)), aes(x=season - .2, y=yards_per_play, group=team, label=team)) + 
  geom_text(data=oh |> filter(season == max(season)), aes(x=season + .2, y=yards_per_play, group=team, label=team)) + 
  geom_text(data=oh |> filter(season == min(season)), aes(x=season - .2, y=yards_per_play, group=team, label=team)) + 
  scale_x_continuous(breaks=c(2024, 2025), limits=c(2023.5,2025.5)) +
  theme_minimal()
```
```{r slope6-check}
grade_this_code()
```

And there you have it. A clean and clear slope chart showing how last season compared to the previous one.

## The Recap

In this lesson, we've explored the power of slope charts for visualizing changes between two time periods in sports data. We started by learning how to prepare and filter data for slope charts, focusing on specific teams and conferences. We then progressed to creating basic slope charts using a combination of geom_line() and geom_point(), and advanced to more complex visualizations by layering multiple teams with distinct colors. We covered important techniques such as manipulating axis scales to create space for labels, adding text annotations to identify teams, and using theme_minimal() for a cleaner look. Throughout the lesson, we've seen how slope charts can effectively show changes in team performance between seasons, allowing for easy comparisons and revealing patterns that might be missed in tabular data or other chart types. Remember, while slope charts are excellent for comparing two time points, they're most effective when used thoughtfully - consider your color choices, labeling, and overall design to ensure your chart tells a clear and compelling story. 

## Terms to Know

- Slope Chart: A type of chart that compares changes in data between two time points, using lines to connect the values for each entity.
- `geom_line()`: A ggplot2 function used to create the connecting lines in a slope chart.
- `geom_point()`: A ggplot2 function used to create the end points in a slope chart.
- Layering: The technique of adding multiple geom functions to create a more complex chart, often used to highlight specific data series in slope charts.
- `scale_x_continuous()`: A ggplot2 function used to customize the x-axis, including setting breaks and limits.
- `geom_text()`: A ggplot2 function used to add text labels to a chart, often used in slope charts to identify entities.