---
title: "Lesson 15: Waffle charts"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to make another chart that shows magnitude and composition.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(waffle)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

# Waffle charts

## The basics

Pie charts are the devil. They should be an instant F in any data visualization class. The problem? How carefully can you evaluate angles and area? Unless they are blindingly obvious and only a few categories, not well. If you've got 25 categories, how can you tell the difference between 7 and 9 percent? You can't.

So let's introduce a better way: The Waffle Chart. Some call it a square pie chart. I personally hate that. Waffles it is. 

**A waffle chart is designed to show you parts of the whole -- proportionality**. How many yards on offense come from rushing or passing. How many singles, doubles, triples and home runs make up a teams hits. How many shots a basketball team takes are two pointers versus three pointers. 

First, the waffle library should be installed, but any time you run across a library you don't have, you can go to the console in R Studio and run something like this: `install.packages("waffle")`

Now load it. For this exercise, we don't need the tidyverse. 

```{r load-waffle, exercise=TRUE}
library(waffle)
```
```{r load-waffle-solution}
library(waffle)
```
```{r load-waffle-check}
grade_this_code()
```

## Making waffles with vectors

Let's look at the debacle that was Nebraska vs. Michigan State in fall 2021 in college football. [Here's the box score](https://www.espn.com/college-football/matchup?gameId=401282784), which we'll use for this part of the walkthrough. 

Maybe the easiest way to do waffle charts, at least at first, is to make vectors of your data and plug them in. To make a vector, we use the `c` or concatenate function. We did this all the way back in Lesson 1. 

So let's look at offense. Rushing vs passing. 

```{r waffle-load-data, message=FALSE, warning=FALSE}
nu <- c("Rushing"=187, "Passing"=255)
ms <- c("Rushing"=71, "Passing"=183)

passing <- c("Nebraska"=255, "Mighigan State"=183)

nu2 <- c("Rushing"=187, "Passing"=255)
ms2 <- c("Rushing"=71, "Passing"=183, 188)
```
```{r waffle-load-data-exercise, exercise = TRUE}
nu <- c("Rushing"=187, "Passing"=255)
ms <- c("Rushing"=71, "Passing"=183)
```
```{r waffle-load-data-exercise-solution}
nu <- c("Rushing"=187, "Passing"=255)
ms <- c("Rushing"=71, "Passing"=183)
```
```{r waffle-load-data-exercise-check}
grade_this_code()
```

So what does the breakdown of the night look like?

### Exercise 1: Make a waffle chart

The waffle library can break this down in a way that's easier on the eyes than a pie chart. We call the `waffle` function, add the data for Nebraska, specify the number of rows (10 is a good start), give it a title and an x value label, and to clean up a quirk of the library, we've got to specify colors. 

```{r waffle, exercise=TRUE, exercise.setup = "waffle-load-data", message=FALSE}
waffle(
        ??, 
        rows = ??, 
        title="Nebraska's offense", 
        xlab="1 square = 1 yard", 
        colors = c("black", "red")
)
```
```{r waffle-solution, exercise.reveal_solution = FALSE}
waffle(
        nu, 
        rows = 10, 
        title="Nebraska's offense", 
        xlab="1 square = 1 yard", 
        colors = c("black", "red")
)
```
```{r waffle-check}
grade_this_code()
```

Or, we could make this two teams in the same chart by just changing the vector up. Nebraska had 255 yards passing that night. Michigan State had 183.

### Exercise 2: Two teams, one waffle

```{r waffle2, exercise=TRUE, exercise.setup = "waffle-load-data", message=FALSE}
passing <- c("Nebraska"=???, "???????? ?????"=183)

waffle(
        ???????, 
        rows = 10, 
        title="Nebraska vs Michigan State: passing", 
        xlab="1 square = 1 yard", 
        colors = c("red", "dark green")
)
```
```{r waffle2-solution, exercise.reveal_solution = FALSE}
passing <- c("Nebraska"=255, "Michigan State"=183)

waffle(
        passing, 
        rows = 10, 
        title="Nebraska vs Michigan State: passing", 
        xlab="1 square = 1 yard", 
        colors = c("red", "dark green")
)
```
```{r waffle2-check}
grade_this_code()
```

No matter how you look at this game, it just comes back to heartache.

## Two waffles = waffle iron

What does it look like if we compare the two teams using the two vectors in the same chart? To do that -- and I am not making this up -- you have to create a waffle iron. Get it? Waffle charts? Iron? 

### Exercise 3: The waffle iron

To make an waffle iron, you wrap your `waffle` functions in an `iron` function. The `iron` is just a wrapper -- it just combines them together. Each waffle functions seperately in the iron. 

In this block, we're going to use the first vectors we made with nu and ms. You'll see, the iron is just two waffles with a comma between them.

```{r waffle3, exercise=TRUE, exercise.setup = "waffle-load-data", message=FALSE}
iron(
 waffle(??, 
        rows = 10, 
        title="Nebraska's offense", 
        xlab="1 square = 1 yard", 
        colors = c("black", "red")
        ),
 waffle(??, 
        rows = 10, 
        title="Michigan State's offense", 
        xlab="1 square = 1 yard", 
        colors = c("black", "dark green")
        )
)
```
```{r waffle3-solution, exercise.reveal_solution = FALSE}
iron(
 waffle(nu, 
        rows = 10, 
        title="Nebraska's offense", 
        xlab="1 square = 1 yard", 
        colors = c("black", "red")
        ),
 waffle(ms, 
        rows = 10, 
        title="Michigan State's offense", 
        xlab="1 square = 1 yard", 
        colors = c("black", "dark green")
        )
)
```
```{r waffle3-check}
grade_this_code()
```

What do you notice about this chart? Notice how the squares aren't the same size? Well, Nebraska out-gained Michigan State by a long way ... AND LOST. So the squares aren't the same size because the numbers aren't the same. 

### Exercise 4: Adding padding

We can fix the uneven box sizes by adding an unnamed padding number so the number of yards add up to the same thing. Nebraka's total yards ended up at 442. To make the squares the same, we need to make the total for everyone be 442. To do that, we need to add 188 to Michigan State. REMEMBER: Don't name it or it'll show up in the legend.

Now, in our waffle iron, if we don't give that padding a color, we'll get an error. So we need to make it white. Which, given our white background, means it will disappear.

```{r waffle4, exercise=TRUE, exercise.setup = "waffle-load-data", message=FALSE}
nu2 <- c("Rushing"=187, "Passing"=255)
ms2 <- c("Rushing"=71, "Passing"=183, 188)

iron(
 waffle(???, 
        rows = 10, 
        title="Nebraska's offense", 
        xlab="1 square = 1 yard", 
        colors = c("black", "red")
        ),
 waffle(???, 
        rows = 10, 
        title="Michigan State's offense", 
        xlab="1 square = 1 yard", 
        colors = c("black", "dark green", "?????")
        )
)
```
```{r waffle4-solution, exercise.reveal_solution = FALSE}
nu2 <- c("Rushing"=187, "Passing"=255)
ms2 <- c("Rushing"=71, "Passing"=183, 188)

iron(
 waffle(nu2, 
        rows = 10, 
        title="Nebraska's offense", 
        xlab="1 square = 1 yard", 
        colors = c("black", "red")
        ),
 waffle(ms2, 
        rows = 10, 
        title="Michigan State's offense", 
        xlab="1 square = 1 yard", 
        colors = c("black", "dark green", "white")
        )
)
```
```{r waffle4-check}
grade_this_code()
```

### Exercise 5: Many units, one box

One last thing we can do is change the 1 square = 1 yard bit -- which makes the squares really small in this case -- by dividing our vector. Because of how R makes vectors, you can just divide it by a number and R will know to divide the numbers inside the vector by that number. Take what you just did and divide it by 2 and see what happens. Reminder: in R, divide is just the slash.

```{r waffle5, exercise=TRUE, exercise.setup = "waffle-load-data", message=FALSE}
nu2 <- c("Rushing"=187, "Passing"=255)
ms2 <- c("Rushing"=71, "Passing"=183, 188)

iron(
 waffle(???/?, 
        rows = 10, 
        title="Nebraska's offense", 
        xlab="1 square = 2 yards", 
        colors = c("black", "red")
        ),
 waffle(???/?, 
        rows = 10, 
        title="Michigan State's offense", 
        xlab="1 square = 2 yards", 
        colors = c("black", "dark green", "?????")
        )
)
```
```{r waffle5-solution, exercise.reveal_solution = FALSE}
nu2 <- c("Rushing"=187, "Passing"=255)
ms2 <- c("Rushing"=71, "Passing"=183, 188)

iron(
 waffle(nu2/2, 
        rows = 10, 
        title="Nebraska's offense", 
        xlab="1 square = 2 yards", 
        colors = c("black", "red")
        ),
 waffle(ms2/2, 
        rows = 10, 
        title="Michigan State's offense", 
        xlab="1 square = 2 yards", 
        colors = c("black", "dark green", "white")
        )
)
```
```{r waffle5-check}
grade_this_code()
```
News flash: Nebraska beat Michigan State everywhere but on the scoreboard and Nebraska is changing its fight song to "Everybody Hurts" by REM. 