---
title: "Lesson 25: Encirlcing points"
output: 
  learnr::tutorial:
    progressive: true
    allow_skip: true
runtime: shiny_prerendered
description: >
  Learn how to draw attention to some points.
---

```{r setup, include=FALSE}
library(learnr)
library(gradethis)
library(tidyverse)
library(ggalt)
knitr::opts_chunk$set(echo = FALSE)
tutorial_options(exercise.completion=FALSE)
```

# Encircling points on a scatterplot

## The basics

One thing we've talked about all semester is drawing attention to the thing you want to draw attention to. We've used color and labels to do that so far. Let's add another layer to it -- a shape around the point or points you want to highlight. 

Remember: The point of all of this is to draw the eye to what you are trying to show your reader. You want people to see the story you are trying to tell. 

It's not hard to draw a shape in ggplot -- it is a challenge to put it in the right place. But, there is a library to the rescue that makes this super easy -- `ggalt`. 

Like usual, you already installed this if you followed the install instructions at the beginning of the course. But if you're having trouble getting this to work, go to your console in R Studio and install it with `install.packages("ggalt")`

There's a bunch of things that `ggalt` does, but one of the most useful for us is the function `encircle`. Let's dive in.

```{r load-tidyverse, exercise=TRUE}
library(tidyverse)
library(ggalt)

```
```{r load-tidyverse-solution}
library(tidyverse)
library(ggalt)
```
```{r load-tidyverse-check}
grade_this_code()
```

Let's say we want to highlight the quiet contributions of Derrick Walker to the Husker basketball team. There are bigger personalities, and bigger hype attached to others. But without Walker, this team would be in more trouble than it's already in. Let's take a look at this season's player data. Per usual: **For purposes of this exercise, you don't need to do this. The data is included here if you want to try this in your own notebook.** 

```{r echo=FALSE, class.output="bg-info", results="asis",  message=FALSE,  warning=FALSE}
library(downloadthis)
library(glue)

dllink <- download_link(
  link = "http://mattwaite.github.io/sportsdatafiles/players22.csv",
  button_label = "Download csv file",
  button_type = "danger",
  has_icon = TRUE,
  icon = "fa fa-save",
  self_contained = FALSE
)

glue("<pre><p><strong>For this walkthrough:</strong></p><p>{dllink}</p></pre>")
```

And while we're loading it, let's filter out anyone who hasn't played 100 minutes and hasn't pulled down a rebound. 

```{r encircle-load-data, message=FALSE, warning=FALSE}
players <- read_csv("http://mattwaite.github.io/sportsdatafiles/players22.csv") %>% filter(MP > 100 & TRB > 0)

nu <- players %>% filter(Team == "Nebraska Cornhuskers")

dw <- nu %>% filter(`FG%` > .75)
```
```{r encircle-load-data-exercise, exercise = TRUE}
players <- read_csv("http://mattwaite.github.io/sportsdatafiles/players22.csv") %>%
  filter(MP > 100 & TRB > 0)
```
```{r encircle-load-data-exercise-solution}
players <- read_csv("http://mattwaite.github.io/sportsdatafiles/players22.csv") %>%
  filter(MP > 100 & TRB > 0)
```
```{r encircle-load-data-exercise-check}
grade_this_code()
```

We've done this before, but let's make a standard scatterplot of shooting percentage and rebounds. 

### Exercise 1: Making the first scatterplot

Who is hitting shots, and who is pulling down boards -- two things the mens basketball team doesn't do particularly well. Our data is players, and our two fields are `FG%` -- note the backticks because of the percent sign -- and TRB.

```{r encircle1, exercise=TRUE, exercise.setup = "encircle-load-data", message=FALSE}
ggplot() + geom_point(data=???????, aes(x=`??%`, y=???))
```
```{r encircle1-solution, exercise.reveal_solution = FALSE}
ggplot() + geom_point(data=players, aes(x=`FG%`, y=TRB))
```
```{r encircle1-check}
grade_this_code()
```

We can see right away that there are some dots at the edges -- lots of rebounds but average shooting or high shooting and not a ton of rebounds. Where are the Huskers in this normal curve? We can get that with filtering and layering.

### Exercise 2: Adding the Huskers as a layer

Like we have done in the past, let's make a dataframe of Nebraska players and add a layer to our graph that puts Nebraska's dots in red. While we're at it, we'll make everyone else grey. 

```{r encircle2, exercise=TRUE, exercise.setup = "encircle-load-data", message=FALSE}
nu <- players %>% filter(Team == "Nebraska Cornhuskers")

ggplot() + 
  geom_point(data=???????, aes(x=`??%`, y=???), color="grey") +
  geom_point(data=??, aes(x=`??%`, y=???), color="red")
```
```{r encircle2-solution, exercise.reveal_solution = FALSE}
nu <- players %>% filter(Team == "Nebraska Cornhuskers")

ggplot() + 
  geom_point(data=players, aes(x=`FG%`, y=TRB), color="grey") + 
  geom_point(data=nu, aes(x=`FG%`, y=TRB), color="red")
```
```{r encircle2-check}
grade_this_code()
```

Given that this chart is about Derrick Walker, I'll give you three guesses who that dot is shooting at a high percentage and leading the team in rebounds. Your first two guesses don't count and his last name rhymes with talker.

But just this quickly, we're on the path to something publishable. We'll need to label a dot and we'll need to drop the default grey and add some headlines and all that. And, for the most part, we've got a solid chart. 

But what if we could really draw the eye to a dot or group of them? In `ggalt`, there is a geom called `geom_encircle`, which ... does what you think it does. It encircles all the dots in a dataset. 

### Exercise 3: Our first encircle

Let's draw a circle around, uh, that one guy. I'm being coy here for a reason. Sometimes the group you want to highlight isn't known to you. Sometimes you want to highlight players by some criteria. In our case, if we look at the graph above, we know we have a player who is shooting better than 70 percent (or .7). To show you how, let's use that x axis interpolation as our filter.

Then we'll add geom_encircle and we'll just copy the data and the aes from our nu geom_point. Then, we need to give the encirclement a shape using s_shape -- which is a number between 0 and 1 -- and then how far away from the dots to draw the circle using expand, which is another number between 0 and 1.

Let's start with `s_shape` 1 and `expand` 1.

```{r encircle3, exercise=TRUE, exercise.setup = "encircle-load-data", message=FALSE}
dw <- nu %>% filter(`FG%` > .75)

ggplot() + 
  geom_point(data=???????, aes(x=`??%`, y=???), color="grey") +
  geom_point(data=??, aes(x=`??%`, y=???), color="red") + 
  geom_encircle(data=dw, aes(x=`??%`, y=???), s_shape=1, expand=1, colour="red")
```
```{r encircle3-solution, exercise.reveal_solution = FALSE}
dw <- nu %>% filter(`FG%` > .75)

ggplot() + 
  geom_point(data=players, aes(x=`FG%`, y=TRB), color="grey") + 
  geom_point(data=nu, aes(x=`FG%`, y=TRB), color="red") + 
  geom_encircle(data=dw, aes(x=`FG%`, y=TRB), s_shape=1, expand=1, colour="red")
```
```{r encircle3-check}
grade_this_code()
```

Whoa. That's ... not good.

Now begins the process of fiddling with `s_shape` and `expand` until you get it just right. But let's add a layer of difficulty here that isn't that difficult to solve: Adding in a text label and using interpolation to move the circle and the text label around. 

### Exercise 4: Fudging it until it looks good

This takes a little bit of finessing, but a shape of .5 means the line will have some bend to it -- it'll look more like someone circled it with a pen. Then, the expand is better if you use hundredths instead of tenths. So .01 instead of .1. 

In terms of interpolation, this is just trial and error. In your `geom_text` and `geom_encirle` aes, you don't *have* to use column names. It's just easier, especially when you have multiple data points. In our case, our dw dataframe has one point in it, so we can use a little math and push it around a little bit.

If we just put in Walker's FG% and TRB into `geom_text`, his name would appear right on top of the dot, with the dot right in the middle of his name. We can use simple addition to move it up and left. Then, we can use that same math to shift the circle to the left, circling his name AND dot instead of just the dot. And, because we're using math instead of just putting hard numbers in, we make this chart future proof -- we can update the data and the dot, the text and the circle all move automatically.

Normally you'll have to fuss with this for a bit. I'm going to give you the fussing here. We're going to move the name and circle up 7 rebounds to get it above Walker's dot, and then we'll drop 5 percent off his shooting numbers to move his name to the left, which prevents it from getting cut off by the graph. 

```{r encircle4, exercise=TRUE, exercise.setup = "encircle-load-data", message=FALSE}
ggplot() + 
  geom_point(data=???????, aes(x=`??%`, y=???), color="grey") +
  geom_point(data=??, aes(x=`??%`, y=???), color="red") + 
  geom_text(data=??, aes(x=`??%`-.05, y=???+7, label=Player)) + 
  geom_encircle(data=dw, aes(x=`??%`-.05, y=???+7), s_shape=.1, expand=.01, colour="red")
```
```{r encircle4-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(data=players, aes(x=`FG%`, y=TRB), color="grey") +
  geom_point(data=nu, aes(x=`FG%`, y=TRB), color="red") + 
  geom_text(data=dw, aes(x=`FG%`-.05, y=TRB+7, label=Player)) + 
  geom_encircle(data=dw, aes(x=`FG%`-.05, y=TRB+7), s_shape=.1, expand=.01, colour="red")
```
```{r encircle4-check}
grade_this_code()
```

Now let's clean this up and make it presentable with labels, a theme and some alterations to that theme.

```{r encircle5, exercise=TRUE, exercise.setup = "encircle-load-data", message=FALSE}
ggplot() + 
  geom_point(data=???????, aes(x=`??%`, y=???), color="grey") +
  geom_point(data=??, aes(x=`??%`, y=???), color="red") + 
  geom_text(data=??, aes(x=`??%`-.05, y=???+7, label=Player)) + 
  geom_encircle(data=dw, aes(x=`??%`-.05, y=???+7), s_shape=.1, expand=.01, colour="red") + 
  labs(title="Derrick Walker is Nebraska's MVP", subtitle="When he shoots, he nearly always scores and he's by far the team's best rebounder.", x="Shooting percentage", y="Rebounds") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )
```
```{r encircle5-solution, exercise.reveal_solution = FALSE}
ggplot() + 
  geom_point(data=players, aes(x=`FG%`, y=TRB), color="grey") + 
  geom_point(data=nu, aes(x=`FG%`, y=TRB), color="red") + 
  geom_text(data=dw, aes(x=`FG%`-.05, y=TRB+7, label=Player)) + 
  geom_encircle(data=dw, aes(x=`FG%`-.05, y=TRB+7), s_shape=.1, expand=.01, colour="red") + 
  labs(title="Derrick Walker is Nebraska's MVP", subtitle="When he shoots, he nearly always scores and he's by far the team's best rebounder.", x="Shooting percentage", y="Rebounds") + 
  theme_minimal() + 
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.title = element_text(size = 8), 
    plot.subtitle = element_text(size=10), 
    panel.grid.minor = element_blank()
    )
```
```{r encircle5-check}
grade_this_code()
```